### 2022-09-02 manuscript.Rmd
create study progress report; started sketching out very rough beginning to intro/method (very very far from complete)

---
title: "Association between trauma severity models and opportunities for improvement"
author: "Victoria Witthoff"
output:
  pdf_document: default
  html_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

You can cite document here like this [@exampleKey9999]. Open the file
bibliography.bib to learn more.

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
## source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2022-09-02 studyprogressreport.Rmd
create study progress report; started sketching out very rough beginning to intro/method (very very far from complete)

---
title: "Association between trauma severity models and opportunities for improvement: A cohort study"
author: "Victoria Witthoff"
output:
  pdf_document: default
  html_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction
============

<!--
Scientific background:

* What is trauma?
* Stats about mortality in Sweden & worldwide? Or maybe just worldwide tbh? 
  + Sourced statistics about how it affects
* What progress has been made in trauma care? Has it gotten better over the years or do statistics stay stagnant?

Rationale:

-> Therefore, improvement of trauma care is important for this patient group -> you can't improve care without having good methods to evaluate what you're currently doing -> it would be good if we could improve our methods for quality assurance

There are multiple strategies for quality assurance. One is M&M that evaluates OFI. Another is using predictive models for benchmarking, such as xyz. Is there an association between OFI and trauma severity models? We don't know?

Aim:

The purpose of this study was to assess how TSM are associated with and to what extent they can predict OFI. 
-->

Severe trauma is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@WHOepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is the leading cause of death for people aged 5-29 years and 90% of trauma-related deaths occur in low- to middle-income countries [@WHOepi]. Nonfatal outcomes, such as permanent disability, also comprise a significant health burden and can drastically reduce quality of life and lifespan in survivors of severe trauma. Reduction of trauma morbidity and mortality is therefore both an imperative, considering the great number of people affected, and also a complex, multi-factorial challenge encompassing a broad range of organisations across continents.

Another characteristic of trauma care is that management and treatment of severe trauma is dependent on a broad chain of actors that span multiple professions, medical specialties, and sometimes multiple clinics. The management of a single trauma case can involve personnel from EMS, the ER, anaesthesia, radiology, and surgery. Improvement of trauma care therefore requires a multidisciplinary approach that examines all steps in the chain of treatment and identifies weak links. One method of doing this is by holding multidisciplinary morbidity and mortality reviews of selected cases in order to identify opportunities for improvement (OFI).

<!--
Thoughts/Notes/Things to potentially say

* Trauma is _____. 
  + Highly heterogenous
* Trauma is one of the leading causes of death among (____) [cite]. 
  + Age group? 
  + Geographical locations? 
* Treatment of severe trauma is time-dependent and is a multi-step process of prehospital management + diagnosis + surgery + perioperative care + post-operative care
  + Affected by routines/protocols/resources at the treating hospital
    + Trauma outcomes are affected by multiple factors
  + Mention levels of trauma centers? 
  + What protocols etc have been developed to manage trauma?
* Because of the above, quality assurance is necessary to evaluate all steps of the chain, since outcomes of trauma treatment are not affected by one single treatment strategy, but by the efficiency and accuracy of each step in the chain

Two methods of QA are M&M + TSM

-->

## Quality assurance and benchmarking

### Trauma severity models

Increasing the quality of trauma care necessitates consideration of several key characteristics of trauma. One of these is that trauma is a highly heterogeneous entity due to the variety in its underlying causes, which affect both the type and severity of the injury. Vehicle collisions are the primary cause of mortality due to physical injury worldwide [@WHOepi], but the list of potential causes of severe physical injury is lengthy, including everything from workplace accidents to physical violence. This heterogeneity makes it poses challenges in comparing trauma outcomes and has spurred the development of multiple trauma severity models (TSM) that aim to overcome this difficulty.

One attempt to solve the problem of the heterogeneity of trauma is the development of multiple trauma severity models (TSM). These models differ in their methodology but all have the aim of establishing a standardised system to compare trauma outcomes and thus aid in quality assurance and benchmarking. The most commonly used algorithm is the Trauma and Injury Severity Score (TRISS)

<!--
What are the differences between the models?
Are there stats on how good they are? Are any better/worse? Why use different ones?

-->

### Morbidity & mortality reviews

<!--
Introduce the concept of OFI here
-->

## Aims

This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

<!-- According to student instructions, 3-5 pages -->

<!-- cite like [@exampleKey9999] -->

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->
This is a single-centre registry-based cohort study.

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

The trauma registry database includes around 21,000 patients treated between 2012 and 2021. The trauma care quality database is a subset of the trauma registry database and includes around 2200 patients who were selected for review between 2014 and 2021. Both databases come from the Karolinska University Hospital in Solna, Sweden. Karolinska University Hospital is classified as a Level 1 trauma center according to American criteria.

<!-- Last sentence prolly not particularly good rn but it's still probably(?) important to mention the context of trauma care at KS; maybe I should find out what trauma patients come to NKS vs other sthlm hospitals; presumably other sthlm hospitals take minor trauma?; https://www.karolinska.se/en/karolinska-university-hospital/news/2020/12/karolinska-university-hospital-becomes-swedens-first-trauma-centre/ -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

* Eligibility 
  + Does it include ALL patients in this time span?
  + Criteria for inclusion in the database, is it simply everyone coming in coded as trauma? 
  + Which patients were selected for review? Presumably cases where the people involved felt that they needed to be discussed? But is there more specific info? 
* Do patients know they are recorded in the registry? Or is this a thing where permission isn't required since it doesn't affect patients directly?

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

The studied outcome is "opportunities for improvement" (OFI), as identified by the  multidisciplinary review board after a mortality and morbidity review. OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement"

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->
* Geographical bias?
* People in the multidisciplinary review committee - is it basically the same people all the time?

Study size 
----------

<!-- Explain how the study size was arrived at -->

- The size of the register? (The big one or the subset?)

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

Should answer these questions, where applicable:

* All statistical methods, incl. those used to control for confounding
* Any methods to examine subgroups and interactions
* How was missing data addressed?
* Cohort study: If applicable, explain how loss to follow-up was addressed 
* Any sensitivity analyses?

Ethical considerations
-------

**Stockholm Research Ethics Review Board approval number 2021-02541 and 2021-03531.**

Potential things to mention here:

* The study doesn't affect treatment of patients in any way
* The data is stored in a secure database (I assume)
* I don't have access to patient journals; only the info in the register

<!-- 
THIS DOESN'T WORK FOR WHATEVER REASON AND MAKES KNITTING THROW AN ERROR
So I'll fix it sometime later
source("main.R")  "imports" the main script file of your project and run any code in it -->

Development and time plan
==========

All data required for this project were collected and available prior to the project beginning. Therefore, the data only need to be analysed.

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/timeplan.png")
```

Reserve plan
------------

Since all of the data has already been collected at the commencement of this project, there is no need for an elaborate back up plan. However, some details of the project may be subject to change during the analysis (?)

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

Vestibulum a velit ullamcorper, porttitor eros ut, tempor dolor. Fusce ac lacus at mauris consequat lobortis et ac sapien. In neque neque, rhoncus quis venenatis semper, iaculis malesuada nunc. Vestibulum nec orci porttitor, mattis nulla efficitur, hendrerit erat. Fusce blandit non neque eget semper. Curabitur at ullamcorper urna. Aliquam placerat pretium dui, et feugiat risus semper non. Donec ac erat et quam pellentesque posuere. Nullam elementum, tellus ac mollis pretium, lorem tellus elementum erat, nec tincidunt elit nibh eget nunc. Quisque ut nulla tellus.

Pellentesque nisi enim, cursus at leo in, pulvinar tincidunt ante. Maecenas eget porta sem. Nunc pulvinar consectetur blandit. Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Ut pellentesque lectus in nunc mattis, nec bibendum ipsum vulputate. Integer laoreet purus vel erat mollis, vitae placerat dolor faucibus. Etiam convallis orci vel neque sodales, sed tincidunt enim facilisis. Curabitur et porta leo. Pellentesque consectetur posuere mauris, ut ullamcorper elit tempor vitae. Maecenas condimentum accumsan felis et tincidunt. Fusce hendrerit id libero a tincidunt.

Suspendisse congue felis neque, ac gravida metus ullamcorper ut. Integer viverra et ex vel malesuada. Vivamus gravida volutpat libero, in accumsan felis tempor id. Integer non venenatis mauris, at sodales nibh. Aenean sit amet velit vitae augue interdum tincidunt vel sed neque. Mauris nec ligula neque. Ut vel posuere justo. Etiam nunc tortor, consequat quis lorem sed, iaculis pellentesque eros. Nulla ornare sapien est, sit amet convallis metus aliquet sit amet. Nulla facilisi.

### 2022-09-08 studyprogressreport.Rmd
Progress on the method section

@@
 ---
-title: "Association between trauma severity models and opportunities for improvement: A cohort study"
+title: "Association between trauma severity models and opportunities for improvement: A retrospective cohort study"
 author: "Victoria Witthoff"
@@
 ```
+
+Abbreviations
+=============
+* KUH - Karolinska University Hospital
+* OFI - Opportunities for improvement
+* TSM - Trauma severity models
 
@@
 The purpose of this study was to assess how TSM are associated with and to what extent they can predict OFI.
--->
 
-Severe trauma is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@WHOepi].
-The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
-One of the characteristics of this particular pandemic is its demographic spread; trauma is the leading cause of death for people aged 5-29 years and 90% of trauma-related deaths occur in low- to middle-income countries [@WHOepi].
-Nonfatal outcomes, such as permanent disability, also comprise a significant health burden and can drastically reduce quality of life and lifespan in survivors of severe trauma.
-Reduction of trauma morbidity and mortality is therefore both an imperative, considering the great number of people affected, and also a complex, multi-factorial challenge encompassing a broad range of organisations across continents.
-
-Another characteristic of trauma care is that management and treatment of severe trauma is dependent on a broad chain of actors that span multiple professions, medical specialties, and sometimes multiple clinics.
-The management of a single trauma case can involve personnel from EMS, the ER, anaesthesia, radiology, and surgery.
-Improvement of trauma care therefore requires a multidisciplinary approach that examines all steps in the chain of treatment and identifies weak links.
-One method of doing this is by holding multidisciplinary morbidity and mortality reviews of selected cases in order to identify opportunities for improvement (OFI).
-
-<!--
 Thoughts/Notes/Things to potentially say
@@
   + What protocols etc have been developed to manage trauma?
-* Because of the above, quality assurance is necessary to evaluate all steps of the chain, since outcomes of trauma treatment are not affected by one single treatment strategy, but by the efficiency and accuracy of each step in the chain
+* Because of the above, quality assurance is necessary to evaluate all steps of the chain, since outcomes of trauma treatment are not affected by one single treatment strategy, but by the efficiency and accuracy of each step in the chain + how they cooperate together
 
 Two methods of QA are M&M + TSM 
+-->
+
+Severe trauma is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@WHOepi].
+The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
+One of the characteristics of this particular pandemic is its demographic spread; trauma is the leading cause of death for people aged 5-29 years and 90% of trauma-related deaths occur in low- to middle-income countries [@WHOepi].
+Nonfatal outcomes, such as permanent disability, also comprise a significant health burden and can drastically reduce quality of life and lifespan in survivors of severe trauma.
+<!-- cite?
+--> Reduction of trauma morbidity and mortality is therefore both an imperative, considering the great number of people affected, and also a complex, multi-factorial challenge encompassing a broad range of organisations across continents.
+
+Increasing the quality of trauma care necessitates consideration of several key characteristics of trauma.
+One of these is that trauma is a highly heterogeneous entity due to the variety in its underlying causes, which affect both the type and severity of the injury.
+Vehicle collisions are the primary cause of mortality due to physical injury worldwide [@WHOepi], but the list of potential causes of severe physical injury is lengthy, including everything from workplace accidents to physical violence.
+This heterogeneity makes it poses challenges in comparing trauma outcomes and has spurred the development of multiple trauma severity models (TSM) that aim to overcome this difficulty.
+These models
+
+One attempt to solve the problem of the heterogeneity of trauma is the development of multiple trauma severity models (TSM).
+These models differ in their methodology but all have the aim of establishing a standardised system to compare trauma outcomes and thus aid in quality assurance and benchmarking.
+
+One characteristic of trauma care is that management and treatment of severe trauma is dependent on a broad chain of actors that span multiple professions, medical specialties, and sometimes multiple clinics.
+For example, the management of a single trauma case could involve personnel from EMS, the ER, anaesthesia, radiology, and surgery.
+Improvement of trauma care therefore requires a multidisciplinary approach that examines all steps in the chain of treatment and identifies weak links.
+One method of doing this is by holding multidisciplinary morbidity and mortality reviews of selected cases in order to identify opportunities for improvement (OFI).
+
+<!--
+
+I NEED TO FIND RESEARCH IN THE AREA AND TALK ABOUT THEM AND CITE THEM
 
@@
 ### Trauma severity models
-
-Increasing the quality of trauma care necessitates consideration of several key characteristics of trauma.
-One of these is that trauma is a highly heterogeneous entity due to the variety in its underlying causes, which affect both the type and severity of the injury.
-Vehicle collisions are the primary cause of mortality due to physical injury worldwide [@WHOepi], but the list of potential causes of severe physical injury is lengthy, including everything from workplace accidents to physical violence.
-This heterogeneity makes it poses challenges in comparing trauma outcomes and has spurred the development of multiple trauma severity models (TSM) that aim to overcome this difficulty.
-
-One attempt to solve the problem of the heterogeneity of trauma is the development of multiple trauma severity models (TSM).
-These models differ in their methodology but all have the aim of establishing a standardised system to compare trauma outcomes and thus aid in quality assurance and benchmarking.
-The most commonly used algorithm is the Trauma and Injury Severity Score (TRISS)
 
@@
 Introduce the concept of OFI here
+
+Actually I don't think this subheading is all that necessary 
 -->
@@
 This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
-
-<!-- Explain the scientific background and rationale for the
-investigation being reported. End by stating the aim or specific
-objectives, including any prespecified hypotheses. The introduction
-should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
-6-8 lines long. Support every statement with a citation -->
-
-<!-- According to student instructions, 3-5 pages -->
-
-<!-- cite like [@exampleKey9999] -->
 
@@
 <!-- Present key elements of study design -->
-This is a single-centre registry-based cohort study.
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry database and the trauma care quality registry.
+The trauma care quality registry is a subset of the trauma registry and contains data on trauma patients who were selected for morbidity and mortality (M&M) reviews.
+These two databases will be linked and then analysis will be performed to estimate the association between the common trauma severity models (TSM) TRISS, NORMIT, and RISCII, and opportunities for improvement (OFI) using logistic regression.
+The predictive performance of these models will be evaluated using measures of discrimination and calibration.
 
@@
 
-The trauma registry database includes around 21,000 patients treated between 2012 and 2021.
-The trauma care quality database is a subset of the trauma registry database and includes around 2200 patients who were selected for review between 2014 and 2021.
-Both databases come from the Karolinska University Hospital in Solna, Sweden.
-Karolinska University Hospital is classified as a Level 1 trauma center according to American criteria.
+KUH is located in Stockholm, Sweden.
+KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@NKStrauma].
+KUH is the primary trauma centre for the region of Stockholm.
+In addition, KUH has agreements with several other regions to provide trauma care for patients with severe trauma.
+KUH is thereby the major trauma centre for a population of almost 3 million people [@NKStrauma].
 
-<!-- Last sentence prolly not particularly good rn but it's still probably(?) important to mention the context of trauma care at KS; maybe I should find out what trauma patients come to NKS vs other sthlm hospitals; presumably other sthlm hospitals take minor trauma?; https://www.karolinska.se/en/karolinska-university-hospital/news/2020/12/karolinska-university-hospital-becomes-swedens-first-trauma-centre/ -->
+The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021.
+This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
+The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
+It also includes tracks follow-up.
+Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
+
+The trauma care quality register is an internal hospital register and registers the outcome of M&M conferences.
+M&M conferences are multidisciplinary and their purpose is to review selected patient cases and determine whether an opportunity for improvement existed in each individual case.
+Examples of opportunities for improvement could be insufficient resources or errors in patient management.
 
@@
 
-<!-- Cohort study: Give the eligibility criteria, and the sources and
-methods of selection of participants. Describe methods of
-follow-up. For matched studies, give matching criteria and number of
-exposed and unexposed -->
+Participants include all patients registered in both the trauma registry and the trauma care quality registry at KUH.
+The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
+Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15.
+Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.
+Since the outcome variable of this study is OF
 
-* Eligibility 
-  + Does it include ALL patients in this time span?
-  + Criteria for inclusion in the database, is it simply everyone coming in coded as trauma? 
-  + Which patients were selected for review? Presumably cases where the people involved felt that they needed to be discussed? But is there more specific info? 
-* Do patients know they are recorded in the registry? Or is this a thing where permission isn't required since it doesn't affect patients directly?
+<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at? -->
 
-Variables and data sources/measurements
----------------------------------------
+The trauma care quality registry contains patients from the trauma registry that were selected for M&M review.
+At KUH, all fatalities are automatically selected for M&M review.
+In addition to this, the electronic medical records of all trauma patients are evaluated by research nursing staff to identify cases that potentially received sub-optimal care.
+Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
+
+Variables
+---------
 
@@
 
-The studied outcome is "opportunities for improvement" (OFI), as identified by the  multidisciplinary review board after a mortality and morbidity review.
-OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement"
+The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the  multidisciplinary review board after a mortality and morbidity review.
+OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement".
+The primary exposures of interest are trauma severity scores according to the TRISS, NORMIT, and RISCII models.
 
@@
 <!-- Describe any efforts to address potential sources of bias -->
-* Geographical bias?
-* People in the multidisciplinary review committee - is it basically the same people all the time?
+
+The method and data analysis model will be written using simulated data to reduce the risk of research bias.
+Real-world data will only be used once the data analysis model is completed and found to work correctly on simulated data.
 
@@
 
-- The size of the register? (The big one or the subset?)
+Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database.
+The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
+
+<!-- Update these numbers later when I have actual numbers -->
 
@@
 
+<!-- Really not sure
+-->
+
 Statistical methods
 -------------------
+
+R, a programming language for statistical computing, will be used for compilation and analysis of data. Variables will be handled according to the SweTrau manual.
+
+<!--
 
@@
 * Cohort study: If applicable, explain how loss to follow-up was addressed 
-* Any sensitivity analyses?
+* Any sensitivity analyses? -->
 
@@
 <!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->
-
-Vestibulum a velit ullamcorper, porttitor eros ut, tempor dolor.
-Fusce ac lacus at mauris consequat lobortis et ac sapien.
-In neque neque, rhoncus quis venenatis semper, iaculis malesuada nunc.
-Vestibulum nec orci porttitor, mattis nulla efficitur, hendrerit erat.
-Fusce blandit non neque eget semper.
-Curabitur at ullamcorper urna.
-Aliquam placerat pretium dui, et feugiat risus semper non.
-Donec ac erat et quam pellentesque posuere.
-Nullam elementum, tellus ac mollis pretium, lorem tellus elementum erat, nec tincidunt elit nibh eget nunc.
-Quisque ut nulla tellus.
-
-Pellentesque nisi enim, cursus at leo in, pulvinar tincidunt ante.
-Maecenas eget porta sem.
-Nunc pulvinar consectetur blandit.
-Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-Ut pellentesque lectus in nunc mattis, nec bibendum ipsum vulputate.
-Integer laoreet purus vel erat mollis, vitae placerat dolor faucibus.
-Etiam convallis orci vel neque sodales, sed tincidunt enim facilisis.
-Curabitur et porta leo.
-Pellentesque consectetur posuere mauris, ut ullamcorper elit tempor vitae.
-Maecenas condimentum accumsan felis et tincidunt.
-Fusce hendrerit id libero a tincidunt.
-
-Suspendisse congue felis neque, ac gravida metus ullamcorper ut.
-Integer viverra et ex vel malesuada.
-Vivamus gravida volutpat libero, in accumsan felis tempor id.
-Integer non venenatis mauris, at sodales nibh.
-Aenean sit amet velit vitae augue interdum tincidunt vel sed neque.
-Mauris nec ligula neque.
-Ut vel posuere justo.
-Etiam nunc tortor, consequat quis lorem sed, iaculis pellentesque eros.
-Nulla ornare sapien est, sit amet convallis metus aliquet sit amet.
-Nulla facilisi.

### 2022-09-09 project-timeline.rmd
detailed project timeline with dates for deadlines

# Project timeline

## Week 1 (29/9 - 2/9)
- Setup Github ([done](https://github.com/victoria-l-w/trauma-severity-ofi))
- Draft study progress report
- Reading

## Week 2 (5/9 - 9/9) 
- Continue drafting study progress report
- Detailed project  timeline

## Week 3 (12/9 - 16/9)
- Complete draft of study progress report so that there's a week+ for feedback
- Start writing code that imports and begins to clean data
- Start drafting a more publication-friendly version

## Week 4 (19/9 - 23/9)
- Revise study progress report based on feedback, submit
- Revise code based on feedback.
- Add code that creates a first table of sample characteristics.
- Continue improving publication-friendly version

## Week 5-10 (26/9 - 4/11)
- **Deadline**: Submit study progress report by 12:00 Monday 26/9 
- **Presentations**: Present study progress report 4-5/10 
- Write code that generate a first set of preliminary results and
  revise weekly based on feedback.

## Week 11-12 (7/11 - 18/11)
- Draft the Results section and upload.
- Revise code based on feedback.

## Week 13-14 (21/11 - 2/12)
- Draft Discussion section and upload.
- Revise Results based on feedback.

## Week 15 (5/12 - 9/12)
- Draft and upload final project paper.

## Week 16 (12/12 - 16/12) 
- Revise final project paper and upload a second version.

## Week 17 (19/12 - 23/12)
- Revise second version of final project paper and submit.

## Week 18-20 (26/12 - 13/1)
- **Deadline**: Submit seminar version by 12:00 Tuesday 27/12
- **Presentations**: Seminar presentations either 5/1 or 9/1
- **Deadline**: Submit final version by 12:00 Friday 13/1
- Revise project paper based on comments and feedback from peers,
  course leaders, examiners, and supervisors

### 2022-09-09 studyprogressreport.Rmd
small corrections
test
lots of progress on introduction

@@
 * KUH - Karolinska University Hospital
+* KUH TCQD - Karolinska University Hospital Trauma Care Quality Database
+* KUH TR - Karolinska University Hospital Trauma Registry
 * OFI - Opportunities for improvement
-* TSM - Trauma severity models
 
@@
 
-<!--
-Scientific background:
+<!-- Why trauma is a big deal -->
 
-* What is trauma?
-* Stats about mortality in Sweden & worldwide? Or maybe just worldwide tbh? 
-  + Sourced statistics about how it affects
-* What progress has been made in trauma care? Has it gotten better over the years or do statistics stay stagnant?
+Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
+The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
+One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
+Nonfatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and lifespan in survivors of severe trauma.
+Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement programs improve outcomes [@ustqi].
 
-Rationale:
+<!-- Explain the what and why of TSM -->
 
--> Therefore, improvement of trauma care is important for this patient group -> you can't improve care without having good methods to evaluate what you're currently doing -> it would be good if we could improve our methods for quality assurance
+Trauma severity scores (TSS) are used in trauma quality improvement programs and research for their usefulness in benchmarking and in comparing trauma severity between patients [@whoguidelines].
+The existence of these models is necessitated by the heterogeneous nature of trauma, due to the variety in its underlying causes, which affect both the type and severity of the injury.
+This heterogeneity poses challenges in comparing and evaluating trauma outcomes and has spurred the development of multiple TSS.
+TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced models have been developed for quality assurance and research purposes.
+These models differ in their methodology but all have the aim of translating injury severity into quantitative values.
 
-There are multiple strategies for quality assurance.
-One is M&M that evaluates OFI.
-Another is using predictive models for benchmarking, such as xyz.
-Is there an association between OFI and trauma severity models?
-We don't know?
+The Trauma and Injury Severity Score (TRISS) is one of the most commonly used models.
+TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss].
+TRISS estimates the possibility of survival for the given parameters, and can be used by hospitals as a benchmark to compare outcomes against a standardised survival estimate.
 
-Aim:
+<!-- Info on other scores here? -->
 
-The purpose of this study was to assess how TSM are associated with and to what extent they can predict OFI.
+<!-- Explain the what and why of M&M -->
 
-Thoughts/Notes/Things to potentially say
+Another cornerstone of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality conferences (MMC).
+Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
+Errors in management are a common cause of preventable deaths [@errors].
+These conferences review selected patient cases and attempt to establish whether suboptimal care was given in each specific case, i.e.
+whether there was room for improvement, and to implement corrective actions.
+The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines].
 
-* Trauma is _____. 
-  + Highly heterogenous
-* Trauma is one of the leading causes of death among (____) [cite]. 
-  + Age group? 
-  + Geographical locations? 
-* Treatment of severe trauma is time-dependent and is a multi-step process of prehospital management + diagnosis + surgery + perioperative care + post-operative care
-  + Affected by routines/protocols/resources at the treating hospital
-    + Trauma outcomes are affected by multiple factors
-  + Mention levels of trauma centers? 
-  + What protocols etc have been developed to manage trauma?
-* Because of the above, quality assurance is necessary to evaluate all steps of the chain, since outcomes of trauma treatment are not affected by one single treatment strategy, but by the efficiency and accuracy of each step in the chain + how they cooperate together
+<!-- Explain why TSM ⇒ M&M -->
 
-Two methods of QA are M&M + TSM 
--->
+One limitation of M&M conferences is that their value in long-term quality improvement is dependent on selection of appropriate cases for review.
+The process by which patient cases are selected for M&M review differs between clinics and one systematic review found that case selection is often non-standardised [@mmreview].
+There is an inherent weakness in relying on healthcare providers to self-report on cases they were involved in.
+One study in South Africa found that using electronic medical records to automatically generate reports led to a dramatic increase in reporting of morbidity and mortality over a system of self-reporting for MMC [@southafrica].
+In institutions that have implemented more standardised approaches to case selection, factors that indicate case review include deaths, adverse events, and complications [@whoguidelines].
 
-Severe trauma is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@WHOepi].
-The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
-One of the characteristics of this particular pandemic is its demographic spread; trauma is the leading cause of death for people aged 5-29 years and 90% of trauma-related deaths occur in low- to middle-income countries [@WHOepi].
-Nonfatal outcomes, such as permanent disability, also comprise a significant health burden and can drastically reduce quality of life and lifespan in survivors of severe trauma.
-<!-- cite?
---> Reduction of trauma morbidity and mortality is therefore both an imperative, considering the great number of people affected, and also a complex, multi-factorial challenge encompassing a broad range of organisations across continents.
-
-Increasing the quality of trauma care necessitates consideration of several key characteristics of trauma.
-One of these is that trauma is a highly heterogeneous entity due to the variety in its underlying causes, which affect both the type and severity of the injury.
-Vehicle collisions are the primary cause of mortality due to physical injury worldwide [@WHOepi], but the list of potential causes of severe physical injury is lengthy, including everything from workplace accidents to physical violence.
-This heterogeneity makes it poses challenges in comparing trauma outcomes and has spurred the development of multiple trauma severity models (TSM) that aim to overcome this difficulty.
-These models
-
-One attempt to solve the problem of the heterogeneity of trauma is the development of multiple trauma severity models (TSM).
-These models differ in their methodology but all have the aim of establishing a standardised system to compare trauma outcomes and thus aid in quality assurance and benchmarking.
-
-One characteristic of trauma care is that management and treatment of severe trauma is dependent on a broad chain of actors that span multiple professions, medical specialties, and sometimes multiple clinics.
-For example, the management of a single trauma case could involve personnel from EMS, the ER, anaesthesia, radiology, and surgery.
-Improvement of trauma care therefore requires a multidisciplinary approach that examines all steps in the chain of treatment and identifies weak links.
-One method of doing this is by holding multidisciplinary morbidity and mortality reviews of selected cases in order to identify opportunities for improvement (OFI).
-
-<!--
-
-I NEED TO FIND RESEARCH IN THE AREA AND TALK ABOUT THEM AND CITE THEM
-
--->
-
-## Quality assurance and benchmarking
-
-### Trauma severity models
-
-<!--
-What are the differences between the models?
-Are there stats on how good they are? Are any better/worse? Why use different ones?
-
--->
-
-### Morbidity & mortality reviews
-
-<!--
-Introduce the concept of OFI here
-
-Actually I don't think this subheading is all that necessary 
--->
+TSS are already used in some clinics to identify cases for peer review [@whoguidelines].
+However, the extent to which TSS are associated with or can predict opportunities for improvement (OFI) as found in M&M conferences is poorly understood.
+If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
 
@@
 Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.
-Since the outcome variable of this study is OF
 
@@
 
-Reserve plan
-------------
-
-Since all of the data has already been collected at the commencement of this project, there is no need for an elaborate back up plan.
-However, some details of the project may be subject to change during the analysis (?)
-
 References

### 2022-09-09 studyprogressreport.Rmd
ethics section

@@
 Errors in management are a common cause of preventable deaths [@errors].
-These conferences review selected patient cases and attempt to establish whether suboptimal care was given in each specific case, i.e.
-whether there was room for improvement, and to implement corrective actions.
+These conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case and to implement corrective actions.
+OFI function in this manner as a proxy for suboptimal care.
 The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines].
@@
 Ethical considerations
--------
+----------------------
+
+All patient information used in this study is anonymised and stored in a secure database.
+Patient medical records will not be accessed.
+Patients were not asked to give consent for inclusion in the KUH trauma registry.
+The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations.
+Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death.
+Patients have the right to be removed from the register if they wish, and also have the right to a copy of data that is stored about them on request.
+
+This project is a retrospective cohort study and the project has thereby had no effect on patient treatment.
+There is no risk that patients are harmed by inclusion in the study.
+This study has the potential to improve patient care by analysing a potential factor that could predict suboptimal trauma care.
+It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified by the fact that this population stands to gain the most from improved trauma care.
 
@@
 
-Potential things to mention here:
+<!--
 
-* The study doesn't affect treatment of patients in any way
-* The data is stored in a secure database (I assume)
-* I don't have access to patient journals; only the info in the register
-
-<!-- 
-THIS DOESN'T WORK FOR WHATEVER REASON AND MAKES KNITTING THROW AN ERROR
-So I'll fix it sometime later
 source("main.R")  "imports" the main script file of your project and run any code in it -->

### 2022-09-09 studyprogressreport.Rmd
method; development and time plan; lots of small fixes

@@
 If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-
-## Aims
-
 This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
@@
 This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry database and the trauma care quality registry.
-The trauma care quality registry is a subset of the trauma registry and contains data on trauma patients who were selected for morbidity and mortality (M&M) reviews.
+The trauma care quality registry is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
 These two databases will be linked and then analysis will be performed to estimate the association between the common trauma severity models (TSM) TRISS, NORMIT, and RISCII, and opportunities for improvement (OFI) using logistic regression.
@@
 The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
-It also includes tracks follow-up.
 Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
@@
 The trauma care quality register is an internal hospital register and registers the outcome of M&M conferences.
-M&M conferences are multidisciplinary and their purpose is to review selected patient cases and determine whether an opportunity for improvement existed in each individual case.
-Examples of opportunities for improvement could be insufficient resources or errors in patient management.
+Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
 
@@
 
-<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at? -->
+<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at NKS? -->
 
@@
 
-<!-- Update these numbers later when I have actual numbers -->
+<!-- Update these numbers later when I have real numbers -->
 
@@
 -->
+The exposure variables are the values given by the studied trauma models. These variables are continuous. The outcome variable, OFI, is a binary variable.
 
-<!-- Really not sure
+<!-- Really not sure exactly what to put here
 -->
@@
 
-R, a programming language for statistical computing, will be used for compilation and analysis of data. Variables will be handled according to the SweTrau manual.
+R, a programming language for statistical computing, will be used for compilation and analysis of data.
+Conversion and handling of variables will be carried out according to the SweTrau manual.
+Results will be presented with a confidence interval of 95%.
+A p-value of <0.05 will be considered significant.
 
+<!-- Complete case study? -->
 <!--
@@
 
-All data required for this project were collected and available prior to the project beginning. Therefore, the data only need to be analysed.
+<!--
+B) En skriftlig redogörelse på ca en A4-sida av projektets utveckling, plus tidsplan för
+resterande del av kursen (använd tidsplanmallen i Canvas). Ange så exakt som möjligt:
+• Vilken typ av data, kategorier, variabler etc du avser att samla in
+• Hur data planeras presenteras (t ex tabeller och hur de kan struktureras, diagramtyper,
+vävnadssnitt, intervjucitat etc)
+• Hur data planeras analyseras; vilka statistiska metoder för jämförelser mellan en eller
+flera grupper, andra matematiska metoder, kvalitativ analys etc
+-->
+
+This is a registry-based project and all data required were collected and available prior to the project beginning.
+The data consists of information about trauma patients at KUH and the outcomes of morbidity & mortality conferences on selected patient cases.
+The next step after submitting this progress report is to begin writing the code that will import and clean data.
+Once this is completed, the code for data analysis will be written.
+The analysis will consist of using logistic regression to see if trauma severity models can predict opportunities for improvement.
+Initially the code will be written using scrambled data in order to minimise researcher bias.
+Once the code for analysis is complete, it will be run on the real-life data to obtain the results for the study.
+I have written a detailed time plan, including dates for deadlines, that is available for my supervisors.
+We will be in contact on a regular basis to discuss progress.
+All of my code and text will be uploaded on a regular basis, which will allow for my supervisors to see my progress.
 

### 2022-09-11 project-timeline.rmd
Comments JA 22.09.11

@@
 # Project timeline
-
+<!--Mycket mycket bra Victoria! -->
 ## Week 1 (29/9 - 2/9)

### 2022-09-11 studyprogressreport.Rmd
Comments JA 22.09.11

@@
 ============
-
+<!--Mycket bra! Röd tråd och "tratt". -->
 <!-- Why trauma is a big deal -->
@@
 TRISS estimates the possibility of survival for the given parameters, and can be used by hospitals as a benchmark to compare outcomes against a standardised survival estimate.
+<!--Jag skulle nämna dina andra scores här -->
 
 <!-- Info on other scores here? -->
-
 <!-- Explain the what and why of M&M --> 
-
+<!-- -->
 Another cornerstone of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality conferences (MMC).
@@
 The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+<!--Kolla exv Strömmers och Ghorbanis pek,https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/29797712/ .
+De försöker använda bl.aa normit och triss för att se korrelans till OFI och preventable deaths.
+Skillnaden mot det du gör är att de bara tittar på patienter som dör och inte hela cohorten.
+Du tittar också på ett betydligt större dataunderlag.
+Så ditt bygger vidare på deras arbete-->
 
@@
 This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
+<!--Introduktionen är mycket bra!
+Egentligen bara fortsätta som du gör, läs på och lägg till meningar/källor när du hittar info.
+För att nå KI:s sidkrav kan du skriva en hel del om de olika Scorsen -->
 
@@
 
-The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021.
-This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
+The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021.<!--Det kanske är så många?
+Var hittade du den siffran?
+--> This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
 The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
 Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
+<!--Jag skulle nämna att trauma registret följer "the utstein template" vilket är en internationell konsensus kring hur register byggs, inkl trauma register i detta fall -->
 
@@
 Participants include all patients registered in both the trauma registry and the trauma care quality registry at KUH.
-The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
+<!--När du kommit igång med analysen så ska faktiskt nummer in här --> The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
 Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15.
@@
 <!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at NKS? -->
+<!--Det stämmer! -->
 
@@
 Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
+<!--Helt korrekt, de använder även "audit filter" i sin screening, ev behöver du nämna detta också, se tid studenters arb. -->
 
@@
 ----------------------
-
+<!--Om du behöver "skapa" ny data genom att manipulera den insamlade (tvek i ditt fall) så ska du beskriva det här. Annars kan detta stycke slås ihop med variables ovan. -->
 <!-- Explain how quantitative variables were handled in the
@@
 The exposure variables are the values given by the studied trauma models. These variables are continuous. The outcome variable, OFI, is a binary variable. 
-
+<!--Om du inte tydligt definerat modellernas variabler behöver du göra det här, alltså för exv triss som är GCS, RR, BT + ISS osv osv-->
 <!-- Really not sure exactly what to put here
@@
 A p-value of <0.05 will be considered significant.
-
+<!-- Jo det kommer nog bli ccs, du kommer exkludera alla patienter som delsa saknar outcome, OFI alternativt där det saknas nog data för att bygga olika scores. -->
 <!-- Complete case study? -->

### 2022-09-12 meetingnotes.rmd
meeting notes 12/9

## 12/9
**Attending**: Me, George, Jonatan

**Discussed**:

* Using RISCII is probably not possible because the database lacks necessary parameters. Just use TRISS and NORMIT, or find a new third (or more?) score to use? Something for me to look into and think about. 
* Meeting KI requirements and deadlines more important than aims to publish, KI is prio 1.

**Plan**

* Access database
* Review what's available in the database + look into scoring systems, decide if there are any suitable scoring systems that would be interesting to add -> this would mean I'd have to calculate any extra scores myself b/c they aren't ready-made in the database
* Complete draft of projektredogörelse by end of week

**Next meeting**: Tuesday 20/12 over lunch, either Zoom or irl.

### 2022-09-12 studyprogressreport.Rmd
lots of text about m&m; other small fixes

@@
   pdf_document: default
-  html_document: default
 bibliography: bibliography.bib
@@
 =============
+* ASA-PS - American Society of Anesthesiology Physical Status
 * KUH - Karolinska University Hospital
@@
 * KUH TR - Karolinska University Hospital Trauma Registry
+* NORMIT - Norwegian Survival Prediction Model in Trauma
 * OFI - Opportunities for improvement
+* Ps - Probability of survival
+* TRISS - Trauma and Injury Severity Score
 
@@
 ============
-<!--Mycket bra! Röd tråd och "tratt". -->
 <!-- Why trauma is a big deal -->
@@
 One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
-Nonfatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and lifespan in survivors of severe trauma.
-Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement programs improve outcomes [@ustqi].
+Nonfatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+
+Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
+Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
+However, it has been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still plenty of progress to be made in improving these programs.
+
+## Morbidity and mortality review
+
+The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
+M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The incidence of OFI function in this manner as a proxy for incidence of suboptimal care.
+The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
+Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqpianglo].
+The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+
+The value of M&M conferences in long-term quality improvement is dependent on selection of appropriate cases for review.
+The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+This weakness of M&M review has been subject to considerable efforts to develop strategies to standardise case selection.
+The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
+The latter benefit is not insignificant considering the large caseload found in most urban trauma centres.
+Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo].
+Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
 
@@
 
+## Trauma severity scores
 Trauma severity scores (TSS) are used in trauma quality improvement programs and research for their usefulness in benchmarking and in comparing trauma severity between patients [@whoguidelines].
@@
 TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced models have been developed for quality assurance and research purposes.
+Today there are hundreds of TSS models.
 These models differ in their methodology but all have the aim of translating injury severity into quantitative values.
 
-The Trauma and Injury Severity Score (TRISS) is one of the most commonly used models.
-TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss].
-TRISS estimates the possibility of survival for the given parameters, and can be used by hospitals as a benchmark to compare outcomes against a standardised survival estimate.
-<!--Jag skulle nämna dina andra scores här -->
+One of the limitations of any TSS model is that their estimates may be unreliable when applied to a population different from the one the model was developed from.
 
-<!-- Info on other scores here? -->
-<!-- Explain the what and why of M&M --> 
-<!-- -->
-Another cornerstone of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality conferences (MMC).
-Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
-Errors in management are a common cause of preventable deaths [@errors].
-These conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case and to implement corrective actions.
-OFI function in this manner as a proxy for suboptimal care.
-The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+The Trauma and Injury Severity Score (TRISS), first developed in 1987 [@triss] and most recently updated in 2010 <!-- Cite 2009 revision ref -->, is one of the most commonly used models worldwide.TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss].
+The resulting value is a percentage that estimates the probability of survival (Ps).
+
+<!-- Triss Limitations -->
+
+<!-- NORMIT -->
+
+<!-- Explain why TSM ⇒ M&M -->
+
  <!--Kolla exv Strömmers och Ghorbanis pek,https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/29797712/ .
@@
 
-<!-- Explain why TSM ⇒ M&M -->
-
-One limitation of M&M conferences is that their value in long-term quality improvement is dependent on selection of appropriate cases for review.
-The process by which patient cases are selected for M&M review differs between clinics and one systematic review found that case selection is often non-standardised [@mmreview].
-There is an inherent weakness in relying on healthcare providers to self-report on cases they were involved in.
-One study in South Africa found that using electronic medical records to automatically generate reports led to a dramatic increase in reporting of morbidity and mortality over a system of self-reporting for MMC [@southafrica].
-In institutions that have implemented more standardised approaches to case selection, factors that indicate case review include deaths, adverse events, and complications [@whoguidelines].
-
-TSS are already used in some clinics to identify cases for peer review [@whoguidelines].
-However, the extent to which TSS are associated with or can predict opportunities for improvement (OFI) as found in M&M conferences is poorly understood.
+ However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is poorly understood.
 If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
@@
 <!-- Present key elements of study design -->
-This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry database and the trauma care quality registry.
-The trauma care quality registry is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases will be linked and then analysis will be performed to estimate the association between the common trauma severity models (TSM) TRISS, NORMIT, and RISCII, and opportunities for improvement (OFI) using logistic regression.
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD).
+The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
+These two databases will be linked and then analysis will be performed to estimate the association between the common TSS models TRISS and NORMIT <!-- And something else?
+--> and OFI using logistic regression.
 The predictive performance of these models will be evaluated using measures of discrimination and calibration.
@@
 KUH is located in Stockholm, Sweden.
-KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@NKStrauma].
+KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
 KUH is the primary trauma centre for the region of Stockholm.
 In addition, KUH has agreements with several other regions to provide trauma care for patients with severe trauma.
-KUH is thereby the major trauma centre for a population of almost 3 million people [@NKStrauma].
+KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
 
-The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021.<!--Det kanske är så många?
-Var hittade du den siffran?
---> This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
+The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
+SweTrau follows the Utstein Trauma Template for reporting of data following major trauma.
 The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
 Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
-<!--Jag skulle nämna att trauma registret följer "the utstein template" vilket är en internationell konsensus kring hur register byggs, inkl trauma register i detta fall -->
 
@@
 
-Participants include all patients registered in both the trauma registry and the trauma care quality registry at KUH.
-<!--När du kommit igång med analysen så ska faktiskt nummer in här --> The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
+Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
+
+### Selection for the trauma registry
+
+The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
 Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15.
@@
 
-<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at NKS? -->
-<!--Det stämmer! -->
+### Selection for the TCQD
 
-The trauma care quality registry contains patients from the trauma registry that were selected for M&M review.
+The TCQD contains patients from the trauma registry that were selected for M&M review.
 At KUH, all fatalities are automatically selected for M&M review.
-In addition to this, the electronic medical records of all trauma patients are evaluated by research nursing staff to identify cases that potentially received sub-optimal care.
+In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care.
+These nurses also use audit filters that automatically highlight cases with abnormal parameters.
 Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
-<!--Helt korrekt, de använder även "audit filter" i sin screening, ev behöver du nämna detta också, se tid studenters arb. -->
 
@@
 of assessment methods if there is more than one group -->
+### Outcome
+The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH.
+OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement".
 
-The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the  multidisciplinary review board after a mortality and morbidity review.
-OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement".
-The primary exposures of interest are trauma severity scores according to the TRISS, NORMIT, and RISCII models.
+### Exposures
+The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models.
+Both of these models estimate the Ps based on expected mortality within 30 days.
+This is expressed a percentage and is a continuous variable.
+The variables integrated into these models are taken from arrival at hospital.
+
+#### TRISS
+
+The TRISS model incorporates patient age, Injury Severity Score (ISS), Revised Trauma Score (RTS), and whether the trauma was blunt or penetrating.
+
+The ISS is given by estimating the severity of physical injuries by anatomical region.
+Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury".
+The value given is points on a scale from 0 to 75.
+The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate.
+Values are taken from arrival at hospital.
+These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.
+
+#### NORMIT
+
+The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+The NISS differs from the ISS in that the NISS is the sum of the squares from only the three most injured anatomical regions.
 
@@
 -->
-The exposure variables are the values given by the studied trauma models. These variables are continuous. The outcome variable, OFI, is a binary variable. 
+
+### Exposures
+
 <!--Om du inte tydligt definerat modellernas variabler behöver du göra det här, alltså för exv triss som är GCS, RR, BT + ISS osv osv-->

### 2022-09-12 studyprogressreport.Rmd
more text about trauma severity models

@@
 
-## Morbidity and mortality review
+## Morbidity and mortality conferences
 
@@
 These models differ in their methodology but all have the aim of translating injury severity into quantitative values.
+However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
+Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions.
+This weakness is one reason why new models continue to be developed and improved around the globe.
 
-One of the limitations of any TSS model is that their estimates may be unreliable when applied to a population different from the one the model was developed from.
+The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide.
+TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
+The resulting value is a percentage that estimates the probability of survival (Ps).
+The ISS gives a score based on the three anatomical regions with the most severe injury.
+A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries.
+The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
+The RTS is difficult to apply in situations where the patient has already been intubated, which gives TRISS limited value in estimating Ps in certain populations.
 
-The Trauma and Injury Severity Score (TRISS), first developed in 1987 [@triss] and most recently updated in 2010 <!-- Cite 2009 revision ref -->, is one of the most commonly used models worldwide.TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss].
-The resulting value is a percentage that estimates the probability of survival (Ps).
+In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations.
+NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
+NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.
 
-<!-- Triss Limitations -->
+## Trauma severity scores as indicators for peer review
 
-<!-- NORMIT -->
-
-<!-- Explain why TSM ⇒ M&M -->
-
- <!--Kolla exv Strömmers och Ghorbanis pek,https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/29797712/ .
-De försöker använda bl.aa normit och triss för att se korrelans till OFI och preventable deaths.
-Skillnaden mot det du gör är att de bara tittar på patienter som dör och inte hela cohorten.
-Du tittar också på ett betydligt större dataunderlag.
-Så ditt bygger vidare på deras arbete-->
-
- However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is poorly understood.
-If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
-<!--Introduktionen är mycket bra!
-Egentligen bara fortsätta som du gör, läs på och lägg till meningar/källor när du hittar info.
-För att nå KI:s sidkrav kan du skriva en hel del om de olika Scorsen -->
+Some clinics already use TSS as audit filters when selecting cases for M&M review [@whoguidelines].
+However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood.
+A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
+However, that study included only deaths and not other morbidities and thereby had a low sample size.
+It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.
+If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
+The aim of this study is to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
 

### 2022-09-12 studyprogressreport.Rmd
progress on method

@@
 Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
-However, it has been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still plenty of progress to be made in improving these programs.
+However, it has also been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
 
@@
 The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
-Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqpianglo].
+Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
 The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
@@
 In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations.
-NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
 NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
@@
 The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases will be linked and then analysis will be performed to estimate the association between the common TSS models TRISS and NORMIT <!-- And something else?
---> and OFI using logistic regression.
+These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression.
 The predictive performance of these models will be evaluated using measures of discrimination and calibration.
@@
 KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
-KUH is the primary trauma centre for the region of Stockholm.
-In addition, KUH has agreements with several other regions to provide trauma care for patients with severe trauma.
+KUH is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma.
 KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
@@
 
-The trauma care quality register is an internal hospital register and registers the outcome of M&M conferences.
-Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
+The TCQD is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
 
@@
 
-Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
+Participants include all patients registered in both the trauma registry and the TCQD at KUH.
+<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> Furthermore, patients <15 years of age will be excluded as paediatric trauma is a seperate clinical entity.
 
@@
 The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models.
-Both of these models estimate the Ps based on expected mortality within 30 days.
-This is expressed a percentage and is a continuous variable.
+Both of these models estimate the Ps.
+This is expressed a probability and is a continuous variable.
 The variables integrated into these models are taken from arrival at hospital.
@@
 
-The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
-The NISS differs from the ISS in that the NISS is the sum of the squares from only the three most injured anatomical regions.
+The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
 
@@
 The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
-
-<!-- Update these numbers later when I have real numbers -->
+The exact study size has not yet been established at the time of writing this report.
 
@@
 ----------------------
-<!--Om du behöver "skapa" ny data genom att manipulera den insamlade (tvek i ditt fall) så ska du beskriva det här. Annars kan detta stycke slås ihop med variables ovan. -->
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
 
-### Exposures
-
-<!--Om du inte tydligt definerat modellernas variabler behöver du göra det här, alltså för exv triss som är GCS, RR, BT + ISS osv osv-->
-<!-- Really not sure exactly what to put here
--->
+The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
 
@@
 Conversion and handling of variables will be carried out according to the SweTrau manual.
+This is a complete case study.
+In cases where data necessary for analysis is missing, i.e.
+the variables referred to above, the participant will be excluded from the study.
 Results will be presented with a confidence interval of 95%.
 A p-value of <0.05 will be considered significant.
-<!-- Jo det kommer nog bli ccs, du kommer exkludera alla patienter som delsa saknar outcome, OFI alternativt där det saknas nog data för att bygga olika scores. -->
-<!-- Complete case study? -->
-<!--
-
-Should answer these questions, where applicable:
-
-* All statistical methods, incl. those used to control for confounding
-* Any methods to examine subgroups and interactions
-* How was missing data addressed?
-* Cohort study: If applicable, explain how loss to follow-up was addressed 
-* Any sensitivity analyses? -->
 
 Ethical considerations
-----------------------
+======================
 

### 2022-09-15 studyprogressreport.Rmd
attempting to make intro longer

@@
 * KUH - Karolinska University Hospital
-* KUH TCQD - Karolinska University Hospital Trauma Care Quality Database
-* KUH TR - Karolinska University Hospital Trauma Registry
 * NORMIT - Norwegian Survival Prediction Model in Trauma
@@
 * Ps - Probability of survival
+* TCQD - Karolinska University Hospital Trauma Care Quality Database
 * TRISS - Trauma and Injury Severity Score
@@
 Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
-However, it has also been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
+Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath].
+However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
 
@@
 
-The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
-M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The incidence of OFI function in this manner as a proxy for incidence of suboptimal care.
+The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
+M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
 The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
@@
 
-The value of M&M conferences in long-term quality improvement is dependent on selection of appropriate cases for review.
+The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
 The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
@@
 The latter benefit is not insignificant considering the large caseload found in most urban trauma centres.
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
 Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo].
 Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
-
-<!-- Explain the what and why of TSM -->
+Another potential quality indicator, that is already used at some clinics [@whoguidelines], are trauma severity scores.
 
 ## Trauma severity scores
-Trauma severity scores (TSS) are used in trauma quality improvement programs and research for their usefulness in benchmarking and in comparing trauma severity between patients [@whoguidelines].
-The existence of these models is necessitated by the heterogeneous nature of trauma, due to the variety in its underlying causes, which affect both the type and severity of the injury.
-This heterogeneity poses challenges in comparing and evaluating trauma outcomes and has spurred the development of multiple TSS.
-TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced models have been developed for quality assurance and research purposes.
-Today there are hundreds of TSS models.
-These models differ in their methodology but all have the aim of translating injury severity into quantitative values.
+Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury.
+An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
+This question has spurred the development of multiple trauma severity scores (TSS).
+TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
+TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
+Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
+
+Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss].
+One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines].
+TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
 However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
-Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions.
-This weakness is one reason why new models continue to be developed and improved around the globe.
+Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality].
+This weakness is one reason why models continue to be developed and improved on.
+Another reason is the efforts to establish which risk factors best predict mortality; different models use different physiological and anatomical parameters.
 
-The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide.
+The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter].
 TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
@@
 The ISS gives a score based on the three anatomical regions with the most severe injury.
-A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries.
+A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries [@trissbetter].
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
-The RTS is difficult to apply in situations where the patient has already been intubated, which gives TRISS limited value in estimating Ps in certain populations.
+The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score can be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
 
-In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations.
+In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit].
 NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
 NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
-NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.
+Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to the ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter].
+NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS.
+For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit].
+NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit]
 
-## Trauma severity scores as indicators for peer review
+## Trauma severity scores as quality indicators
 
-Some clinics already use TSS as audit filters when selecting cases for M&M review [@whoguidelines].
+As already mentioned, some clinics use TSS as audit filters to aid in identifying cases for M&M review.
 However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood.
@@
 It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.
+
 If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-The aim of this study is to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
+Conversely, if TSS models are not associated with OFI, using TSS as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
+
+There are far more TSS models than the two described above.
+These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is the location of this institution.
+It seems reasonable to begin by investigating these two models.
+The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.
+
+<!-- It is possible that I will add other scores later, but in the interest of having this document done in time to submit to school, I'm going with TRISS & NORMIT for now.
+I mention under the "development and time plan" heading that I may add scores later -->
 
@@
 
-Participants include all patients registered in both the trauma registry and the TCQD at KUH.
-<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> Furthermore, patients <15 years of age will be excluded as paediatric trauma is a seperate clinical entity.
+Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
 
@@
 Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
+
+### Study-specific exclusion criteria
+Patients <15 years of age will be excluded as paediatric trauma is a separate clinical entity.
+Furthermore, this is a complete case study and all patients will be excluded where there is not enough data recorded in order to carry out the analysis.
 
@@
 Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury".
-The value given is points on a scale from 0 to 75.
+The values from the three anatomical regions with the most severe injury are combined into a score that is given is points on a scale from 0 to 75.
 The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate.
@@
 The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+
+NISS is based on the same variables as ISS.
+It differs from ISS in that it takes into account multiple injuries within one anatomical region and calculates a score based on the three most severe injuries regardless of anatomical region.
 
@@
 
-Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database.
-The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
+Since all eligible participants must exist in both the trauma registry and the TCQD, and the TCQD is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the TCQD.
+The TCQD includes around 2200 patients who were selected for review between 2014 and 2021.
+In order to establish the exact study size, the eligibility criteria need to first be applied.
 The exact study size has not yet been established at the time of writing this report.
@@
 The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
+The scores will be calculated according to the formulas given by the respective developers of each score.
 
@@
 Conversion and handling of variables will be carried out according to the SweTrau manual.
-This is a complete case study.
 In cases where data necessary for analysis is missing, i.e.
 the variables referred to above, the participant will be excluded from the study.
+Logistic regression will be used to estimate the predictive value of each TSS in regards to OFI.
 Results will be presented with a confidence interval of 95%.
@@
 Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death.
-Patients have the right to be removed from the register if they wish, and also have the right to a copy of data that is stored about them on request.
+Patients have the right to be removed from the register on request.
+They also have the right to a copy of data that is stored about them on request.
 
@@
 
+It is possible that scores other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to.
+I will decide this within the first week or two of starting my data analysis code.
+
 ```{r, echo=FALSE, out.width = "400px"}

### 2022-09-15 main.R

@@ -21,6 +21,8 @@
 noacsr::source_all_functions()
 
 ## Import data
-data <- import_data(test = TRUE)
+datasets <- import_data(test = TRUE)
+
+
 
 ## Whatever you do next, maybe clean data?

### 2022-09-15 studyprogressreport.Rmd
fixes, make slightly longer

@@
 ============
-<!-- Why trauma is a big deal -->
-
 Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
@@
 However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
+A large population worldwide stands to gain from the improvement of trauma outcomes.
+These gains benefit not only the individual but also a broader population by reducing the societal and healthcare burden of disability.
+Reducing the economic burden of trauma is particularly relevant in light of the fact that low- to middle-income countries make up the majority of poor trauma outcomes.
 
@@
 The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
-The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
-This weakness of M&M review has been subject to considerable efforts to develop strategies to standardise case selection.
+In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed.
+This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
+Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
 The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
@@
 Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-Another potential quality indicator, that is already used at some clinics [@whoguidelines], are trauma severity scores.
+Another potential audit filter, that is already used at some clinics [@whoguidelines], are trauma severity scores.
 
@@
 
+<!-- Potentially talk about other limitations of TRISS here? Feels somewhat unnecessary because the project isn't really about investigating every weakness of TRISS -->
+
 In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit].
@@
 For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit].
-NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit]
+NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].
 
@@
 
+<!-- Also feels like I maybe have too many different sources cited for the intro, ~22 feels like overkill? -->
+
 Methods 
@@
 
-KUH is located in Stockholm, Sweden.
+KUH is located in Solna, Sweden.
 KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
@@
 
-The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
+The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
+The inclusion criteria are given below.
+All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma clinic.
+Data is collected from the patient's electronic medical records.
+The database contains data on patients from 2012 to 2021.
+
 SweTrau follows the Utstein Trauma Template for reporting of data following major trauma.
@@
 
-The TCQD is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
+The TCQD is an internal hospital register and registers the outcome of M&M conferences.
+Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
+The database contains data on patients from 2014 to 2021.
 
@@
 
-Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
+Participants include all patients registered in both the trauma registry and the TCQD at KUH that were not excluded by study-specific exclusion criteria.
+<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
 
-### Selection for the trauma registry
+### Inclusion criteria for the trauma registry
 
@@
 
-### Selection for the TCQD
+### Inclusion criteria for the TCQD
 
@@
 Patients <15 years of age will be excluded as paediatric trauma is a separate clinical entity.
-Furthermore, this is a complete case study and all patients will be excluded where there is not enough data recorded in order to carry out the analysis.
+In addition, this is a complete case study and all patients will be excluded where there is not enough data recorded in order to carry out the analysis.
 
@@
 
+<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. -->
+
 #### NORMIT
@@
 The TCQD includes around 2200 patients who were selected for review between 2014 and 2021.
-In order to establish the exact study size, the eligibility criteria need to first be applied.
+In order to establish the exact study size, the study-specific exclusion criteria need to first be applied.
 The exact study size has not yet been established at the time of writing this report.

### 2022-09-15 studyprogressreport.Rmd
final fixes for today i swear

@@
 One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
-Nonfatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
 
@@
 The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
-Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo].
+Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
 Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-Another potential audit filter, that is already used at some clinics [@whoguidelines], are trauma severity scores.
+Another potential audit filter, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
 
@@
 This weakness is one reason why models continue to be developed and improved on.
-Another reason is the efforts to establish which risk factors best predict mortality; different models use different physiological and anatomical parameters.
+Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.
 
@@
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
-The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score can be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
+The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
 

### 2022-09-15 manuscript.Rmd
making a test table one

@@
 ---
-title: "Association between trauma severity models and opportunities for improvement"
+title: "Association between trauma severity models and opportunities for improvement: A retrospective cohort study"
 author: "Victoria Witthoff"
@@
 ```
-
-<!-- This is an R Markdown document. Markdown is a simple formatting
-syntax for authoring HTML, PDF, and MS Word documents. For more
-details on using R Markdown see http://rmarkdown.rstudio.com -->
-
-<!-- This is a comment and will not be present in the compiled
-document. Try "knitting" this document by clicking the knit button up
-to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
-the console and you should only see headings -->
-
-<!-- Structure and most comments are from the Strengthening the
-reporting of observational studies in epidemiology (STROBE) statement,
-see https://www.strobe-statement.org/. For more explanations see the
-paper Strengthening the Reporting of Observational Studies in
-Epidemiology (STROBE):explanation and elaboration.  -->
-
-<!-- Indicate the study’s design with a commonly used term in the
-title, for example: "Associations between resuscitation interventions
-and opportunities for improvement in adult trauma patients: A cohort
-study" -->
 
@@
 6-8 lines long. Support every statement with a citation -->
-
-You can cite document here like this [@exampleKey9999]. Open the file
-bibliography.bib to learn more.
 
@@
 
-You can include code in this document like this:
-
-```{r main, echo=FALSE}
-## source("main.R") ## This "imports" the main script file of your project and run any code in it
+```{r main, echo=FALSE, message = FALSE}
+  source("main.R") ## This "imports" the main script file of your project and run any code in it
+  as.data.frame(t1)
 ```
-
-You can also embed plots:
-
-```{r plot, echo=FALSE}
-plot(pressure)
-```
-
-You can also mix text and code, so called inline code, like this: `r 2+5`.
 
@@
 ==========
-
-<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2022-09-15 main.R

@@ -1,16 +1,16 @@
 ## Load 
 library("rofi")
 library("dplyr")
-library("labelled")
-library("tableone")
+library(labelled)
+library(table1)
+library(printr, quietly=TRUE)
 source("functions/clean_data.r")
+source("functions/make_table_one.r")
 noacsr::source_all_functions()
 
 
 ## Step 1: Import data
 datasets <- import_data(test = TRUE)
-
-
 
 ## Step 2: Merge data? 
 ## I did this ugly workaround to make the merge_data function in the rofi package work, but perhaps I'm not supposed to use this function
@@ -25,13 +25,11 @@ datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scramb
 
 merged_data <- merge_data(datasets)
 
-
 ## Step 3: Clean data, see functions/clean_data.r
 
-cleaned_data <- clean_data(merged_data)
-
+cd <- clean_data(merged_data)
 
 ## Step 4: Table 1. Patient demographics.
-## Age, sex, ISS, NISS, ASA, penetrating/blunt, what more?? 
 
+t1 <- make_table_one(cd)
 

### 2022-09-16 studyprogressreport.Rmd
study progress report ready for feedback

@@
 ---
-title: "Association between trauma severity models and opportunities for improvement: A retrospective cohort study"
+title: 'Association between trauma severity models and opportunities for improvement:
+  A retrospective cohort study'
 author: "Victoria Witthoff"
@@
   pdf_document: default
+  word_document: default
+  html_document:
+    df_print: paged
+    number_sections: true
 bibliography: bibliography.bib
 csl: vancouver.csl
+fontsize: 11pt
+linestretch: 1
+always_allow_html: yes
 ---
@@
 
+```{r main, echo=FALSE, message = FALSE}
+  source("main.R") ## This "imports" the main script file of your project and run any code in it
+```
+
 Abbreviations
@@
 * ASA-PS - American Society of Anesthesiology Physical Status
+* ISS - Injury Severity Score
 * KUH - Karolinska University Hospital
+* M&M - Morbidity and mortality
+* NISS - New Injury Severity Score
 * NORMIT - Norwegian Survival Prediction Model in Trauma
@@
 * Ps - Probability of survival
-* TCQD - Karolinska University Hospital Trauma Care Quality Database
+* RTS - Revised Trauma Score
 * TRISS - Trauma and Injury Severity Score
@@
 Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.
 
@@
 Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath].
-However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
-A large population worldwide stands to gain from the improvement of trauma outcomes.
-These gains benefit not only the individual but also a broader population by reducing the societal and healthcare burden of disability.
-Reducing the economic burden of trauma is particularly relevant in light of the fact that low- to middle-income countries make up the majority of poor trauma outcomes.
-
-## Morbidity and mortality conferences
-
-The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
-M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
-The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
-Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
-The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
-
-The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
-In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed.
-This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
-Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
-The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
-The latter benefit is not insignificant considering the large caseload found in most urban trauma centres.
-The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
-Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
-Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-Another potential audit filter, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
+However, the implementation of QI programs differs widely between clinics [@tqipanglo], with smaller, non-specialised hospitals tending to have less organised programs [@whoguidelines].
 
@@
 
+<!-- I know that a lot of the following details about TRISS/NORMIT are probably not really relevant to my aim, it's mainly here to fill out on length -->
+
 The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter].
@@
 
-<!-- Potentially talk about other limitations of TRISS here? Feels somewhat unnecessary because the project isn't really about investigating every weakness of TRISS -->
-
 In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit].
@@
 NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
-Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to the ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter].
+Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter].
 NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS.
@@
 
+## Morbidity and mortality conferences
+
+The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
+M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
+The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
+Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
+The WHO themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+
+The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
+In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed.
+This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
+The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
+The latter benefit is not insignificant considering the large caseload found in many urban trauma centres.
+Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
+Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
+Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
+
 ## Trauma severity scores as quality indicators
 
-As already mentioned, some clinics use TSS as audit filters to aid in identifying cases for M&M review.
+Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
 However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood.
@@
 There are far more TSS models than the two described above.
-These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is the location of this institution.
+These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located.
 It seems reasonable to begin by investigating these two models.
@@
 
-<!-- Also feels like I maybe have too many different sources cited for the intro, ~22 feels like overkill? -->
+<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? -->
 
@@
 
-<!-- Present key elements of study design -->
-This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD).
-The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database.
+The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
 These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression.
@@
 
-The TCQD is an internal hospital register and registers the outcome of M&M conferences.
+The trauma care quality database is an internal hospital register and registers the outcome of M&M conferences.
 Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
@@
 
-Participants include all patients registered in both the trauma registry and the TCQD at KUH that were not excluded by study-specific exclusion criteria.
+Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria.
 <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
@@
 
-### Inclusion criteria for the TCQD
+### Inclusion criteria for the trauma care quality database
 
-The TCQD contains patients from the trauma registry that were selected for M&M review.
+<!-- 
+``` {r, echo = FALSE,} 
+  ofi_diagram() ## Flowchart of the process of arriving at OFI 
+```
+Removed this b/c it only works in html, diagrammersvg doesn't seem to be possible to install
+-->
+
+The trauma care quality database contains patients from the trauma registry that were selected for M&M review.
 At KUH, all fatalities are automatically selected for M&M review.
@@
 
-<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. -->
+<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? -->
 
@@
 
-Since all eligible participants must exist in both the trauma registry and the TCQD, and the TCQD is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the TCQD.
-The TCQD includes around 2200 patients who were selected for review between 2014 and 2021.
+Since all eligible participants must exist in both the trauma registry and the trauma care quality database, and the trauma care quality database is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the trauma care quality database.
+The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
 In order to establish the exact study size, the study-specific exclusion criteria need to first be applied.
@@
 The scores will be calculated according to the formulas given by the respective developers of each score.
+
+<!-- BTW this is here because, according to me and what I see in the database, it seems like I'll have to calculate NORMIT/TRISS for everyone based on parameters in the database -->
 
@@
 They also have the right to a copy of data that is stored about them on request.
+<!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->
 
@@
 **Stockholm Research Ethics Review Board approval number 2021-02541 and 2021-03531.**
-
-<!--
-
-source("main.R")  "imports" the main script file of your project and run any code in it -->
 

### 2022-09-16 functions/clean_data.r

@@ -8,12 +8,15 @@
 ##  tra_Dodsfallanalysgenomford -- Maybe?? Difference between this and VK_avslutad?
 
 clean_data <- function(dirty_data) {
-
-  ## Remove rows with NA in required fields
-  cleaned_data <- dirty_data %>% filter_at(vars(ISS, NISS, pt_age_yrs, inj_dominant, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, VK_avslutad, Problemomrade_.FMP),all_vars(!is.na(.)))
   
-  ## Remove cases with no quality review (?, unclear if it's only VK_avslutad or also tra_Dodsfallsanalysgenomford, ask about this)
-  cleaned_data <- cleaned_data %>% filter(VK_avslutad == 'Ja' | VK_avslutad == 'ja')
+  ## Remove rows with NA in required fields
+  cleaned_data <- dirty_data %>% filter_at(vars(ISS, NISS, pt_age_yrs, inj_dominant, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, Problemomrade_.FMP),all_vars(!is.na(.)))
+  
+  ## Create OFI variable
+  cleaned_data$ofi <- create_ofi(cleaned_data)
+  
+  ## Remove cases where OFI is NA
+  cleaned_data <- cleaned_data %>% filter (ofi != "NA")
   
   ## Remove rows where required values are unknown & filter away age groups. Systolic blood pressure should prolly be >0, unless the patient is DOA? Same for RR.
   cleaned_data <- cleaned_data %>% filter(inj_dominant != 999 & ed_gcs_sum != 999 & ed_sbp_value > 0 & ed_rr_value > 0 & pt_asa_preinjury != 999 & pt_age_yrs >= 15)

### 2022-09-16 functions/make_table_one.r

@@ -1,0 +1,89 @@
+## Age
+## ISS, NISS, pt_age_yrs, inj_dominant, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, ofi, pt_asa_preinjury
+
+make_table_one <- function(cd) {
+
+  ## Renaming data labels
+  
+  cd$pt_Gender <- factor(
+    cd$pt_Gender,
+    levels = c(1, 2),
+    labels = c("Male", "Female")
+  )
+  
+  cd$res_survival <- factor(
+    cd$res_survival,
+    levels = c(1, 2),
+    labels = c("Yes", "No")
+  )
+  
+  cd$host_care_level <- factor(
+    cd$host_care_level,
+    levels = c(1,2,3,4,5), 
+    labels = c("Emergency Department", "General Ward", "Operating Theatre", "High Dependency Unit", "Intensive Care Unit")
+  )
+  
+  cd$inj_dominant <- factor(
+    cd$inj_dominant,
+    levels = c(1, 2),
+    labels = c("Blunt", "Penetrating"),
+  )
+
+  cd$ofi <- factor(
+    cd$ofi,
+    levels = c("Yes", "No"),
+    labels = c("Opportunity for improvement", "No opportunities for improvement"),
+  )
+  
+  cd$pt_asa_preinjury <- factor(
+    cd$pt_asa_preinjury,
+    levels = c(1,2,3,4),
+    labels = c(1,2,3,4)
+  )
+  
+  ## Renaming column labels
+  var_label(cd) <- list (
+    pt_age_yrs = "Age (years)",
+    pt_Gender = "Gender",
+    res_survival = "30 day survival",
+    host_care_level = "Highest level of care",
+    ISS = "ISS",
+    NISS = "NISS",
+    inj_dominant = "Dominating Type of Injury",
+    ed_gcs_sum = "GCS",
+    ed_rr_value = "Respiratory Rate",
+    ed_sbp_value = "Systolic blood pressure (mmHg)",
+    pt_asa_preinjury = "ASA score",
+    ofi = "Opportunity for improvement"
+  )
+
+  
+  ## Not being used
+  ## rndr <- function(x, name, ...) {
+  ##  if (!is.numeric(x)) return(render.categorical.default(x))
+  ##  what <- switch(name,
+  ##                 pt_age_years = "Mean (SD)",
+  ##                 ISS  = "Mean (SD)")
+  ##  parse.abbrev.render.code(c("", what))(x)
+  ## }
+
+
[…]

### 2022-09-16 main.R

@@ -1,22 +1,29 @@
-## Load 
-library("rofi")
-library("dplyr")
+## Load libaries
+library(rofi)
+library(dplyr)
 library(labelled)
 library(table1)
 library(printr, quietly=TRUE)
-source("functions/clean_data.r")
-source("functions/make_table_one.r")
+library(DiagrammeR)
+library(miceadds) ## To make sourcing functions/ work
+library(trelliscope) ## To aid in making diagrams that work in .docx 
+
+## Sources all my functions, which are basically just code broken up into smaller parts
+## It looks like previous students put everything in one file (?), if you would prefer that let me know! No big deal to change
+## I did it like this to make it easier for me to see the overall workflow.
+source_function_path <- "~/R/trauma-severity-ofi/functions/"
+source.all(source_function_path, grepstring="*", print.source = FALSE)
+
 noacsr::source_all_functions()
 
-
 ## Step 1: Import data
+
 datasets <- import_data(test = TRUE)
 
-## Step 2: Merge data? 
-## I did this ugly workaround to make the merge_data function in the rofi package work, but perhaps I'm not supposed to use this function
-## It works tho! Testing it for now, I can do something else later if necessary
 
-## Make dataset names that match merge_data() requirements (lol)
+## Step 2: Merge data
+
+## Make dataset names that match merge_data() requirements
 datasets[['swetrau']] <- datasets$swetrau_scrambled
 datasets[['fmp']] <- datasets$fmp_scrambled
 datasets[['atgarder']] <- datasets$atgarder_scrambled
@@ -25,11 +32,12 @@ datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scramb
 
 merged_data <- merge_data(datasets)
 
+
 ## Step 3: Clean data, see functions/clean_data.r
 
 cd <- clean_data(merged_data)
 
+
 ## Step 4: Table 1. Patient demographics.
 
 t1 <- make_table_one(cd)
-

### 2022-09-17 studyprogressreport.Rmd
Comment on progress report

--- 2022-09-15 studyprogressreport.Rmd
@@
 * KUH - Karolinska University Hospital
+* KUH TCQD - Karolinska University Hospital Trauma Care Quality Database
+* KUH TR - Karolinska University Hospital Trauma Registry
 * NORMIT - Norwegian Survival Prediction Model in Trauma
@@
 * Ps - Probability of survival
-* TCQD - Karolinska University Hospital Trauma Care Quality Database
 * TRISS - Trauma and Injury Severity Score
@@
 ============
+<!-- Why trauma is a big deal -->
+
 Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
-The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
+The global health burden of physical injury<!-- I typically define trauma as the clinical entity composed of physical injury and the body's associated response.
+I would suggest that you stick to trauma and don't use physical injury and trauma interchangeably --> has led to trauma being described as a worldwide pandemic.
 One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
@@
 Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
-Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath].
-However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
-A large population worldwide stands to gain from the improvement of trauma outcomes.
-These gains benefit not only the individual but also a broader population by reducing the societal and healthcare burden of disability.
-Reducing the economic burden of trauma is particularly relevant in light of the fact that low- to middle-income countries make up the majority of poor trauma outcomes.
+However, it has also been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.
 
@@
 
-The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
-M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
+The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
+M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The incidence of OFI function in this manner as a proxy for incidence of suboptimal care.
 The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
@@
 
-The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
-In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed.
-This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
-Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+The value of M&M conferences in long-term quality improvement is dependent on selection of appropriate cases for review.
+The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+This weakness of M&M review has been subject to considerable efforts to develop strategies to standardise case selection.
 The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
 The latter benefit is not insignificant considering the large caseload found in most urban trauma centres.
-The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
 Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo].
 Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-Another potential audit filter, that is already used at some clinics [@whoguidelines], are trauma severity scores.
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
+
+<!-- Explain the what and why of TSM -->
 
 ## Trauma severity scores
-Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury.
-An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
-This question has spurred the development of multiple trauma severity scores (TSS).
-TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
-TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
-Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
+Trauma severity scores (TSS)<!-- Avoid non-standard abbreviations, as it makes your text harder to read.
+For example, I don't think spelling out trauma severity scores is such a big deal, especially considering that you'll be dealing with so many score names that are abbreviations themselves -->  are used in trauma quality improvement programs and research for their usefulness in benchmarking and in comparing trauma severity between patients [@whoguidelines].
+The existence of these models is necessitated by the heterogeneous nature of trauma, due to the variety in its underlying causes, which affect both the type and severity of the injury.
+This heterogeneity poses challenges in comparing and evaluating trauma outcomes and has spurred the development of multiple TSS.
+TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced models have been developed for quality assurance and research purposes.
+Today there are hundreds of TSS models.
+These models differ in their methodology but all have the aim of translating injury severity into quantitative values.
+However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
+Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions.
+This weakness is one reason why new models continue to be developed and improved around the globe.
 
-Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss].
-One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines].
-TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
-However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
-Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality].
-This weakness is one reason why models continue to be developed and improved on.
-Another reason is the efforts to establish which risk factors best predict mortality; different models use different physiological and anatomical parameters.
-
-The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter].
+The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide.
 TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
@@
 The ISS gives a score based on the three anatomical regions with the most severe injury.
-A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries [@trissbetter].
+A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries.
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
-The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score can be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
+The RTS is difficult to apply in situations where the patient has already been intubated, which gives TRISS limited value in estimating Ps in certain populations.
 
-<!-- Potentially talk about other limitations of TRISS here? Feels somewhat unnecessary because the project isn't really about investigating every weakness of TRISS -->
-
-In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit].
+In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations.
 NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
 NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
-Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to the ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter].
-NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS.
-For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit].
-NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].
+NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.
 
-## Trauma severity scores as quality indicators
+## Trauma severity scores as indicators for peer review
 
-As already mentioned, some clinics use TSS as audit filters to aid in identifying cases for M&M review.
+Some clinics already use TSS as audit filters when selecting cases for M&M review [@whoguidelines].
 However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood.
-A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
+A study at this institution<!-- This institution?
+--> found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
 However, that study included only deaths and not other morbidities and thereby had a low sample size.
 It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.
+If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
+The aim of this study is to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
 
-If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-Conversely, if TSS models are not associated with OFI, using TSS as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
-
-There are far more TSS models than the two described above.
-These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is the location of this institution.
-It seems reasonable to begin by investigating these two models.
-The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.
-
-<!-- It is possible that I will add other scores later, but in the interest of having this document done in time to submit to school, I'm going with TRISS & NORMIT for now.
-I mention under the "development and time plan" heading that I may add scores later -->
-
-<!-- Also feels like I maybe have too many different sources cited for the intro, ~22 feels like overkill? -->
+<!-- Note that you mix trauma severity scores, trauma severity models, prediction models, etc. Stick to one of these concepts -->
 
@@
 <!-- Present key elements of study design -->
-This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD).
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD)<!-- I would avoid abbreviating these databases as well-->.
 The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
@@
 
-KUH is located in Solna, Sweden.
+KUH is located in Stockholm, Sweden.
 KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
@@
 
-The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
-The inclusion criteria are given below.
-All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma clinic.
-Data is collected from the patient's electronic medical records.
-The database contains data on patients from 2012 to 2021.
-
+The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines.
 SweTrau follows the Utstein Trauma Template for reporting of data following major trauma.
@@
 
-The TCQD is an internal hospital register and registers the outcome of M&M conferences.
-Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
-The database contains data on patients from 2014 to 2021.
+The TCQD is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
 
@@
 
-Participants include all patients registered in both the trauma registry and the TCQD at KUH that were not excluded by study-specific exclusion criteria.
-<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
+Participants include all patients registered in both the trauma registry and the TCQD at KUH.
+<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> <!-- Don't include numbers here,  that's for Results --> Furthermore, patients <15 years of age will be excluded as paediatric trauma is a seperate clinical entity.
 
-### Inclusion criteria for the trauma registry
+### Selection<!-- By selection, do you mean eligibility criteria?--> for the trauma registry
 
@@
 
-### Inclusion criteria for the TCQD
+### Selection for the TCQD
 
@@
 Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
-
-### Study-specific exclusion criteria
-Patients <15 years of age will be excluded as paediatric trauma is a separate clinical entity.
-In addition, this is a complete case study and all patients will be excluded where there is not enough data recorded in order to carry out the analysis.
 
@@
 The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH.
-OFI is coded as either "Yes - at least one opportunity for improvement" or "No - no opportunities for improvement".
+OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
 
@@
 Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury".
-The values from the three anatomical regions with the most severe injury are combined into a score that is given is points on a scale from 0 to 75.
+The value given is points on a scale from 0 to 75.
 The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate.
@@
 
-<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. -->
-
 #### NORMIT
@@
 The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
-
-NISS is based on the same variables as ISS.
-It differs from ISS in that it takes into account multiple injuries within one anatomical region and calculates a score based on the three most severe injuries regardless of anatomical region.
 
@@
 
-Since all eligible participants must exist in both the trauma registry and the TCQD, and the TCQD is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the TCQD.
-The TCQD includes around 2200 patients who were selected for review between 2014 and 2021.
-In order to establish the exact study size, the study-specific exclusion criteria need to first be applied.
+Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database.
+The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
 The exact study size has not yet been established at the time of writing this report.
@@
 The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
-The scores will be calculated according to the formulas given by the respective developers of each score.
 
@@
 Conversion and handling of variables will be carried out according to the SweTrau manual.
+This is a complete case study.
 In cases where data necessary for analysis is missing, i.e.
 the variables referred to above, the participant will be excluded from the study.
-Logistic regression will be used to estimate the predictive value of each TSS in regards to OFI.
 Results will be presented with a confidence interval of 95%.
 A p-value of <0.05 will be considered significant.
+<!-- The aim includes both to determine association between the severity estimates and OFI as well as how well they predict OFI.
+These two components require different methods.
+To determine the strength of the association you could use logistic regression, then the resulting OR would be your effect size.
+Because all these models returns a probability you don't need to standardise the scores to be able to compare the ORs across models to determine which score that is most strongly associated with OFI.
+To assess how well they predict OFI you need to calculate the discrimination (often expressed as the area under the receiver operating characteristics curve, or just AUC) and calibration (using for example the integrated calibration index.
+Note that this is slightly contrived, as that would equate the probability of death with the probability of OFI-->
 
@@
 Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death.
-Patients have the right to be removed from the register on request.
-They also have the right to a copy of data that is stored about them on request.
+Patients have the right to be removed from the register if they wish, and also have the right to a copy of data that is stored about them on request.
 
@@
 
-It is possible that scores other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to.
-I will decide this within the first week or two of starting my data analysis code.
-
 ```{r, echo=FALSE, out.width = "400px"}

### 2022-09-19 studyprogressreport.Rmd

--- 2022-09-16 studyprogressreport.Rmd
@@
 Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
-The global health burden of physical injury has led to trauma being described as a worldwide pandemic.
+The global health burden of physical injury has led to trauma being described as a worldwide pandemic<!-- I typically define trauma as the clinical entity composed of physical injury and the body's associated response.
+I would suggest that you stick to trauma and don't use physical injury and trauma interchangeably -->.
 One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
@@
 An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
-This question has spurred the development of multiple trauma severity scores (TSS).
+This question has spurred the development of multiple trauma severity scores (TSS)<!-- Avoid non-standard abbreviations, as it makes your text harder to read.
+For example, I don't think spelling out trauma severity scores is such a big deal, especially considering that you'll be dealing with so many score names that are abbreviations themselves -->.
 TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
@@
 
+<!-- Note that you mix trauma severity scores, trauma severity models, prediction models, etc. Stick to one of these concepts -->
+
 <!-- It is possible that I will add other scores later, but in the interest of having this document done in time to submit to school, I'm going with TRISS & NORMIT for now.
@@
 
-<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? -->
+<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? --> <!-- That's fine -->
 
@@
 Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria.
-<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->
+<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> <!-- Don't include numbers here, those belongs in Reults -->
 
@@
 
-<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail. Same for NORMIT below. Do I need to have the formulas? -->
+<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail.
+Same for NORMIT below.
+Do I need to have the formulas?
+--> <!-- Yes, you need the formulas, but not for this report -->
 
@@
 A p-value of <0.05 will be considered significant.
+
+<!-- The aim includes both to determine association between the severity estimates and OFI as well as how well they predict OFI.
+These two components require different methods.
+To determine the strength of the association you could use logistic regression, then the resulting OR would be your effect size.
+Because all these models returns a probability you don't need to standardise the scores to be able to compare the ORs across models to determine which score that is most strongly associated with OFI.
+To assess how well they predict OFI you need to calculate the discrimination (often expressed as the area under the receiver operating characteristics curve, or just AUC) and calibration (using for example the integrated calibration index.
+Note that this is slightly contrived, as that would equate the probability of death with the probability of OFI-->
 

### 2022-09-22 functions/clean_data.R
small fixes

@@ -1,0 +1,25 @@
+## All values that are necessary:
+##  For TRISS: ISS, pt_age_years, inj_dominant (blunt or penetrating), RTS: ed_gcs_sum (GCS on arrival at hosp), ed_sbp_value (systolic BP), ed_rr_value (RR)
+##  For NORMIT: NISS, pt_age_years, pt_asa_preinjury, RTS: ed_gcs_sum (GCS on arrival at hosp), ed_sbp_value (systolic BP), ed_rr_value (RR)
+##  For OFI: VK_avslutad, problemomrade_.FMP
+##  Interpretation of OFI stuff:
+##  VK_avslutad -- Whether quality review was done ja/nej
+##  Problemomrade_.FMP -- Whether there was OFI, OK = no OFI
+##  tra_Dodsfallanalysgenomford -- 
+##  Add numbers for numbers excluded at each step 
+##  So need to return a list? HMMMMM
+## For GCS = 99, take prehosp value, (?) exclude if no prehosp value
+
+clean_data <- function(dirty.data) {
+  
+  ## Remove rows with NA in required fields
+  cleaned.data <- dirty.data %>% filter_at(vars(ISS, NISS, pt_age_yrs, inj_dominant, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, Problemomrade_.FMP),all_vars(!is.na(.)))
+  
+  ## Remove cases where OFI is NA
+  cleaned.data <- cleaned.data %>% filter (ofi != "NA")
+  
+  ## Remove rows where required values are unknown & filter away age groups. Systolic blood pressure should prolly be >0, unless the patient is DOA? Same for RR.
+  cleaned.data <- cleaned.data %>% filter(inj_dominant != 999 & ed_gcs_sum != 999 & ed_sbp_value > 0 & ed_rr_value > 0 & pt_asa_preinjury != 999 & pt_age_yrs >= 15)
+  
+  return (cleaned.data)
+}

### 2022-09-22 functions/make_table_one.R

@@ -1,0 +1,87 @@
+## Age
+## ISS, NISS, pt_age_yrs, inj_dominant, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, ofi, pt_asa_preinjury
+
+make_table_one <- function(cd) {
+  
+  ## Renaming data labels
+  
+  cd$pt_Gender <- factor(
+    cd$pt_Gender,
+    levels = c(1, 2),
+    labels = c("Male", "Female")
+  )
+  
+  cd$res_survival <- factor(
+    cd$res_survival,
+    levels = c(1, 2),
+    labels = c("Yes", "No")
+  )
+  
+  cd$host_care_level <- factor(
+    cd$host_care_level,
+    levels = c(1,2,3,4,5), 
+    labels = c("Emergency Department", "General Ward", "Operating Theatre", "High Dependency Unit", "Intensive Care Unit")
+  )
+  
+  cd$inj_dominant <- factor(
+    cd$inj_dominant,
+    levels = c(1, 2),
+    labels = c("Blunt", "Penetrating"),
+  )
+  
+  cd$ofi <- factor(
+    cd$ofi,
+    levels = c("Yes", "No"),
+    labels = c("Opportunity for improvement", "No opportunities for improvement"),
+  )
+  
+  cd$pt_asa_preinjury <- factor(
+    cd$pt_asa_preinjury,
+    levels = c(1,2,3,4),
+    labels = c(1,2,3,4)
+  )
+  
+  ## Renaming column labels
+  var_label(cd) <- list (
+    pt_age_yrs = "Age (years)",
+    pt_Gender = "Gender",
+    res_survival = "30 day survival",
+    host_care_level = "Highest level of care",
+    ISS = "ISS",
+    NISS = "NISS",
+    inj_dominant = "Dominating Type of Injury",
+    ed_gcs_sum = "GCS",
+    ed_rr_value = "Respiratory Rate",
+    ed_sbp_value = "Systolic blood pressure (mmHg)",
+    pt_asa_preinjury = "ASA score",
+    ofi = "Opportunity for improvement"
+  )
+  
+  
+  ## Not being used
+  ## rndr <- function(x, name, ...) {
+  ##  if (!is.numeric(x)) return(render.categorical.default(x))
+  ##  what <- switch(name,
+  ##                 pt_age_years = "Mean (SD)",
+  ##                 ISS  = "Mean (SD)")
+  ##  parse.abbrev.render.code(c("", what))(x)
+  ## }
+  
+  
[…]

### 2022-09-22 functions/ofi_diagram.R

@@ -1,0 +1,36 @@
+## Makes a flowchart on how patients are selected for M&M
+
+ofi_diagram <- function() {
+  
+ofi_diagram <- DiagrammeR::grViz("digraph flowchart {
+  # node definitions with substituted label text
+  node [fontname = Helvetica, shape = rectangle]        
+  tab1 [label = '@@1']
+  tab2 [label = '@@2']
+  tab3 [label = '@@3']
+  tab4 [label = '@@4']
+  tab5 [label = '@@5']
+  tab6 [label = '@@6']
+  tab7 [label = '@@7'] 
+  
+  # edge definitions with the node IDs
+  tab1 -> tab2 -> tab3 -> tab4 -> tab5;
+  tab5 -> tab6;
+  tab5 -> tab7;
+}
+
+  [1]: 'Trauma team activation'
+  [2]: 'Inclusion in registry according to SweTrau criteria described in 3.3.3'
+  [3]: 'Audit filters and primary review by experienced nurseand audit filters'
+  [4]: 'Secondary review by another experienced nurse'
+  [5]: 'Multidisciplinary morbidity and mortality conference'
+  [6]: 'Opportunities for improvement found'
+  [7]: 'No opportunities for improvement found'
+") 
+
+## %>% DiagrammeRsvg::export_svg() %>% charToRaw() %>% rsvg::rsvg_pdf("swetrau_flowchart.pdf")
+## Cannot to install DiagrammeRsvg b/c can't install v8 b/c "fatal error: v8.h: no such file or directory"
+}
+  
+
+

### 2022-09-22 main.R

@@ -5,21 +5,16 @@ library(labelled)
 library(table1)
 library(printr, quietly=TRUE)
 library(DiagrammeR)
-library(miceadds) ## To make sourcing functions/ work
-library(trelliscope) ## To aid in making diagrams that work in .docx 
 
-## Sources all my functions, which are basically just code broken up into smaller parts
-## It looks like previous students put everything in one file (?), if you would prefer that let me know! No big deal to change
-## I did it like this to make it easier for me to see the overall workflow.
-source_function_path <- "~/R/trauma-severity-ofi/functions/"
-source.all(source_function_path, grepstring="*", print.source = FALSE)
+## list of libraries i tried to install but failed, mb try again later
+## rsvg/diagrammersvg, kableextra
 
+## reminder: needs to be .R i.e. capital R 
 noacsr::source_all_functions()
 
 ## Step 1: Import data
 
 datasets <- import_data(test = TRUE)
-
 
 ## Step 2: Merge data
 
@@ -30,14 +25,16 @@ datasets[['atgarder']] <- datasets$atgarder_scrambled
 datasets[['problem']] <- datasets$problem_scrambled
 datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scrambled
 
-merged_data <- merge_data(datasets)
+merged.data <- merge_data(datasets)
 
+## Step 3: Create OFI variable
+
+merged.data$ofi <- create_ofi(merged.data)
 
 ## Step 3: Clean data, see functions/clean_data.r
 
-cd <- clean_data(merged_data)
-
+cleaned.data <- clean_data(merged.data)
 
 ## Step 4: Table 1. Patient demographics.
 
-t1 <- make_table_one(cd)
+t1 <- make_table_one(cleaned.data)

### 2022-09-22 studyprogressreport.Rmd
updated according to martins comments, done

@@
 ============
-Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
-The global health burden of physical injury has led to trauma being described as a worldwide pandemic<!-- I typically define trauma as the clinical entity composed of physical injury and the body's associated response.
-I would suggest that you stick to trauma and don't use physical injury and trauma interchangeably -->.
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response <!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
+The global health burden of trauma has led to it being described as a worldwide pandemic.
 One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
@@
 
-## Trauma severity scores
+## Trauma severity models
 Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury.
 An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
-This question has spurred the development of multiple trauma severity scores (TSS)<!-- Avoid non-standard abbreviations, as it makes your text harder to read.
-For example, I don't think spelling out trauma severity scores is such a big deal, especially considering that you'll be dealing with so many score names that are abbreviations themselves -->.
-TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
-TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
-Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
+This question has spurred the development of multiple trauma severity models.
+Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
+Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
+Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
 
-Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss].
+Trauma severity models have become a cornerstone of modern trauma QI programs [@historytss].
 One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines].
-TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
-However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
+Trauma severity models also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
+However, an inherent weakness of any trauma severity model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
 Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality].
@@
 Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.
-
-<!-- I know that a lot of the following details about TRISS/NORMIT are probably not really relevant to my aim, it's mainly here to fill out on length -->
 
@@
 
-## Trauma severity scores as quality indicators
+## Trauma severity models as quality indicators
 
 Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
-However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood.
+However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
 A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
 However, that study included only deaths and not other morbidities and thereby had a low sample size.
-It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.
+It is not known whether trauma severity models have better performance when taking into account the entire population of trauma patients.
 
-If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-Conversely, if TSS models are not associated with OFI, using TSS as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
+If it could be found that an existing trauma severity model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
+Conversely, if trauma severity models are not associated with OFI, using trauma severity models as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
 
-There are far more TSS models than the two described above.
+There are far more trauma severity models than the two described above.
 These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located.
@@
 
-<!-- Note that you mix trauma severity scores, trauma severity models, prediction models, etc. Stick to one of these concepts -->
-
 <!-- It is possible that I will add other scores later, but in the interest of having this document done in time to submit to school, I'm going with TRISS & NORMIT for now.
 I mention under the "development and time plan" heading that I may add scores later -->
-
-<!-- Also feels like I maybe have too many different sources cited for the intro? Is ~22 sources overkill? Or maybe that's normal? --> <!-- That's fine -->
 
@@
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression.
+These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
 The predictive performance of these models will be evaluated using measures of discrimination and calibration.
@@
 Participants include all patients registered in both the trauma registry and the trauma care quality database at KUH that were not excluded by study-specific exclusion criteria.
-<!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> <!-- Don't include numbers here, those belongs in Reults -->
 
@@
 ### Exposures
-The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models.
+The primary exposures of interest are the trauma severity models TRISS and NORMIT.
 Both of these models estimate the Ps.
@@
 These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.
-
-<!-- I already described all this in the intro so not sure if I really need to repeat it here, or in how much detail.
-Same for NORMIT below.
-Do I need to have the formulas?
---> <!-- Yes, you need the formulas, but not for this report -->
 
@@
 
-The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
-The scores will be calculated according to the formulas given by the respective developers of each score.
-
-<!-- BTW this is here because, according to me and what I see in the database, it seems like I'll have to calculate NORMIT/TRISS for everyone based on parameters in the database -->
+NORMIT and TRISS will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
+The scores will be calculated according to the formulas given by the respective developers of each model.
 
@@
 the variables referred to above, the participant will be excluded from the study.
-Logistic regression will be used to estimate the predictive value of each TSS in regards to OFI.
+Logistic regression will be used to determine association between the estimates of the trauma severity models and OFI.
+To assess how well the models predict OFI measures of discrimination and calibration will be used.
 Results will be presented with a confidence interval of 95%.
@@
 
-It is possible that scores other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to.
+It is possible that models other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to.
 I will decide this within the first week or two of starting my data analysis code.

### 2022-09-23 manuscript.Rmd
a working flowchart for inclusion/exclusion numbers

@@
 ---
-title: "Association between trauma severity models and opportunities for improvement: A retrospective cohort study"
+title: 'Association between trauma severity models and opportunities for improvement:
+  A retrospective cohort study'
 author: "Victoria Witthoff"
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
+  word_document: default
 bibliography: bibliography.bib
@@
 ============
+
+```{r main, echo=FALSE, message = FALSE}
+  source("main.R") 
+  exclusion_flowchart(inclusion.counts)
+```
 
@@
 
-```{r main, echo=FALSE, message = FALSE}
-  source("main.R") ## This "imports" the main script file of your project and run any code in it
-  as.data.frame(t1)
-```
-
 Discussion

### 2022-09-23 functions/clean_data.R

@@ -17,7 +17,7 @@ clean_data <- function(dirty.data, numbers = FALSE) {
   inclusion.counts <- data.frame (step  = c("original"),
                     included = c(original.count),
                     excluded = c(0)
-  )
+                                  )
   
   ## OFI exclusion
   cleaned.data <- cleaned.data %>% filter (ofi != "NA")

### 2022-09-23 functions/exclusion_flowchart.R

@@ -1,0 +1,40 @@
+exclusion_flowchart <- function(inclusion.counts) {
+  
+  exclusion.flowchart <- grViz("digraph flowchart {
+    ## the real nodes
+    node [shape = rectangle, width = 8, fontsize = 18] 
+      1 [label = '@@1']
+      2 [label = '@@2']
+      3 [label = '@@3']
+      4 [label = '@@4']
+      5 [label = '@@5']
+      m1 [label = '@@6']
+      m2 [label = '@@7']
+      m3 [label = '@@8']
+      m4 [label = '@@9']
+    
+    ## the fake nodes to create arrows halfway
+    node [shape=none, width=0, height=0, label='']
+    p1 -> 2; p2 -> 3; p3 -> 4; p4 -> 5;
+      {rank=same; p1 -> m1}
+      {rank=same; p2 -> m2}
+      {rank=same; p3 -> m3}
+      {rank=same; p4 -> m4}
+
+    edge [dir=none]
+    1 -> p1; 2 -> p2; 3 -> p3; 4 -> p4;
+  }
+
+  [1]: paste0(inclusion.counts[1,2], ' patients in trauma registry')
+  [2]: paste0(inclusion.counts[2,2], ' patients')
+  [3]: paste0(inclusion.counts[3,2], ' patients')
+  [4]: paste0(inclusion.counts[4,2], ' patients')
+  [5]: paste0(inclusion.counts[5,2], ' patients included')
+  [6]: paste0('Excluded due to no outcome from M&M review \\n(n =  ', inclusion.counts[2,3], ')')
+  [7]: paste0('Excluded due to <15 years of age \\n(n = ', inclusion.counts[3,3], ')')
+  [8]: paste0('Excluded due to DOA \\n(n = ', inclusion.counts[4,3], ')')
+  [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n(n = ', inclusion.counts[5,3], ')')
+                               
+                               ")
+  return(exclusion.flowchart)
+}

### 2022-09-23 main.R

@@ -33,11 +33,13 @@ merged.data$ofi <- create_ofi(merged.data)
 cleaned.data <- clean_data(merged.data)
 
 ## This (in a very ugly way) returns data on how many patients are kept and excluded at each step 
-exclusion.numbers <- clean_data(merged.data, numbers = TRUE)
+inclusion.counts <- clean_data(merged.data, numbers = TRUE)
 
 ## add scores
 cleaned.data <- make_rts(cleaned.data)
 cleaned.data <- make_triss(cleaned.data)
 cleaned.data <- make_normit(cleaned.data)
 
-## exclusion_flowchart(exclusion.numbers)
+
+
+

### 2022-09-23 studyprogressreport.Rmd
made ofi_diagram() a bit better perhaps

@@
 output:
+  word_document: default
   pdf_document: default
-  word_document: default
   html_document:

### 2022-09-23 functions/exclusion_flowchart.R

@@ -1,8 +1,14 @@
 exclusion_flowchart <- function(inclusion.counts) {
   
   exclusion.flowchart <- grViz("digraph flowchart {
+  
+    label='Fig. x. Flow diagram for application of exclusion criteria';
+    labelloc=bottom;
+    labeljust=left;
+    labelfontsize=16;
+  
     ## the real nodes
-    node [shape = rectangle, width = 8, fontsize = 18] 
+    node [shape = rectangle, width = 8, fontsize = 16] 
       1 [label = '@@1']
       2 [label = '@@2']
       3 [label = '@@3']

### 2022-09-23 functions/ofi_diagram.R

@@ -2,34 +2,42 @@
 
 ofi_diagram <- function() {
   
-ofi_diagram <- DiagrammeR::grViz("digraph flowchart {
-  # node definitions with substituted label text
-  node [fontname = Helvetica, shape = rectangle]        
-  tab1 [label = '@@1']
-  tab2 [label = '@@2']
-  tab3 [label = '@@3']
-  tab4 [label = '@@4']
-  tab5 [label = '@@5']
-  tab6 [label = '@@6']
-  tab7 [label = '@@7'] 
+  ofi_diagram <- DiagrammeR::grViz("digraph flowchart {
   
-  # edge definitions with the node IDs
-  tab1 -> tab2 -> tab3 -> tab4 -> tab5;
-  tab5 -> tab6;
-  tab5 -> tab7;
-}
+    label='Fig. x. OFI decision-making workflow';
+    labelloc=bottom;
+    labeljust=left;
+    labelfontsize=16;
 
-  [1]: 'Trauma team activation'
-  [2]: 'Inclusion in registry according to SweTrau criteria described in 3.3.3'
-  [3]: 'Audit filters and primary review by experienced nurseand audit filters'
-  [4]: 'Secondary review by another experienced nurse'
-  [5]: 'Multidisciplinary morbidity and mortality conference'
-  [6]: 'Opportunities for improvement found'
-  [7]: 'No opportunities for improvement found'
-") 
+    # node definitions with substituted label text
+    node [shape = rectangle, width = 8, fontsize = 16] 
+    tab1 [label = '@@1']
+    tab2 [label = '@@2']
+    tab3 [label = '@@3']
+    tab4 [label = '@@4']
+    tab5 [label = '@@5']
+    tab6 [label = '@@6']
+    tab7 [label = '@@7'] 
+    
+    # edge definitions with the node IDs
+    tab1 -> tab2 -> tab3 -> tab4 -> tab5;
+    tab5 -> tab6;
+    tab5 -> tab7;
+  }
+  
+    [1]: 'Trauma team activation'
+    [2]: 'Inclusion in registry according to SweTrau criteria'
+    [3]: 'Audit filters and primary review by experienced nurseand audit filters'
+    [4]: 'Secondary review by another experienced nurse'
+    [5]: 'Multidisciplinary morbidity and mortality conference'
+    [6]: 'Opportunities for improvement found'
[…]

### 2022-10-13 manuscript.Rmd
made ofi_diagram() a bit better perhaps
some progress on results

@@
 csl: vancouver.csl
+always_allow_html: true
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+```{r main, echo=FALSE, message = FALSE}
+  source("main.R") 
 ```
@@
 
-```{r main, echo=FALSE, message = FALSE}
-  source("main.R") 
-  exclusion_flowchart(inclusion.counts)
-```
-
 <!-- Explain the scientific background and rationale for the
@@
 ------------
+<!-- Cohort study: Give the eligibility criteria, and the sources and
+methods of selection of participants. -->
+Participants were patients registered in both the trauma registry and the trauma care quality database at KUH.
+The eligibility criteria for this study were the criteria for the two databases used in addition to some study-specific inclusion criteria.
 
-<!-- Cohort study: Give the eligibility criteria, and the sources and
-methods of selection of participants. Describe methods of
-follow-up. For matched studies, give matching criteria and number of
-exposed and unexposed -->
+### Inclusion criteria for the trauma registry
 
-<!-- Case-control study: Give the eligibility criteria, and the
-sources and + methods of case ascertainment and control
-selection. Give the rationale for the choice of cases and
-controls. For matched studies, give matching criteria and the number
-of controls per case -->
+The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated.
+Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15.
+Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.
 
-<!-- Cross-sectional study: Give the eligibility criteria, and the
-sources and methods of selection of participants -->
+### Inclusion criteria for the trauma care quality database
+
+The trauma care quality database contains patients from the trauma registry that were selected for M&M review.
+At KUH, all fatalities are automatically selected for M&M review.
+In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care.
+These nurses also use audit filters that automatically highlight cases with abnormal parameters.
+Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
+Figure 1 gives an overview of the workflow that leads to determination of OFI.
+
+``` {r, echo = FALSE,} 
+  ofi_diagram() ## Only works in HTML
+```
+
+### Study-specific exclusion criteria
+Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity.
+Patients that were dead on arrival were also excluded.
+In addition, this is a complete case study and all patients that lacked parameters necessary to carry out the analysis were excluded.
 
@@
 
-<!--
-
 Participants
 ------------
-
+<!-- 
 (a) Report numbers of individuals at each stage of study—eg numbers
@@
 
+-->
+A total of `r counts[6,3]` patients were excluded during processing of data.
+`r counts[2,3]` patients were excluded due to not having a recorded outcome from an M&M conference.
+`r counts[3,3]` patients were excluded for being under 15 years of age.
+`r counts[4,3]` patients were excluded due to being dead on arrival at hospital.
+`r counts[5,3]` patients were excluded due to missing required data.
+Figure 2 shows the selection process for eligible participants.
+
+```{r, echo = FALSE}
+  exclusion_flowchart(counts) ## Only works in HTML
+```
+
 Descriptive data
@@
 
+<!--
 (a) Give characteristics of study participants (eg demographic,
@@
 (c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
+amount) -->
+
+A total of `r counts[5,2]` participants were identified as eligible for inclusion.
+Their demographical and clinical characteristics are described in Table 1.
+`r numbers['ofi.true']` cases had at least one OFI and `r numbers['ofi.false']` cases had no OFI.
+The mean age for all patients was `r numbers['mean.age']` and `r numbers['pc.male']`% were male.
+
+For patients with at least one OFI, the mean TRISS score was `r numbers['t.ofi.t']` and the mean NORMIT score was `r numbers['n.ofi.t']`.
+For patients with no OFI, the mean TRISS score was `r numbers['t.ofi.f']` and the mean NORMIT score was `r numbers['n.ofi.f']`.
+
+```{r, echo = FALSE} 
+  table.one ## will be improved later
+```
 
@@
 
-Cohort study — Report numbers of outcome events or summary measures
-over time
-
-Case-control study — Report numbers in each exposure category, or
-summary measures of exposure
-
-Cross-sectional study — Report numbers of outcome events or summary
-measures
+<!-- Cohort study — Report numbers of outcome events or summary measures
+over time -->
 
@@
 
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
-
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
+The odds ratio for TRISS was `r t.or["or"]` (95% CI: `r t.or["5 %"]`-`r t.or["95 %"]`). When plotting the ROC, the AUC was `r t.auc` (95% CI: `r t.auc.ci[1]`-`r t.auc.ci[3]`).
 
@@
 
-Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses 
--->
+<!-- Report other analyses done—eg analyses of subgroups and interactions,
+and sensitivity analyses -->
 

### 2022-10-13 functions/exclusion_flowchart.R

@@ -1,14 +1,14 @@
-exclusion_flowchart <- function(inclusion.counts) {
+exclusion_flowchart <- function(counts) {
   
   exclusion.flowchart <- grViz("digraph flowchart {
   
-    label='Fig. x. Flow diagram for application of exclusion criteria';
+    label='Fig. 2. Flow diagram for application of exclusion criteria';
     labelloc=bottom;
     labeljust=left;
     labelfontsize=16;
   
     ## the real nodes
-    node [shape = rectangle, width = 8, fontsize = 16] 
+    node [shape = rectangle, width = 8, fontsize = 18] 
       1 [label = '@@1']
       2 [label = '@@2']
       3 [label = '@@3']
@@ -31,15 +31,15 @@ exclusion_flowchart <- function(inclusion.counts) {
     1 -> p1; 2 -> p2; 3 -> p3; 4 -> p4;
   }
 
-  [1]: paste0(inclusion.counts[1,2], ' patients in trauma registry')
-  [2]: paste0(inclusion.counts[2,2], ' patients')
-  [3]: paste0(inclusion.counts[3,2], ' patients')
-  [4]: paste0(inclusion.counts[4,2], ' patients')
-  [5]: paste0(inclusion.counts[5,2], ' patients included')
-  [6]: paste0('Excluded due to no outcome from M&M review \\n(n =  ', inclusion.counts[2,3], ')')
-  [7]: paste0('Excluded due to <15 years of age \\n(n = ', inclusion.counts[3,3], ')')
-  [8]: paste0('Excluded due to DOA \\n(n = ', inclusion.counts[4,3], ')')
-  [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n(n = ', inclusion.counts[5,3], ')')
+  [1]: paste0(counts[1,2], ' patients in trauma registry')
+  [2]: paste0(counts[2,2], ' patients')
+  [3]: paste0(counts[3,2], ' patients')
+  [4]: paste0(counts[4,2], ' patients')
+  [5]: paste0(counts[5,2], ' patients included')
+  [6]: paste0('Excluded due to no outcome from M&M review \\n(n =  ', counts[2,3], ')')
+  [7]: paste0('Excluded due to <15 years of age \\n(n = ', counts[3,3], ')')
+  [8]: paste0('Excluded due to DOA \\n(n = ', counts[4,3], ')')
+  [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n(n = ', counts[5,3], ')')
[…]

### 2022-10-13 functions/make_table_one.R

@@ -1,50 +1,50 @@
 ## Age
-## iss, Niss, pt_age_yrs, dom.inj, ed_gcs_sum, ed_sbp_value, ed_rr_value, pt_asa_preinjury, ofi, pt_asa_preinjury
+## iss, Niss, pt_age_yrs, dom.inj, ed_gcs_sum, ed.sbp_value, ed.rr_value, pt_asa_preinjury, ofi, pt_asa_preinjury
 
 make_table_one <- function(cd) {
   
   ## Renaming data labels
   
-  cd$gender <- factor(
-    cd$gender,
+  df$gender <- factor(
+    df$gender,
     levels = c(1, 2),
     labels = c("Male", "Female")
   )
   
-  cd$survival <- factor(
-    cd$survival,
+  df$survival <- factor(
+    df$survival,
     levels = c(1, 2),
     labels = c("Yes", "No")
   )
   
-  cd$care.level <- factor(
-    cd$care.level,
+  df$care.level <- factor(
+    df$care.level,
     levels = c(1,2,3,4,5), 
     labels = c("Emergency Department", "General Ward", "Operating Theatre", "High Dependency Unit", "Intensive Care Unit")
   )
   
-  cd$dom.inj <- factor(
-    cd$dom.inj,
+  df$dom.inj <- factor(
+    df$dom.inj,
     levels = c(1, 2),
     labels = c("Blunt", "Penetrating"),
   )
   
-  cd$ofi <- factor(
-    cd$ofi,
-    levels = c("Yes", "No"),
+  df$ofi <- factor(
+    df$ofi,
+    levels = c("1", "0"),
     labels = c("Opportunity for improvement", "No opportunities for improvement"),
   )
   
-  cd$asa <- factor(
-    cd$asa,
+  df$asa <- factor(
+    df$asa,
     levels = c(1,2,3,4),
     labels = c(1,2,3,4)
   )
   
   ## Renaming column labels
-  var_label(cd) <- list (
-    age = "Age (years)",
-    gender = "Gender",
+  var_label(df) <- list (
+    age = "Age — yrs",
+    gender = "Sex",
     survival = "30 day survival",
     care.level = "Highest level of care",
     iss = "ISS",
@@ -55,11 +55,20 @@ make_table_one <- function(cd) {
     ed.sbp = "Systolic blood pressure (mmHg)",
     asa = "ASA score",
     ofi = "Opportunity for improvement",
[…]

### 2022-10-13 functions/ofi_diagram.R

@@ -4,10 +4,10 @@ ofi_diagram <- function() {
   
   ofi_diagram <- DiagrammeR::grViz("digraph flowchart {
   
-    label='Fig. x. OFI decision-making workflow';
+    label='Fig. 1. OFI decision-making workflow';
     labelloc=bottom;
     labeljust=left;
-    labelfontsize=16;
+    labelfontsize=18;
 
     # node definitions with substituted label text
     node [shape = rectangle, width = 8, fontsize = 16] 

### 2022-10-13 functions/prep_data.R

@@ -47,10 +47,6 @@ prep_data <- function(df, numbers = FALSE) {
   ## Exclusion: ofi
   ##
   df <- df %>% filter (ofi != "NA")
-
-  ## Convert ofu values to 0 & 1 to make it easier later
-  df$ofi[df$ofi == "Yes"] <- 1
-  df$ofi[df$ofi == "No"] <- 0
   
   ## Add ofi inclusion/exclusion counts
   ofi.kept <- nrow(df)
@@ -103,6 +99,13 @@ prep_data <- function(df, numbers = FALSE) {
   total.excluded <- original.count - param.kept
   inclusion.counts[nrow(inclusion.counts) + 1,] = c("total", param.kept, total.excluded)
   
+  ##
+  ## Cleaning
+  ## 
+  
+  df$ofi[df$ofi == "Yes"] <- 1
+  df$ofi[df$ofi == "No"] <- 0
+
   ## 
   ## Testing
   ## 

### 2022-10-13 functions/stats.R

@@ -25,6 +25,5 @@ make_stats <-function(df) {
   
   ## return all the stats that we've created as a single list; this seemed more efficient than having multiple different functions that do similar things
   stats.ret <- list(model = model, perf = perf, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci)
-  
   return(stats.ret)
 }

### 2022-10-13 functions/table_one_stats.R

@@ -1,0 +1,25 @@
+
+table_one_stats <- function(df) {
+
+  numbers <- c()
+  numbers['ofi.true'] <- nrow(df[df$ofi == 1,])
+  numbers['ofi.false'] <- nrow(df[df$ofi == 0,])
+  numbers['mean.age'] <- round(mean(df$age), digits = 1)
+  numbers['pc.male'] <- round(nrow(df[df$gender == 1,]) / nrow(df) * 100, digits = 1)
+  numbers['t.mean'] <- round(mean(df$triss), digits = 2)
+  numbers['n.mean'] <- round(mean(df$normit), digits = 2)
+
+  t.numbers <- df %>% group_by(ofi) %>%
+    summarise(mean = round(mean(triss), digits = 2))
+  
+  numbers['t.ofi.t'] <- t.numbers[2,2]
+  numbers['t.ofi.f'] <- t.numbers[1,2]
+  
+  n.numbers <- df %>% group_by(ofi) %>%
+    summarise(mean = round(mean(normit), digits = 2))
+  
+  numbers['n.ofi.t'] <- n.numbers[2,2]
+  numbers['n.ofi.f'] <- n.numbers[1,2]
+  
+  return(numbers)
+}

### 2022-10-13 main.R

@@ -1,5 +1,7 @@
-## Libraries I tried to install but failed: rsvg/diagrammersvg, kableextra, phantomjs
-## Todo: CI for AUC
+## Libraries I tried to install but failed: rsvg/diagrammersvg, kableextra
+## Todo:
+## table1
+## finish unit tests
 library(rofi)
 library(dplyr)
 library(labelled)
@@ -14,6 +16,8 @@ library(caret)
 library(caTools)
 library(gmish)
 library(pROC)
+library(gt)
+library(webshot)
 
 noacsr::source_all_functions()
 
@@ -26,11 +30,11 @@ datasets[['atgarder']] <- datasets$atgarder_scrambled
 datasets[['problem']] <- datasets$problem_scrambled
 datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scrambled
 
-## setup data + store how many included/excluded
+## setup data
 df <- merge_data(datasets)
 pd <- prep_data(df)
 df <- pd[['df']]
-counts <- pd[['counts']]
+counts <- pd[['counts']] ## stores inclusion counts
 
 ## add scores
 df$rts <- apply(df, 1, make_rts)
@@ -52,18 +56,22 @@ df.normit <- df.normit %>% rename(score = normit)
 ## triss
 t.stats <- make_stats(df.triss)
 t.model <- t.stats[['model']]
-t.model.vals <- c(coeff = coef(summary(t.model))[2,1], pval = coef(summary(t.model))[2,4])
-t.roc <- t.stats[['perf']]
-t.roc.vals <- c(auc = t.stats[['auc']], auc.ci = t.stats[['auc.ci']], ici = t.stats[['ici']])
+t.auc <- round(t.stats[['auc']], digits = 2)
+t.auc.ci <- round(t.stats[['auc.ci']], digits = 2)
+## very confusing because level = 0.95 gives "97.5" and "2.5" in the output? at least i can easily change this later
+t.or <- exp(cbind("or" = coef(t.model), confint.default(t.model, level = 0.90)))
+t.or <- round(t.or[2,], digits = 2)
+t.perf <- t.stats[['perf']]
 t.acc <- t.stats[['acc']]
-plot(t.roc)
 
 ## normit
 n.stats <- make_stats(df.normit)
 n.model <- n.stats[['model']]
[…]

### 2022-10-17 manuscript.Rmd
updates to results

@@
 
+computed CIs for OR using wald estimator. 
 <!--
@@
 
-The odds ratio for TRISS was `r t.or["or"]` (95% CI: `r t.or["5 %"]`-`r t.or["95 %"]`). When plotting the ROC, the AUC was `r t.auc` (95% CI: `r t.auc.ci[1]`-`r t.auc.ci[3]`).
-
-Other analyses 
---------------
-
-<!-- Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses -->
+The odds ratio for TRISS was `r t.stats[['or']]` (95% CI: `r t.stats[['or.ci']]`, p-value: `r t.stats[['or.p']]`).
+When plotting the ROC, the AUC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
+The odds ratio for NORMIT was `r n.stats[['or']]` (95% CI: `r n.stats[['or.ci']]`, p-value: `r n.stats[['or.p']]`).
+When plotting the ROC, the AUC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
 

### 2022-10-17 functions/stats.R

@@ -17,13 +17,20 @@ make_stats <-function(df) {
   perf <- performance(pred, "tpr", "fpr")
   
   auc <- unlist(slot(performance(pred, "auc"), "y.values"))
+  auc <- round(auc, digits = 2)
   acc <- performance(pred, measure = "acc")
   ici <- ici(pred.model, df$ofi)
   
   proc <- roc(ofi ~ score, data = df)
-  auc.ci <- ci.auc(proc)
+  auc.ci <- round(ci.auc(proc), digits = 2)
+  
+  ## very confusing because level = 0.95 gives "97.5" and "2.5" in the output? at least i can easily change this later
+  or <- round(exp(cbind("or" = coef(model), confint.default(model, level = 0.90))), digits = 2)
+  or.ci <- paste(or[2,2], "-", or[2,3])
+  or <- or[2,1]
+  or.p <- summary(model)$coefficients[2,4]
   
   ## return all the stats that we've created as a single list; this seemed more efficient than having multiple different functions that do similar things
-  stats.ret <- list(model = model, perf = perf, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci)
+  stats.ret <- list(model = model, perf = perf, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci, or = or, or.p = or.p, or.ci = or.ci)
   return(stats.ret)
 }

### 2022-10-17 main.R

@@ -47,36 +47,25 @@ df$triss <- as.numeric(df$triss)
 df$normit <- as.numeric(df$normit)
 
 ## some subsetted data to make be friendly to stats functions
-df.triss <- df[, c("ofi", "triss")]
-df.triss <- df.triss %>% rename(score = triss)
-df.normit <- df[, c("ofi", "normit")]
-df.normit <- df.normit %>% rename(score = normit)
+t.df <- df[, c("ofi", "triss")] %>% rename(score = triss)
+n.df <- df[, c("ofi", "normit")] %>% rename(score = normit)
 
 ## make_stats returns a named list with all the things i want; then assign all list members to individual variables for ease of use later
 ## triss
-t.stats <- make_stats(df.triss)
+t.stats <- make_stats(t.df)
 t.model <- t.stats[['model']]
-t.auc <- round(t.stats[['auc']], digits = 2)
-t.auc.ci <- round(t.stats[['auc.ci']], digits = 2)
-## very confusing because level = 0.95 gives "97.5" and "2.5" in the output? at least i can easily change this later
-t.or <- exp(cbind("or" = coef(t.model), confint.default(t.model, level = 0.90)))
-t.or <- round(t.or[2,], digits = 2)
 t.perf <- t.stats[['perf']]
 t.acc <- t.stats[['acc']]
 
 ## normit
-n.stats <- make_stats(df.normit)
+n.stats <- make_stats(n.df)
 n.model <- n.stats[['model']]
-n.auc <- round(n.stats[['auc']], digits = 2)
-n.auc.ci <- round(n.stats[['auc.ci']], digits = 2)
-n.or <- exp(cbind("or" = coef(n.model), confint.default(n.model, level = 0.90)))
-n.or <- round(n.or[2,])
 n.perf <- n.stats[['perf']]
 n.acc <- n.stats[['acc']]
 
 ## things that can be plotted
 ## models: plot(t.model)
-## roc: plot(t.roc)
+## roc: plot(t.perf)
 ## accuracy: plot(t.acc)
 
 table.one <- make_table_one(df)

### 2022-10-18 manuscript.Rmd
plots

@@
 When plotting the ROC, the AUC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
+The integrated calibration index for TRISS was `r t.stats[['ici']]`.
+
 The odds ratio for NORMIT was `r n.stats[['or']]` (95% CI: `r n.stats[['or.ci']]`, p-value: `r n.stats[['or.p']]`).
 When plotting the ROC, the AUC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
+The integrated calibration index for NORMIT was `r n.stats[['ici']]`.
+
+`r plot(t.perf, main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE) `
+`r plot(n.perf, main="NORMIT Receiver Operating Characteristic Curve", colorize = TRUE) `
+
+`r plot(t.acc, main="TRISS Accuracy", colorize = TRUE) `
+`r plot(n.acc, main="NORMIT Accuracy", colorize = TRUE) `
 

### 2022-10-18 functions/stats.R
fix?

@@ -20,6 +20,7 @@ make_stats <-function(df) {
   auc <- round(auc, digits = 2)
   acc <- performance(pred, measure = "acc")
   ici <- ici(pred.model, df$ofi)
+  ici <- round(ici, digits = 2)
   
   proc <- roc(ofi ~ score, data = df)
   auc.ci <- round(ci.auc(proc), digits = 2)
@@ -28,7 +29,7 @@ make_stats <-function(df) {
   or <- round(exp(cbind("or" = coef(model), confint.default(model, level = 0.90))), digits = 2)
   or.ci <- paste(or[2,2], "-", or[2,3])
   or <- or[2,1]
-  or.p <- summary(model)$coefficients[2,4]
+  or.p <- round(summary(model)$coefficients[2,4], digits = 2)
   
   ## return all the stats that we've created as a single list; this seemed more efficient than having multiple different functions that do similar things
   stats.ret <- list(model = model, perf = perf, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci, or = or, or.p = or.p, or.ci = or.ci)

### 2022-10-18 manuscript.Rmd
fix?
plots actually work

--- 2022-10-17 manuscript.Rmd
@@
 always_allow_html: true
+editor_options: 
+   chunk_output_type: inline
 ---
@@
 When plotting the ROC, the AUC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
+The integrated calibration index for TRISS was `r t.stats[['ici']]`.
+
 The odds ratio for NORMIT was `r n.stats[['or']]` (95% CI: `r n.stats[['or.ci']]`, p-value: `r n.stats[['or.p']]`).
 When plotting the ROC, the AUC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
+The integrated calibration index for NORMIT was `r n.stats[['ici']]`.
+
+`r t.plot.roc`
+`r t.plot.acc`
+
+`r n.plot.roc`
+`r n.plot.acc`
 

### 2022-10-18 main.R

@@ -72,3 +72,8 @@ table.one <- make_table_one(df)
 
 ## make some numbers to quote in results bc i dont see a way to extract them from table1
 numbers <- table_one_stats(df)
+
+t.plot.roc <- plot(t.perf, main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
+t.plot.acc <- plot(t.acc, main="TRISS Accuracy")
+n.plot.roc <- plot(n.perf, main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
+n.plot.acc <- plot(n.acc, main="TRISS Accuracy")

### 2022-10-18 manuscript.Rmd
made a table

@@
 
-Outcome data
-------------
+Detailed totals for missing data for each required parameter are given in table 2.
+The total number of missing parameters is greater than the number of participants excluded for missing parameters.
+This is because some patients lacked multiple required parameters.
 
-<!-- Cohort study — Report numbers of outcome events or summary measures
-over time -->
+```{r gt table, echo = FALSE}
+missing
+```
 
@@
 
-The odds ratio for TRISS was `r t.stats[['or']]` (95% CI: `r t.stats[['or.ci']]`, p-value: `r t.stats[['or.p']]`).
-When plotting the ROC, the AUC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
+The odds ratio for TRISS was `r t.or['or']` (95% CI: `r t.or[2]`-`r t.or[3]`, p-value: `r t.stats[['or.p']]`).
+The ROC is plotted in Figure 3.
+The AUC for the ROC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
 The integrated calibration index for TRISS was `r t.stats[['ici']]`.
+The accuracy curve is shown in Figure 4.
+The best accuracy was `r t.stats[['acc.nrs']][1]` and was found at the cutoff `r t.stats[['acc.nrs']][2]`.
 
-The odds ratio for NORMIT was `r n.stats[['or']]` (95% CI: `r n.stats[['or.ci']]`, p-value: `r n.stats[['or.p']]`).
-When plotting the ROC, the AUC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
+```{r, echo = FALSE}
+plot(t.stats[['perf']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+```
+
+```{r, echo = FALSE}
+t.plot.acc <- plot(t.stats[['acc']], main="Fig 4. TRISS Accuracy Curve.")
+```
+
+The odds ratio for NORMIT was `r n.or['or']` (95% CI: `r n.or[2]`-`r n.or[3]`, p-value: `r n.stats[['or.p']]`).
+The ROC is plotted in Figure 5.
+The AUC for the ROC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
 The integrated calibration index for NORMIT was `r n.stats[['ici']]`.
+The accuracy curve for NORMIT is shown in Figure 5.
+The best accuracy was `r n.stats[['acc.nrs']][1]` and was found at the cutoff `r n.stats[['acc.nrs']][2]`.
 
-`r t.plot.roc`
-`r t.plot.acc`
+```{r, echo = FALSE}
+plot(n.stats[['perf']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+```
 
-`r n.plot.roc`
-`r n.plot.acc`
+```{r, echo = FALSE}
+n.plot.acc <- plot(n.stats[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
+```
 

### 2022-10-18 functions/prep_data.R

@@ -72,16 +72,19 @@ prep_data <- function(df, numbers = FALSE) {
   ## Exclusion: parameters
   
   ## Jonatan wanted me to make a table of exactly how many were missing for each parameter
-  m.gcs <- sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999,]) ## This does not take into account patients removed where GCS == 99 & there's no prehospital GCS, fix later
-  m.asa <- sum(is.na(df$asa)) + nrow(df[df$asa == 999,])
-  m.rr <- sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0,])
-  m.sbp <- sum(is.na(df$ed.sbp)) + nrow(df[df$ed.sbp == 0,])
-  m.dominj <- sum(is.na(df$dom.inj)) + nrow(df[df$dom.inj == 999,])
+  ## Age is higher up, above the <15 exclusion
+  m.gcs <- sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),]) ## This does not take into account patients removed where GCS == 99 & there's no prehospital GCS, fix later
+  m.asa <- sum(is.na(df$asa)) + nrow(df[df$asa == 999 & !is.na(df$asa),])
+  m.rr <- sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0 & !is.na(df$ed.rr),])
+  m.sbp <- sum(is.na(df$ed.sbp)) + nrow(df[df$ed.sbp == 0 & !is.na(df$ed.sbp),])
+  m.dominj <- sum(is.na(df$dom.inj)) + nrow(df[df$dom.inj == 999 & !is.na(df$dom.inj),])
   m.age <- sum(is.na(df$age))
   m.iss <- sum(is.na(df$iss))
   m.niss <- sum(is.na(df$niss))
-  missing <- c(gcs = m.gcs, asa = m.asa, rr = m.rr, sbp = m.sbp, dom.inj = m.dominj, age = m.age, iss = m.iss, niss = m.niss)
-  
+  m.names <- c("GCS", "ASA", "RR", "SBP", "Dominant injury type", "Age", "ISS", "NISS")
+  m.numbers <- c(m.gcs, m.asa, m.rr, m.sbp, m.dominj, m.age, m.iss, m.niss)
+  missing <- tibble("Required parameter" = m.names, "Total no. cases without required parameter" = m.numbers)
[…]

### 2022-10-18 functions/stats.R

@@ -20,7 +20,9 @@ make_stats <-function(df) {
   acc <- performance(pred, measure = "acc") ## This is plottable
   ind <- which.max( slot(acc, "y.values")[[1]] ) ## Find the index of the highest accuracy
   acc.max <- slot(acc, "y.values")[[1]][ind] ## Stores the highest accuracy
-  cutoff = slot(acc, "x.values")[[1]][ind] ## Stores the cutoff at the highest accuracy
+  acc.max <- round(acc.max, digits = 2)
+  cutoff <- slot(acc, "x.values")[[1]][ind] ## Stores the cutoff at the highest accuracy
+  cutoff <- round(cutoff, digits = 2)
   acc.nrs <- c(acc.max = acc.max, co = cutoff) ## Make a vector to return
   
   ## This makes a vector with the OR and the upper/lower bounds of the CI

### 2022-10-18 main.R

@@ -14,6 +14,7 @@ library(caTools)
 library(gmish)
 library(pROC)
 library(gt)
+library(gtsummary)
 
 noacsr::source_all_functions()
 
@@ -30,6 +31,8 @@ datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scramb
 df <- merge_data(datasets)
 pd <- prep_data(df) ## returns a list with the prepped dataset, inclusion counts, and missing parameters
 df <- pd[['df']]
+
+## exclusion data
 counts <- pd[['counts']] ## inclusion/exclusion counts
 missing <- pd[['missing']] ## nr of rows with missing data for each parameter
 
@@ -50,7 +53,11 @@ n.df <- df[, c("ofi", "normit")] %>% rename(score = normit)
 
 ## make_stats returns a named list with all the things i want; see stats.R
 t.stats <- make_stats(t.df)
+t.or <- t.stats[['or']]
+t.or <- t.or[2,]
 n.stats <- make_stats(n.df)
+n.or <- n.stats[['or']]
+n.or <- n.or[2,]
 
 ## descriptive data
 table.one <- make_table_one(df)

### 2022-10-18 manuscript.Rmd
copied text from progress report; to be improved upon later

@@
 output:
-  html_document: default
+  html_document:
+    df_print: paged
+    number_sections: true
   pdf_document: default
@@
 csl: vancouver.csl
-always_allow_html: true
+fontsize: 11pt
+linestretch: 1
+always_allow_html: yes
 editor_options: 
@@
 
-Background
-----------
-
-Methods
--------
-
-Results
--------
-
-Conclusion
-----------
+Abbreviations
+=============
+* ASA-PS - American Society of Anesthesiology Physical Status
+* ISS - Injury Severity Score
+* KUH - Karolinska University Hospital
+* M&M - Morbidity and mortality
+* NISS - New Injury Severity Score
+* NORMIT - Norwegian Survival Prediction Model in Trauma
+* OFI - Opportunities for improvement
+* Ps - Probability of survival
+* QI - Quality Improvement
+* RTS - Revised Trauma Score
+* TRISS - Trauma and Injury Severity Score
+* WHO - World Health Organisation
 
@@
 ============
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response <!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
+The global health burden of trauma has led to it being described as a worldwide pandemic.
+One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi].
+Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.
 
-<!-- Explain the scientific background and rationale for the
-investigation being reported. End by stating the aim or specific
-objectives, including any prespecified hypotheses. The introduction
-should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
-6-8 lines long. Support every statement with a citation -->
+<!-- S wants me to write about the UN 2030 goals here -->
+
+Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
+Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
+Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath].
+However, the implementation of QI programs differs widely between clinics [@tqipanglo], with smaller, non-specialised hospitals tending to have less organised programs [@whoguidelines].
+
+## Trauma severity models
+Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury.
+An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
+This question has spurred the development of multiple trauma severity models.
+Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
+Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
+Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
+
+Trauma severity models have become a cornerstone of modern trauma QI programs [@historytss].
+One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines].
+Trauma severity models also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
+However, an inherent weakness of any trauma severity model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
+Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality].
+This weakness is one reason why models continue to be developed and improved on.
+Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.
+
+The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter].
+TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
+The resulting value is a percentage that estimates the probability of survival (Ps).
+The ISS gives a score based on the three anatomical regions with the most severe injury.
+A limitation of the ISS is that it doesn't account for situations where one anatomical region has multiple injuries [@trissbetter].
+The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
+The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
+
+In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit].
+NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region.
+Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter].
+NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS.
+For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit].
+NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].
+
+## Morbidity and mortality conferences
+
+The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
+M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
+The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
+Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
+The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+
+The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
+In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed.
+This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
+The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
+The latter benefit is not insignificant considering the large caseload found in many urban trauma centres.
+Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
+Some clinics have already made efforts to standardise case election; for example, by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
+Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
+
+## Trauma severity models as quality indicators
+
+Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
+However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
+A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
+However, that study included only deaths and not other morbidities and thereby had a low sample size.
+It is not known whether trauma severity models have better performance when applied to the entire population of trauma patients.
+
+If it could be found that an existing trauma severity model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
+Conversely, if trauma severity models are not associated with OFI, using trauma severity models as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
+
+There are far more trauma severity models than the two described above.
+These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located.
+It seems reasonable to begin by investigating these two models.
+
+<!-- S says to clarify the knowledge gap & motive -->
+
+## Aim
+The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.
+
+Methods 
+=======
 
@@
 
-<!-- Present key elements of study design -->
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database.
+The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
+These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
+The predictive performance of these models will be evaluated using measures of discrimination and calibration.
 
@@
 periods of recruitment, exposure, follow-up, and data collection -->
+
+KUH is located in Solna, Sweden.
+KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
+KUH is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma.
+KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
+
+The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
+The inclusion criteria are given below.
+All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma clinic.
+Data is collected from the patient's electronic medical records.
+The database contains data on patients from 2012 to 2021.
+
+SweTrau follows the Utstein Trauma Template for reporting of data following major trauma.
+The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
+Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
+
+The trauma care quality database is an internal hospital register and registers the outcome of M&M conferences.
+Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
+The database contains data on patients from 2014 to 2021.
 
@@
 
-Variables and data sources/measurements
----------------------------------------
+Variables
+---------
 
@@
 of assessment methods if there is more than one group -->
+### Outcome
+The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH.
+OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
+
+### Exposures
+The primary exposures of interest are the trauma severity models TRISS and NORMIT.
+Both of these models estimate the Ps.
+This is expressed a probability and is a continuous variable.
+The variables integrated into these models are taken from arrival at hospital.
+
+#### TRISS
+
+The TRISS model incorporates patient age, Injury Severity Score (ISS), Revised Trauma Score (RTS), and whether the trauma was blunt or penetrating.
+
+The ISS is given by estimating the severity of physical injuries by anatomical region.
+Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury".
+The values from the three anatomical regions with the most severe injury are combined into a score that is given is points on a scale from 0 to 75.
+The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate.
+Values are taken from arrival at hospital.
+These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.
+
+#### NORMIT
+
+The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.
+
+NISS is based on the same variables as ISS.
+It differs from ISS in that it takes into account multiple injuries within one anatomical region and calculates a score based on the three most severe injuries regardless of anatomical region.
 
@@
 
+The method and data analysis model will be written using simulated data to reduce the risk of research bias.
+Real-world data will only be used once the data analysis model is completed and found to work correctly on simulated data.
+
 Study size 
@@
 
+Since all eligible participants must exist in both the trauma registry and the trauma care quality database, and the trauma care quality database is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the trauma care quality database.
+The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
+In order to establish the exact study size, the study-specific exclusion criteria need to first be applied.
+The exact study size has not yet been established at the time of writing this report.
+
 Quantitative variables
@@
 
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
+NORMIT and TRISS will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
+The scores will be calculated according to the formulas given by the respective developers of each model.
 
@@
 
-computed CIs for OR using wald estimator. 
-<!--
+R, a programming language for statistical computing, will be used for compilation and analysis of data.
+Conversion and handling of variables will be carried out according to the SweTrau manual.
+In cases where data necessary for analysis is missing, i.e.
+the variables referred to above, the participant will be excluded from the study.
+Logistic regression will be used to determine association between the estimates of the trauma severity models and OFI.
+To assess how well the models predict OFI measures of discrimination and calibration will be used.
+Results will be presented with a confidence interval of 95%.
+A p-value of <0.05 will be considered significant.
 
-(a) Describe all statistical methods, including those used to control
-for confounding
+<!-- computed CIs for OR using wald estimator. -->
 
-(b) Describe any methods used to examine subgroups and interactions
+Ethical considerations
+==========
 
-(c) Explain how missing data were addressed
+All patient information used in this study is anonymised and stored in a secure database.
+Patient medical records will not be accessed.
+Patients were not asked to give consent for inclusion in the KUH trauma registry.
+The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations.
+Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death.
+Patients have the right to be removed from the register on request.
+They also have the right to a copy of data that is stored about them on request.
+<!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->
 
-(d)
+This project is a retrospective cohort study and the project has thereby had no effect on patient treatment.
+There is no risk that patients are harmed by inclusion in the study.
+This study has the potential to improve patient care by analysing a potential factor that could predict suboptimal trauma care.
+It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified by the fact that this population stands to gain the most from improved trauma care.
 
-Cohort study: If applicable, explain how loss to follow-up was addressed
-
-Case-control study:If applicable, explain how matching of cases and
-controls was addressed
-
-Cross-sectional study: If applicable, describe analytical methods
-taking account of sampling strategy
-
-(e) Describe any sensitivity analyses
--->
+**Stockholm Research Ethics Review Board approval number 2021-02541 and 2021-03531.**
 
@@
 ------------
-<!-- 
-(a) Report numbers of individuals at each stage of study—eg numbers
-potentially eligible, examined for eligibility, confirmed eligible,
-included in the study, completing follow-up, and analysed
-
-(b) Give reasons for non-participation at each stage
-
-(c) Consider use of a flow diagram
-
--->
 A total of `r counts[6,3]` patients were excluded during processing of data.
@@
 ----------------
-
-<!--
-(a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
-
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount) -->
-
 A total of `r counts[5,2]` participants were identified as eligible for inclusion.

### 2022-10-25 manuscript.Rmd

@@
 ============
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response <!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi].
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015]<!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]<!-- Cite the most recent Global Burden of Disease study instead-->.
 The global health burden of trauma has led to it being described as a worldwide pandemic.
@@
 Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
-Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).
+Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps)<!-- Consider not abbreviating Ps, as it's quite short anyway and is a "non-standard" abbreviation-->.
 
@@
 The ISS gives a score based on the three anatomical regions with the most severe injury.
-A limitation of the ISS is that it doesn't account for situations where one anatomical region has multiple injuries [@trissbetter].
+A limitation of the ISS is that it doesn't<!-- Don't use contractions --> account for situations where one anatomical region has multiple injuries [@trissbetter].
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
@@
 M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths<!-- I would not include preventable deaths here-->.
 The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
@@
 However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
-A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
+A study at this institution<!-- You haven't mentioned "this institution" before, so it's hard to understand what you refer to.
+I suggest that state the actual centre--> found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
 However, that study included only deaths and not other morbidities and thereby had a low sample size.
@@
 
-Methods 
-=======
-
 Study design
@@
 
-This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database.
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH)<!-- Use the full hospital name-->: the trauma registry and the trauma care quality database.
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
@@
 
-The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
+The trauma registry at KUH is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
 The inclusion criteria are given below.
-All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma clinic.
+All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma centre.
 Data is collected from the patient's electronic medical records.
-The database contains data on patients from 2012 to 2021.
+The database<!-- Trauma registry?--> contains data on patients from 2012 to 2021.
 
@@
 The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
-Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.
+Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS<!-- Do you mean GOS?--> at discharge, highest level of care, and transfers to other units.
 
@@
 ### Study-specific exclusion criteria
-Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity.
+Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity<!-- And are not part of the same review process-->.
 Patients that were dead on arrival were also excluded.
@@
 of assessment methods if there is more than one group -->
+
 ### Outcome
@@
 
-All patient information used in this study is anonymised and stored in a secure database.
+All patient information used in this study is pseudononymised and stored in a secure database.
 Patient medical records will not be accessed.

### 2022-10-25 manuscript.Rmd
fix based on martins comments

@@
 * ISS - Injury Severity Score
-* KUH - Karolinska University Hospital
 * M&M - Morbidity and mortality
@@
 * OFI - Opportunities for improvement
-* Ps - Probability of survival
 * QI - Quality Improvement
@@
 ============
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015]<!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]<!-- Cite the most recent Global Burden of Disease study instead-->.
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@globalburden].
 The global health burden of trauma has led to it being described as a worldwide pandemic.
@@
 Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
-Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps)<!-- Consider not abbreviating Ps, as it's quite short anyway and is a "non-standard" abbreviation-->.
+Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival.
 
@@
 The ISS gives a score based on the three anatomical regions with the most severe injury.
-A limitation of the ISS is that it doesn't<!-- Don't use contractions --> account for situations where one anatomical region has multiple injuries [@trissbetter].
+A limitation of the ISS is that it does not account for situations where one anatomical region has multiple injuries [@trissbetter].
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
@@
 M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths<!-- I would not include preventable deaths here-->.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care.
 The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
@@
 However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
-A study at this institution<!-- You haven't mentioned "this institution" before, so it's hard to understand what you refer to.
-I suggest that state the actual centre--> found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
+A study at Karolinska University Hospital found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
 However, that study included only deaths and not other morbidities and thereby had a low sample size.
@@
 
-This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH)<!-- Use the full hospital name-->: the trauma registry and the trauma care quality database.
+This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital: the trauma registry and the trauma care quality database.
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
@@
 KUH is located in Solna, Sweden.
-KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
-KUH is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma.
-KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
+Karolinska University Hospital's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
+Karolinska University Hospital is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma.
+Karolinska University Hospital is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
 
-The trauma registry at KUH is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
+The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
 The inclusion criteria are given below.
@@
 Data is collected from the patient's electronic medical records.
-The database<!-- Trauma registry?--> contains data on patients from 2012 to 2021.
+The trauma registry contains data on patients from 2012 to 2021.
 
@@
 methods of selection of participants. -->
-Participants were patients registered in both the trauma registry and the trauma care quality database at KUH.
+Participants were patients registered in both the trauma registry and the trauma care quality database at Karolinska University Hospital.
 The eligibility criteria for this study were the criteria for the two databases used in addition to some study-specific inclusion criteria.
@@
 The trauma care quality database contains patients from the trauma registry that were selected for M&M review.
-At KUH, all fatalities are automatically selected for M&M review.
+At Karolinska University Hospital, all fatalities are automatically selected for M&M review.
 In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care.
@@
 ### Study-specific exclusion criteria
-Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity<!-- And are not part of the same review process-->.
+Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity and is not part of the same review process.
 Patients that were dead on arrival were also excluded.
@@
 ### Outcome
-The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH.
+The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital.
 OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
@@
 Patient medical records will not be accessed.
-Patients were not asked to give consent for inclusion in the KUH trauma registry.
+Patients were not asked to give consent for inclusion in the Karolinska University Hospital trauma registry.
 The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations.

### 2022-10-25 functions/make_ps12.R
added ps12

@@ -1,0 +1,77 @@
+## https://www.tarn.ac.uk/Content.aspx?c=1895
+## TARN PS12 seems to have basically 5 groupings of coefficients:
+## GCS (or intubated)
+## Age
+## Gender
+## ISS
+## Female + age
+## So this function first works out the value for each of these groups
+## And then combines them
+## b0, b7 and b8 seem to be indepedent constants? The page explicitly says b0 is a constant but doesn't elaborate on b7/b8
+
+make_ps12 <- function (input.data) {
+  
+  b0 <- 4.9146
+  
+  ## ed.gcs == 99 -> intubated
+  if (input.data["ed.gcs"] == 99) {gcs.val <- -2.31186} ## b6
+    else if (input.data["gcs"] >= 13) {gcs.val <- 0} ## b1
+    else if (input.data["gcs"] >= 9) {gcs.val <- -1.27734} ## b2
+    else if (input.data["gcs"] >= 5) {gcs.val <- -1.68936} ## b3
+    else if (input.data["gcs"] >= 4) {gcs.val <- -2.52661} # b4
+    else {gcs.val <- -3.62339} #b5
+
+  ## coefficient for gender
+  if (input.data["gender"] == 1) {gender.val <- 0} # b9
+    else {gender.val <- -0.024416} #b10
+  
+  ## coefficient for age; ignored coefficients for ages <15
+  if (input.data["age"] > 74) {age.val <- -3.01315} ## b18
+    else if (input.data["age"] > 64) {age.val <- -1.74081} ## b17
+    else if (input.data["age"] > 54) {age.val <- -0.995816} ## b16
+    else if (input.data["age"] > 44) {age.val <- -0.557926} ## b15
+    else if (input.data["age"] > 15) {age.val <- 0} ## b14
+    else {age.val <- 0.210467} ## b13
+    
+  if (input.data["gender"] == 2) {
+    if (input.data["age"] > 74) {gen.age.val <- 0.299887} ## b26
+      else if (input.data["age"] > 64) {gen.age.val <- 0.081554} ## b25
+      else if (input.data["age"] > 54) {gen.age.val <- -0.085054} ## b24
+      else if (input.data["age"] > 44) {gen.age.val <- 0.003235} ## b23
+      else if (input.data["age"] > 15) {gen.age.val <- 0} ## b22
[…]

### 2022-10-25 functions/prep_data.R

@@ -87,8 +87,8 @@ prep_data <- function(df, numbers = FALSE) {
 
   ## Not 100% sure if really need to remove sbp/rr == 0
   ## sbp/rr == 0 should perhaps be under "DOA" instead?
-  df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa),all_vars(!is.na(.)))
-  df <- df %>% filter(dom.inj != 999 & ed.sbp > 0 & ed.rr > 0 & asa != 999 & ed.gcs != 999)
+  df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa, gender),all_vars(!is.na(.)))
+  df <- df %>% filter(dom.inj != 999 & ed.sbp > 0 & ed.rr > 0 & asa != 999 & ed.gcs != 999 & gender != 999)
   
   ## Either ed.gcs or pre.gcs should have a usable GCS. ed.gcs == 999 and NA was removed earlier 
   v <- c(3:15)

### 2022-10-25 main.R

@@ -40,16 +40,19 @@ missing <- pd[['missing']] ## nr of rows with missing data for each parameter
 df$rts <- apply(df, 1, make_rts)
 df$triss <- apply(df, 1, make_triss)
 df$normit <- apply(df, 1, make_normit)
+df$ps12 <- apply(df, 1, make_ps12)
 
 ## cast to numeric to avoid Problems
 df$ofi <- as.numeric(df$ofi)
 df$triss <- as.numeric(df$triss)
 df$normit <- as.numeric(df$normit)
+df$ps12 <- as.numeric(df$ps12)
 
 ## some subsetted data to make be friendly to stats functions
 ## this means my stats functions can be generic in case i add scores later
 t.df <- df[, c("ofi", "triss")] %>% rename(score = triss)
 n.df <- df[, c("ofi", "normit")] %>% rename(score = normit)
+p.df <- df[, c("ofi", "ps12")] %>% rename(score = ps12)
 
 ## make_stats returns a named list with all the things i want; see stats.R
 t.stats <- make_stats(t.df)
@@ -58,6 +61,9 @@ t.or <- t.or[2,]
 n.stats <- make_stats(n.df)
 n.or <- n.stats[['or']]
 n.or <- n.or[2,]
+p.stats <- make_stats(p.df)
+p.or <- p.stats[['or']]
+p.or <- p.or[2,]
 
 ## descriptive data
 table.one <- make_table_one(df)

### 2022-10-25 functions/make_ps12.R
fix broken ps12 calc

@@ -1,49 +1,49 @@
 ## https://www.tarn.ac.uk/Content.aspx?c=1895
-## TARN PS12 seems to have basically 5 groupings of coefficients:
-## GCS (or intubated)
-## Age
-## Gender
-## ISS
-## Female + age
-## So this function first works out the value for each of these groups
-## And then combines them
-## b0, b7 and b8 seem to be indepedent constants? The page explicitly says b0 is a constant but doesn't elaborate on b7/b8
 
 make_ps12 <- function (input.data) {
   
-  b0 <- 4.9146
+  ed.gcs <- as.numeric(input.data["ed.gcs"])
+  gcs <- as.numeric(input.data["gcs"] )
+  age <- as.numeric(input.data["age"])
+  gender <- as.numeric(input.data["gender"])
+  iss <- as.numeric(input.data["iss"])
   
-  ## ed.gcs == 99 -> intubated
-  if (input.data["ed.gcs"] == 99) {gcs.val <- -2.31186} ## b6
-    else if (input.data["gcs"] >= 13) {gcs.val <- 0} ## b1
-    else if (input.data["gcs"] >= 9) {gcs.val <- -1.27734} ## b2
-    else if (input.data["gcs"] >= 5) {gcs.val <- -1.68936} ## b3
-    else if (input.data["gcs"] >= 4) {gcs.val <- -2.52661} # b4
-    else {gcs.val <- -3.62339} #b5
+  b0 <- 4.9146
 
+  ## ed.gcs == 99 -> intubated
+  if (ed.gcs == 99) {gcs.val <- -2.31186} ## b6
+    else { 
+    if (gcs >= 13) {gcs.val <- 0} ## b1
+    else if (gcs >= 9) {gcs.val <- -1.27734} ## b2
+    else if (gcs >= 5) {gcs.val <- -1.68936} ## b3
+    else if (gcs >= 4) {gcs.val <- -2.52661} # b4
+    else {gcs.val <- -3.62339} #b5
+    }
+  
+  
   ## coefficient for gender
-  if (input.data["gender"] == 1) {gender.val <- 0} # b9
+  if (gender == 1) {gender.val <- 0} # b9
     else {gender.val <- -0.024416} #b10
   
+  
   ## coefficient for age; ignored coefficients for ages <15
-  if (input.data["age"] > 74) {age.val <- -3.01315} ## b18
-    else if (input.data["age"] > 64) {age.val <- -1.74081} ## b17
[…]

### 2022-10-25 functions/prep_data.R

@@ -72,7 +72,6 @@ prep_data <- function(df, numbers = FALSE) {
   ## Exclusion: parameters
   
   ## Jonatan wanted me to make a table of exactly how many were missing for each parameter
-  ## Age is higher up, above the <15 exclusion
   m.gcs <- sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),]) ## This does not take into account patients removed where GCS == 99 & there's no prehospital GCS, fix later
   m.asa <- sum(is.na(df$asa)) + nrow(df[df$asa == 999 & !is.na(df$asa),])
   m.rr <- sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0 & !is.na(df$ed.rr),])
@@ -81,8 +80,9 @@ prep_data <- function(df, numbers = FALSE) {
   m.age <- sum(is.na(df$age))
   m.iss <- sum(is.na(df$iss))
   m.niss <- sum(is.na(df$niss))
-  m.names <- c("GCS", "ASA", "RR", "SBP", "Dominant injury type", "Age", "ISS", "NISS")
-  m.numbers <- c(m.gcs, m.asa, m.rr, m.sbp, m.dominj, m.age, m.iss, m.niss)
+  m.gender <- sum(is.na(df$gender)) + nrow(df[df$gender == 999 & !is.na(df$gender),])
+  m.names <- c("GCS", "ASA", "RR", "SBP", "Dominant injury type", "Age", "ISS", "NISS", "Gender")
+  m.numbers <- c(m.gcs, m.asa, m.rr, m.sbp, m.dominj, m.age, m.iss, m.niss, m.gender)
   missing <- tibble("Required parameter" = m.names, "Total no. cases without required parameter" = m.numbers)
 
   ## Not 100% sure if really need to remove sbp/rr == 0

### 2022-10-25 main.R

@@ -65,6 +65,9 @@ p.stats <- make_stats(p.df)
 p.or <- p.stats[['or']]
 p.or <- p.or[2,]
 
+
+test.ps$ps12 <- apply(test.ps, 1, make_ps12)
+
 ## descriptive data
 table.one <- make_table_one(df)
 numbers <- table_one_stats(df) ## some manually generated descriptive statistics

### 2022-10-26 manuscript.Rmd
added text about UN sdg

@@
 ============
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@globalburden].
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting around 9% of global deaths annually [@globalburden].
 The global health burden of trauma has led to it being described as a worldwide pandemic.
@@
 Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 <!-- [@spg] -->.
+This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 <!-- [@spg2020] -->.
 Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.
-
-<!-- S wants me to write about the UN 2030 goals here -->
 
@@
 ```{r, echo = FALSE}
-plot(t.stats[['perf']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(t.stats[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r, echo = FALSE}
-plot(n.stats[['perf']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(n.stats[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 <!-- 
-
 Key results
@@
 
+-->
+
 Limitations
@@
 
+<!-- 
 Discuss limitations of the study, taking into account sources of
 potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias
+of any potential bias -->
 
+<!-- 
 Interpretation

### 2022-10-26 functions/auc_diff.R
added AUC diff + ci

@@ -1,0 +1,13 @@
+
+
+auc_diff <- function(roc1, roc2) {
+  
+  roctest <- roc.test(roc1, roc2, method="delong")
+  
+  diff <- roctest[['estimate']][[1]] - roctest[['estimate']][[2]]
+  ci.low <- roctest[['conf.int']][[1]]
+  ci.hi <- roctest[['conf.int']][[2]]
+  
+  ci.diff <- c(diff, ci.low, ci.hi)
+  return(ci.diff)
+}

### 2022-10-26 functions/stats.R

@@ -5,13 +5,11 @@ make_stats <-function(df) {
   ## Sets up prediction and performance objects
   pred.model<- predict(model, type="response")
   pred <- prediction(pred.model, df$ofi)
-
-  perf <- performance(pred, "tpr", "fpr") ## ROC
-  auc <- unlist(slot(performance(pred, "auc"), "y.values")) ## AUC
-  auc <- round(auc, digits = 2)
-  proc <- roc(ofi ~ score, data = df) ## This is so I can use the pROC package to calculate AUC CI
-  auc.ci <- round(ci.auc(proc), digits = 2)
   
+  roc <- roc(ofi ~ score, data = df) ## pROC
+  auc <- auc(roc)
+  auc.ci <- round(ci.auc(roc), digits = 2)
+
   ## The console output of this says that it uses "mgcv::gam instead of loess to calculate ICI"
   ici <- ici(pred.model, df$ofi)
   ici <- round(ici, digits = 2)
@@ -34,7 +32,7 @@ make_stats <-function(df) {
   
   ## Summary of the things I should now have:
   ## model = the logistic regression model
-  ## perf = the ROC
+  ## roc = the ROC
   ## auc = the AUC for the ROC
   ## auc.ci = the CI for the AUC
   ## acc = a plottable accuracy/cutoff
@@ -44,6 +42,6 @@ make_stats <-function(df) {
   ## acc.nrs = the accuracy and cutoff at the highest accuracy
   
   ## return all the stats that we've created as a single list; this seemed more efficient than having multiple different functions that do similar things
-  stats.ret <- list(model = model, perf = perf, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci, or = or, or.p = or.p, acc.nrs = acc.nrs)
+  stats.ret <- list(model = model, roc = roc, auc = auc, acc = acc, ici = ici, auc.ci = auc.ci, or = or, or.p = or.p, acc.nrs = acc.nrs)
   return(stats.ret)
 }

### 2022-10-26 main.R

@@ -65,15 +65,19 @@ p.stats <- make_stats(p.df)
 p.or <- p.stats[['or']]
 p.or <- p.or[2,]
 
-
-test.ps$ps12 <- apply(test.ps, 1, make_ps12)
+## differences between AUC
+tn.aucdiff <- auc_diff(t.stats[['roc']], n.stats[['roc']]) ## TRISS AUC - NORMIT AUC
+tp.aucdiff <- auc_diff(t.stats[['roc']], p.stats[['roc']]) ## TRISS AUC - PS12 AUC
+np.aucdiff <- auc_diff(n.stats[['roc']], p.stats[['roc']]) ## NORMIT AUC - PS12 AUC
 
 ## descriptive data
 table.one <- make_table_one(df)
 numbers <- table_one_stats(df) ## some manually generated descriptive statistics
 
 ## plots to be used in manuscript
-t.plot.roc <- plot(t.stats[['perf']], main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
+t.plot.roc <- plot(t.stats[['roc']], main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
 t.plot.acc <- plot(t.stats[['acc']], main="TRISS Accuracy")
-n.plot.roc <- plot(n.stats[['perf']], main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
-n.plot.acc <- plot(n.stats[['acc']], main="TRISS Accuracy")
+n.plot.roc <- plot(n.stats[['roc']], main="NORMIT Receiver Operating Characteristic Curve", colorize = TRUE)
+n.plot.acc <- plot(n.stats[['acc']], main="NORMIT Accuracy")
+p.plot.roc <- plot(n.stats[['roc']], main="TARN Ps12 Receiver Operating Characteristic Curve", colorize = TRUE)
+p.plot.acc <- plot(n.stats[['acc']], main="TARN Ps12 Accuracy")

### 2022-11-02 functions/prep_data.R
fix doa

@@ -34,7 +34,7 @@ prep_data <- function(df, numbers = FALSE) {
                       ed.rr = ed_rr_value,
                       ed.sbp = ed_sbp_value,
                       pre.gcs = pre_gcs_sum,
-                      age = pt_age_yrs,
+                      age = pt_age_yrsg
                       )
   
   ## Storing the inclusion/exclusion counts at each step so I can use those later
@@ -55,7 +55,7 @@ prep_data <- function(df, numbers = FALSE) {
   
   ## Exclusion: age
   
-  df <- df %>% filter (age >= 15)
+  ## df <- df %>% filter (age >= 15)
   
   age.kept <- nrow(df)
   age.excluded <- ofi.kept - age.kept
@@ -64,6 +64,7 @@ prep_data <- function(df, numbers = FALSE) {
   ## Exclusion: doa
   
   df <- df %>% filter(is.na(doa)|doa != 1)
+  df <- df %>% filter (ed.sbp != 0 & ed.rr != 0 & ed.gcs != 3)
   
   doa.kept <- nrow(df)
   doa.excluded <- age.kept - doa.kept
@@ -85,10 +86,8 @@ prep_data <- function(df, numbers = FALSE) {
   m.numbers <- c(m.gcs, m.asa, m.rr, m.sbp, m.dominj, m.age, m.iss, m.niss, m.gender)
   missing <- tibble("Required parameter" = m.names, "Total no. cases without required parameter" = m.numbers)
 
-  ## Not 100% sure if really need to remove sbp/rr == 0
-  ## sbp/rr == 0 should perhaps be under "DOA" instead?
   df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa, gender),all_vars(!is.na(.)))
-  df <- df %>% filter(dom.inj != 999 & ed.sbp > 0 & ed.rr > 0 & asa != 999 & ed.gcs != 999 & gender != 999)
+  df <- df %>% filter(dom.inj != 999 & asa != 999 & ed.gcs != 999 & gender != 999)
   
   ## Either ed.gcs or pre.gcs should have a usable GCS. ed.gcs == 999 and NA was removed earlier 
   v <- c(3:15)

### 2022-11-02 main.R

@@ -74,10 +74,5 @@ np.aucdiff <- auc_diff(n.stats[['roc']], p.stats[['roc']]) ## NORMIT AUC - PS12 
 table.one <- make_table_one(df)
 numbers <- table_one_stats(df) ## some manually generated descriptive statistics
 
-## plots to be used in manuscript
-t.plot.roc <- plot(t.stats[['roc']], main="TRISS Receiver Operating Characteristic Curve", colorize = TRUE)
-t.plot.acc <- plot(t.stats[['acc']], main="TRISS Accuracy")
-n.plot.roc <- plot(n.stats[['roc']], main="NORMIT Receiver Operating Characteristic Curve", colorize = TRUE)
-n.plot.acc <- plot(n.stats[['acc']], main="NORMIT Accuracy")
-p.plot.roc <- plot(n.stats[['roc']], main="TARN Ps12 Receiver Operating Characteristic Curve", colorize = TRUE)
-p.plot.acc <- plot(n.stats[['acc']], main="TARN Ps12 Accuracy")
+test <- merge_data(datasets)
+test <- test %>% filter (age >= 15)

### 2022-11-02 functions/prep_data.R
fix issues w/ NA being removed at wrong time + fix missing GCS counting

@@ -34,7 +34,7 @@ prep_data <- function(df, numbers = FALSE) {
                       ed.rr = ed_rr_value,
                       ed.sbp = ed_sbp_value,
                       pre.gcs = pre_gcs_sum,
-                      age = pt_age_yrsg
+                      age = pt_age_yrs
                       )
   
   ## Storing the inclusion/exclusion counts at each step so I can use those later
@@ -45,7 +45,7 @@ prep_data <- function(df, numbers = FALSE) {
                                   )
   ## Exclusion: ofi
 
-  df <- df %>% filter (ofi != "NA")
+  df <- df %>% filter (!is.na(ofi))
   
   ## So what I'm doing is:
   ## After every exclusion counting the new number of rows, then comparing it with the previous number of rows to see how many were excluded
@@ -55,7 +55,7 @@ prep_data <- function(df, numbers = FALSE) {
   
   ## Exclusion: age
   
-  ## df <- df %>% filter (age >= 15)
+  df <- df %>% filter (age >= 15 | is.na(df$asa))
   
   age.kept <- nrow(df)
   age.excluded <- ofi.kept - age.kept
@@ -83,15 +83,17 @@ prep_data <- function(df, numbers = FALSE) {
   m.niss <- sum(is.na(df$niss))
   m.gender <- sum(is.na(df$gender)) + nrow(df[df$gender == 999 & !is.na(df$gender),])
   m.names <- c("GCS", "ASA", "RR", "SBP", "Dominant injury type", "Age", "ISS", "NISS", "Gender")
-  m.numbers <- c(m.gcs, m.asa, m.rr, m.sbp, m.dominj, m.age, m.iss, m.niss, m.gender)
-  missing <- tibble("Required parameter" = m.names, "Total no. cases without required parameter" = m.numbers)
 
   df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa, gender),all_vars(!is.na(.)))
   df <- df %>% filter(dom.inj != 999 & asa != 999 & ed.gcs != 999 & gender != 999)
   
   ## Either ed.gcs or pre.gcs should have a usable GCS. ed.gcs == 999 and NA was removed earlier 
+  nrow1 <- nrow(df)
   v <- c(3:15)
   df <- df %>% filter(pre.gcs %in% v | ed.gcs %in% v)
+  nrow2 <- nrow(df)
[…]

### 2022-11-02 main.R

@@ -73,6 +73,3 @@ np.aucdiff <- auc_diff(n.stats[['roc']], p.stats[['roc']]) ## NORMIT AUC - PS12 
 ## descriptive data
 table.one <- make_table_one(df)
 numbers <- table_one_stats(df) ## some manually generated descriptive statistics
-
-test <- merge_data(datasets)
-test <- test %>% filter (age >= 15)

### 2022-11-02 functions/prep_data.R
prep_data should be completed now

@@ -73,7 +73,7 @@ prep_data <- function(df, numbers = FALSE) {
   ## Exclusion: parameters
   
   ## Jonatan wanted me to make a table of exactly how many were missing for each parameter
-  m.gcs <- sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),]) ## This does not take into account patients removed where GCS == 99 & there's no prehospital GCS, fix later
+  m.gcs <- sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),])
   m.asa <- sum(is.na(df$asa)) + nrow(df[df$asa == 999 & !is.na(df$asa),])
   m.rr <- sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0 & !is.na(df$ed.rr),])
   m.sbp <- sum(is.na(df$ed.sbp)) + nrow(df[df$ed.sbp == 0 & !is.na(df$ed.sbp),])
@@ -92,7 +92,7 @@ prep_data <- function(df, numbers = FALSE) {
   v <- c(3:15)
   df <- df %>% filter(pre.gcs %in% v | ed.gcs %in% v)
   nrow2 <- nrow(df)
-  m.pregcs <- nrow1 - nrow2
+  m.pregcs <- nrow1 - nrow2 ## adding missing pregcs when edgcs == 99 to missing parameter count
   m.gcs <- m.gcs + m.pregcs
   
   ## Make a new gcs variable that takes into account ed.gcs = 99 

### 2022-11-02 functions/make_ps12.R
minor fixes

@@ -13,13 +13,12 @@ make_ps12 <- function (input.data) {
   ## ed.gcs == 99 -> intubated
   if (ed.gcs == 99) {gcs.val <- -2.31186} ## b6
     else { 
-    if (gcs >= 13) {gcs.val <- 0} ## b1
-    else if (gcs >= 9) {gcs.val <- -1.27734} ## b2
-    else if (gcs >= 5) {gcs.val <- -1.68936} ## b3
-    else if (gcs >= 4) {gcs.val <- -2.52661} # b4
+    if (ed.gcs >= 13) {gcs.val <- 0} ## b1
+    else if (ed.gcs >= 9) {gcs.val <- -1.27734} ## b2
+    else if (ed.gcs >= 5) {gcs.val <- -1.68936} ## b3
+    else if (ed.gcs >= 4) {gcs.val <- -2.52661} # b4
     else {gcs.val <- -3.62339} #b5
     }
-  
   
   ## coefficient for gender
   if (gender == 1) {gender.val <- 0} # b9

### 2022-11-02 functions/stats.R

@@ -24,9 +24,6 @@ make_stats <-function(df) {
   acc.nrs <- c(acc.max = acc.max, co = cutoff) ## Make a vector to return
   
   ## This makes a vector with the OR and the upper/lower bounds of the CI
-  ## confint.default is confusing because when I do level=0.95, the output says "97.5%" and "2.5%"
-  ## So I changed it to 0.90 because then it gives 95%/5%
-  ## Really not sure if this is the correct function to use or if I'm using it correctly
   or <- round(exp(cbind("or" = coef(model), confint.default(model, level = 0.90))), digits = 2)
   or.p <- round(summary(model)$coefficients[2,4], digits = 2) ## Taking the p-value from glm()
   

### 2022-11-02 manuscript.Rmd
fix UN cites

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 ```
+
 ```{r main, echo=FALSE, message = FALSE}
@@
 * OFI - Opportunities for improvement
+* PS - Probability of Survival score
 * QI - Quality Improvement
@@
 Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
-The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 <!-- [@spg] -->.
-This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 <!-- [@spg2020] -->.
+The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg].
+This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 [@sdg20].
 Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.
@@
 The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
-The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
+The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating probability of survival in certain populations [@trissbetter].
 
@@
 
-KUH is located in Solna, Sweden.
+Karolinska University Hospital is located in Solna, Sweden.
 Karolinska University Hospital's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
@@
 
-``` {r, echo = FALSE,} 
+``` {r} 
   ofi_diagram() ## Only works in HTML
@@
 
-```{r, echo = FALSE}
+```{r}
   exclusion_flowchart(counts) ## Only works in HTML
@@
 
-```{r, echo = FALSE} 
+```{r} 
   table.one ## will be improved later
@@
 
-```{r, echo = FALSE}
+```{r}
 plot(t.stats[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
@@
 
-```{r, echo = FALSE}
+```{r}
 t.plot.acc <- plot(t.stats[['acc']], main="Fig 4. TRISS Accuracy Curve.")

### 2022-11-02 functions/make_triss.R
changed triss coefficients according to my understanding of the 2009 paper

@@ -1,29 +1,63 @@
+## Source: https://journals.lww.com/jtrauma/Fulltext/2010/04000/Trauma_and_Injury_Severity_Score__TRISS_.2.aspx
+## Using the "imputed NTDB" coefficients as they were recommended
+## I can't see a coefficient that is directly for RTS so I substituted the formula for RTS into the TRISS formula
+
 make_triss <- function(input.data) {
   
-  ## Saving the coefficients for later
-  lookup <- data.frame(
-    inj.type = c(1, 2), ## 1 = blunt, 2 = penetrating
-    b0 = c(-0.4499, -2.5355),
-    b1 = c(0.8085, 0.9934),
-    b2 = c(-0.0835, -0.651),
-    b3 = c(-1.743, -1.136)
-  )
-  
   ## Cast variables to numeric to ensure they can be used for calculations
-  t.dom.inj <- as.numeric(input.data["dom.inj"])
-  t.rts <- as.numeric(input.data["rts"])
-  t.iss <- as.numeric(input.data["iss"])
-  t.age <- as.numeric(input.data["age"])
+  dom.inj <- as.numeric(input.data["dom.inj"])
+  iss <- as.numeric(input.data["iss"])
+  age <- as.numeric(input.data["age"])
+  gcs <- as.numeric(input.data["gcs"])
+  sbp <- as.numeric(input.data["ed.sbp"])
+  rr <- as.numeric(input.data["ed.rr"])
+
+  if (dom.inj == 1) { ## 1 = blunt
+    b0 <- 1.6494 ## intercept
+    b1 <- 0.0095 ## RR
+    b2 <- 0.4260 ## SBP
+    b3 <- 0.6307 ## GCS
+    b4 <- -0.0795 ## ISS
+    b5 <- -1.6216 ## Age
+  } else {
+    b0 <- -0.5757
+    b1 <- 0.1517
+    b2 <- 0.5237
+    b3 <- 0.8310
+    b4 <- -0.0872
+    b5 <- -0.8714
+  }
+  
+  b1 <- as.numeric(b1)
+  b2 <- as.numeric(b2)
+  b3 <- as.numeric(b3)
+  b4 <- as.numeric(b4)
+  b5 <- as.numeric(b5)
   
   ## Assign the coefficient for age
-  if (t.age > 54) {age.val <- 1} else {age.val <- 0}
-
[…]

### 2022-11-02 main.R

@@ -50,25 +50,25 @@ df$ps12 <- as.numeric(df$ps12)
 
 ## some subsetted data to make be friendly to stats functions
 ## this means my stats functions can be generic in case i add scores later
-t.df <- df[, c("ofi", "triss")] %>% rename(score = triss)
-n.df <- df[, c("ofi", "normit")] %>% rename(score = normit)
-p.df <- df[, c("ofi", "ps12")] %>% rename(score = ps12)
+df.t <- df[, c("ofi", "triss")] %>% rename(score = triss)
+df.n <- df[, c("ofi", "normit")] %>% rename(score = normit)
+df.p <- df[, c("ofi", "ps12")] %>% rename(score = ps12)
 
 ## make_stats returns a named list with all the things i want; see stats.R
-t.stats <- make_stats(t.df)
-t.or <- t.stats[['or']]
-t.or <- t.or[2,]
-n.stats <- make_stats(n.df)
-n.or <- n.stats[['or']]
-n.or <- n.or[2,]
-p.stats <- make_stats(p.df)
-p.or <- p.stats[['or']]
-p.or <- p.or[2,]
+stats.t <- make_stats(df.t)
+or.t <- stats.t[['or']]
+or.t <- or.t[2,]
+stats.n <- make_stats(df.n)
+or.n <- stats.n[['or']]
+or.n <- or.n[2,]
+stats.p <- make_stats(df.p)
+or.p <- stats.p[['or']]
+or.p <- or.p[2,]
 
 ## differences between AUC
-tn.aucdiff <- auc_diff(t.stats[['roc']], n.stats[['roc']]) ## TRISS AUC - NORMIT AUC
-tp.aucdiff <- auc_diff(t.stats[['roc']], p.stats[['roc']]) ## TRISS AUC - PS12 AUC
-np.aucdiff <- auc_diff(n.stats[['roc']], p.stats[['roc']]) ## NORMIT AUC - PS12 AUC
+aucdiff.tn <- auc_delta(stats.t[['roc']], stats.n[['roc']]) ## TRISS AUC - NORMIT AUC
+aucdiff.tp <- auc_delta(stats.t[['roc']], stats.p[['roc']]) ## TRISS AUC - PS12 AUC
+aucdiff.np <- auc_delta(stats.n[['roc']], stats.p[['roc']]) ## NORMIT AUC - PS12 AUC
 
 ## descriptive data
 table.one <- make_table_one(df)

### 2022-11-02 manuscript.Rmd
updating method

@@
 Trauma severity models have become a cornerstone of modern trauma QI programs [@historytss].
-One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines].
+One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high predicted probability of survival.
+[@whoguidelines].
 Trauma severity models also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
@@
 TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
-The resulting value is a percentage that estimates the probability of survival (Ps).
+The resulting value the probability of survival given as a percentage.
 The ISS gives a score based on the three anatomical regions with the most severe injury.
@@
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
-The predictive performance of these models will be evaluated using measures of discrimination and calibration.
+These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
+The predictive performance of these models was evaluated using measures of discrimination and calibration.
 
@@
 Karolinska University Hospital is located in Solna, Sweden.
-Karolinska University Hospital's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats].
-Karolinska University Hospital is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma.
-Karolinska University Hospital is thereby the major trauma centre for a population of almost 3 million people [@nksstats].
+The trauma centre at Karolinska University Hospital is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats] and it is the primary trauma centre for the region of Stockholm.
+Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].
+<!-- Potential things to add about NKS:
+1. How many trauma patients per year?
+2. How often do conferences happen, or alternatively, roughly how many per year?
+3. Which specialties/professions participate in conferences? 
+4. Any other info about conferences at NKS that might be relevant like if there's a particular format or it follows particular guidelines etc 
+5. Also if there's any stats about trauma patients at NKS in general, like the avg ISS or similar
+... although I guess I have the data to calculate this myself-->
 
 The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
-The inclusion criteria are given below.
-All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma centre.
+All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre.
 Data is collected from the patient's electronic medical records.
 The trauma registry contains data on patients from 2012 to 2021.
+<!-- Double check these dates, maybe give exact dates of registration in the study -->
+<!-- Also is there a time limit for when patients are registered in the registry? Like x days/weeks following admission -->
 
@@
 The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment.
-Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS<!-- Do you mean GOS?--> at discharge, highest level of care, and transfers to other units.
+Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, Glasgow Outcome Scale score at discharge, highest level of care, and transfers to other units.
 
-The trauma care quality database is an internal hospital register and registers the outcome of M&M conferences.
+The trauma care quality database is an internal hospital registry and registers the outcome of M&M conferences.
 Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
 The database contains data on patients from 2014 to 2021.
+<!-- Double check these dates -->
+
+<!-- How exactly is OFI identified, is it according to specific criteria? -->
 
@@
 Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15.
-Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.
+Patients where the only traumatic injury is a chronic subdural haematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.
 
@@
 
-### Study-specific exclusion criteria
+### Study-specific eligibility criteria
 Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity and is not part of the same review process.
 Patients that were dead on arrival were also excluded.
-In addition, this is a complete case study and all patients that lacked parameters necessary to carry out the analysis were excluded.
+This is a complete case study and all patients with missing data were excluded.
 
@@
 ### Exposures
-The primary exposures of interest are the trauma severity models TRISS and NORMIT.
-Both of these models estimate the Ps.
-This is expressed a probability and is a continuous variable.
-The variables integrated into these models are taken from arrival at hospital.
+The primary exposures of interest are the trauma severity models TRISS, TARN PS12, and NORMIT.
+All three of these models estimate the probability of death.
+This is expressed a percentage and is a continuous variable.
+The scores given by each of these models were calculated using recorded physiological and anatomical variables.
+Conversion and handling of these variables was carried out according to the SweTrau manual.
 
@@
 
-The TRISS model incorporates patient age, Injury Severity Score (ISS), Revised Trauma Score (RTS), and whether the trauma was blunt or penetrating.
+The TRISS model incorporates patient age, Injury Severity Score (ISS), GCS, systolic blood pressure, and respiratory rate.
+There are two sets of coefficients for these variables according to whether the trauma was blunt or penetrating.
 
-The ISS is given by estimating the severity of physical injuries by anatomical region.
+The ISS is recorded in the trauma registry as part of original data collection.
+It is given by estimating the severity of physical injuries by anatomical region.
 Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury".
 The values from the three anatomical regions with the most severe injury are combined into a score that is given is points on a scale from 0 to 75.
-The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate.
-Values are taken from arrival at hospital.
-These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.
 
@@
 
+The NISS is recorded in the trauma registry as part of original data collection.
 NISS is based on the same variables as ISS.
 It differs from ISS in that it takes into account multiple injuries within one anatomical region and calculates a score based on the three most severe injuries regardless of anatomical region.
+
+The RTS was computed using GCS, systolic blood pressure, and respiratory rate.
+For the systolic blood pressure and respiratory rate, values are taken from arrival at hospital.
+The GCS is also taken from arrival at hospital except in cases where the patient is intubated pre-hospitally; in these cases, the pre-hospital GCS is used.
+These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.
+
+### PS
+
+The PS model incorporates patient age, gender, GCS and ISS.
 
@@
 
-NORMIT and TRISS will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.
-The scores will be calculated according to the formulas given by the respective developers of each model.
+RTS, NORMIT and PS12 were calculated according to the published algorithms from their developers [@rtsalgorithm, @normit, @psalgorithm].
+TRISS was calculated using to the revised 2009 coefficients [@triss2010].
+
+For all parameters other than GCS, the values used to compute scores were solely taken from arrival at hospital.
+For GCS, values were used based on arrival at hospital except in cases where the patient was intubated pre-hospitally.
+PS and NORMIT have methods of handling this scenario, as described above.
+TRISS does not have a published method to handle patients intubated pre-hospitally.
+In order to avoid excluding patients who were intubated pre-hospitally, the GCS parameter for TRISS was handled in the same manner as NORMIT, i.e.
+the pre-hospital GCS value was used.
 
@@
 
-R, a programming language for statistical computing, will be used for compilation and analysis of data.
-Conversion and handling of variables will be carried out according to the SweTrau manual.
-In cases where data necessary for analysis is missing, i.e.
-the variables referred to above, the participant will be excluded from the study.
-Logistic regression will be used to determine association between the estimates of the trauma severity models and OFI.
-To assess how well the models predict OFI measures of discrimination and calibration will be used.
-Results will be presented with a confidence interval of 95%.
-A p-value of <0.05 will be considered significant.
-
-<!-- computed CIs for OR using wald estimator. -->
+R, a programming language for statistical computing, was used for data collection and analysis.
+Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR.
+The CI for the OR was computed using the Wald estimator.
+Receiver operator characteristic (ROC) curves were used to assess diagnostic performance.
+The area under the curve (AUC) was calculated to compare the models with one another.
+The Integrated Calibration Index (ICI) was used to measure calibration.
+Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff.
+Results werepresented with a confidence interval of 95%.
+A p-value of <0.05 was considered significant.
 
@@
 
-The odds ratio for TRISS was `r t.or['or']` (95% CI: `r t.or[2]`-`r t.or[3]`, p-value: `r t.stats[['or.p']]`).
+### TRISS
+
+The odds ratio for TRISS was `r or.t['or']` (95% CI: `r or.t[2]`-`r or.t[3]`, p-value: `r stats.t[['or.p']]`).
 The ROC is plotted in Figure 3.
-The AUC for the ROC was `r t.stats[['auc']]` (95% CI: `r t.stats[['auc.ci']][1]`-`r t.stats[['auc.ci']][3]`).
-The integrated calibration index for TRISS was `r t.stats[['ici']]`.
+The AUC for the ROC was `r stats.t[['auc']]` (95% CI: `r stats.t[['auc.ci']][1]`-`r stats.t[['auc.ci']][3]`).
+The integrated calibration index for TRISS was `r stats.t[['ici']]`.
 The accuracy curve is shown in Figure 4.
-The best accuracy was `r t.stats[['acc.nrs']][1]` and was found at the cutoff `r t.stats[['acc.nrs']][2]`.
+The best accuracy was `r stats.t[['acc.nrs']][1]` and was found at the cutoff `r stats.t[['acc.nrs']][2]`.
 
 ```{r}
-plot(t.stats[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(stats.t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r}
-t.plot.acc <- plot(t.stats[['acc']], main="Fig 4. TRISS Accuracy Curve.")
+t.plot.acc <- plot(stats.t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
 ```
 
-The odds ratio for NORMIT was `r n.or['or']` (95% CI: `r n.or[2]`-`r n.or[3]`, p-value: `r n.stats[['or.p']]`).
+### NORMIT 
+The odds ratio for NORMIT was `r or.n['or']` (95% CI: `r or.n[2]`-`r or.n[3]`, p-value: `r stats.n[['or.p']]`).
 The ROC is plotted in Figure 5.
-The AUC for the ROC was `r n.stats[['auc']]` (95% CI: `r n.stats[['auc.ci']][1]`-`r n.stats[['auc.ci']][3]`).
-The integrated calibration index for NORMIT was `r n.stats[['ici']]`.
+The AUC for the ROC was `r stats.n[['auc']]` (95% CI: `r stats.n[['auc.ci']][1]`-`r stats.n[['auc.ci']][3]`).
+The integrated calibration index for NORMIT was `r stats.n[['ici']]`.
 The accuracy curve for NORMIT is shown in Figure 5.
-The best accuracy was `r n.stats[['acc.nrs']][1]` and was found at the cutoff `r n.stats[['acc.nrs']][2]`.
+The best accuracy was `r stats.n[['acc.nrs']][1]` and was found at the cutoff `r stats.n[['acc.nrs']][2]`.
 
 ```{r, echo = FALSE}
-plot(n.stats[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(stats.n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r, echo = FALSE}
-n.plot.acc <- plot(n.stats[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
+n.plot.acc <- plot(stats.n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
 ```
+
+### PS
+
+### Comparison of scores
 

### 2022-11-02 manuscript.Rmd
fixed method more

@@
 
-<!-- Clearly define all outcomes, exposures, predictors, potential
-confounders, and effect modifiers. Give diagnostic criteria, if
-applicable. For each variable of interest, give sources of data and
-details of methods of assessment (measurement). Describe comparability
-of assessment methods if there is more than one group -->
-
 ### Outcome
@@
 
-The PS model incorporates patient age, gender, GCS and ISS.
-
-Bias 
-----
-
-<!-- Describe any efforts to address potential sources of bias -->
-
-The method and data analysis model will be written using simulated data to reduce the risk of research bias.
-Real-world data will only be used once the data analysis model is completed and found to work correctly on simulated data.
-
-Study size 
-----------
-
-<!-- Explain how the study size was arrived at -->
-
-Since all eligible participants must exist in both the trauma registry and the trauma care quality database, and the trauma care quality database is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the trauma care quality database.
-The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.
-In order to establish the exact study size, the study-specific exclusion criteria need to first be applied.
-The exact study size has not yet been established at the time of writing this report.
+The algorithm for PS includes the patient's age, gender, GCS and ISS.
 
@@
 
-RTS, NORMIT and PS12 were calculated according to the published algorithms from their developers [@rtsalgorithm, @normit, @psalgorithm].
-TRISS was calculated using to the revised 2009 coefficients [@triss2010].
+RTS, NORMIT and PS12 were calculated according to the original published algorithms [@rtsalgorithm, @normit, @psalgorithm]. TRISS was calculated using the revised 2009 coefficients [@triss2010].
 
@@
 the pre-hospital GCS value was used.
+
+Bias 
+----
+
+The code for data analysis was written using simulated data to reduce the risk of research bias.
+Real-world data was only used once the data analysis model was completed and found to work correctly on simulated data.
 

### 2022-11-02 manuscript.Rmd
fix

@@
 The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital.
-OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
+OFI is registered in the trauma care quality database and is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
 
 ### Exposures
-The primary exposures of interest are the trauma severity models TRISS, TARN PS12, and NORMIT.
-All three of these models estimate the probability of death.
-This is expressed a percentage and is a continuous variable.
-The scores given by each of these models were calculated using recorded physiological and anatomical variables.
+The exposures of interest are the trauma severity models TRISS, NORMIT, and PS.
+All three of these models estimate the probability of death, expressed as a probability.
+The scores given by each of these models were calculated using physiological and anatomical parameters recorded in the trauma registry.
 Conversion and handling of these variables was carried out according to the SweTrau manual.

### 2022-11-02 functions/prep_data.R
fix mistake in identifying doa

@@ -39,7 +39,7 @@ prep_data <- function(df, numbers = FALSE) {
   
   ## Storing the inclusion/exclusion counts at each step so I can use those later
   original.count <- nrow(df)
-  inclusion.counts <- data.frame (step  = c("original"),
+  eligibility <- data.frame (step  = c("original"),
                     included = c(original.count),
                     excluded = c(0)
                                   )
@@ -51,7 +51,7 @@ prep_data <- function(df, numbers = FALSE) {
   ## After every exclusion counting the new number of rows, then comparing it with the previous number of rows to see how many were excluded
   ofi.kept <- nrow(df)
   ofi.excluded <- original.count - ofi.kept
-  inclusion.counts[nrow(inclusion.counts) + 1,] = c("ofi", ofi.kept, ofi.excluded)
+  eligibility[nrow(eligibility) + 1,] = c("ofi", ofi.kept, ofi.excluded)
   
   ## Exclusion: age
   
@@ -59,16 +59,16 @@ prep_data <- function(df, numbers = FALSE) {
   
   age.kept <- nrow(df)
   age.excluded <- ofi.kept - age.kept
-  inclusion.counts[nrow(inclusion.counts) + 1,] = c("age", age.kept, age.excluded)
+  eligibility[nrow(eligibility) + 1,] = c("age", age.kept, age.excluded)
   
   ## Exclusion: doa
   
   df <- df %>% filter(is.na(doa)|doa != 1)
-  df <- df %>% filter (ed.sbp != 0 & ed.rr != 0 & ed.gcs != 3)
+  df <- df %>% filter (ed.sbp != 0 | ed.rr != 0 | ed.gcs != 3 | is.na(ed.sbp) | is.na(ed.rr) | is.na(ed.gcs))
   
   doa.kept <- nrow(df)
   doa.excluded <- age.kept - doa.kept
-  inclusion.counts[nrow(inclusion.counts) + 1,] = c("doa", doa.kept, doa.excluded)
+  eligibility[nrow(eligibility) + 1,] = c("doa", doa.kept, doa.excluded)
 
   ## Exclusion: parameters
   
@@ -88,11 +88,13 @@ prep_data <- function(df, numbers = FALSE) {
   df <- df %>% filter(dom.inj != 999 & asa != 999 & ed.gcs != 999 & gender != 999)
   
[…]

### 2022-11-02 manuscript.Rmd
added missing parameters table

@@
 ------------
-A total of `r counts[6,3]` patients were excluded during processing of data.
-`r counts[2,3]` patients were excluded due to not having a recorded outcome from an M&M conference.
-`r counts[3,3]` patients were excluded for being under 15 years of age.
-`r counts[4,3]` patients were excluded due to being dead on arrival at hospital.
-`r counts[5,3]` patients were excluded due to missing required data.
+A total of `r eligibility[6,3]` patients were excluded during processing of data.
+`r eligibility[2,3]` patients were excluded due to not having a recorded outcome from an M&M conference.
+`r eligibility[3,3]` patients were excluded for being under 15 years of age.
+`r eligibility[4,3]` patients were excluded due to being dead on arrival at hospital.
+`r eligibility[5,3]` patients were excluded due to incomplete data.
 Figure 2 shows the selection process for eligible participants.
@@
 ```{r}
-  exclusion_flowchart(counts) ## Only works in HTML
+  exclusion_flowchart(eligibility) ## Only works in HTML
+```
+
+Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table 1.
+Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.
+
+```{r}
+  as_raw_html(incomplete.data.tbl)
 ```
@@
 ----------------
-A total of `r counts[5,2]` participants were identified as eligible for inclusion.
+A total of `r eligibility[5,2]` participants were identified as eligible for inclusion.
 Their demographical and clinical characteristics are described in Table 1.
@@
   table.one ## will be improved later
-```
-
-Detailed totals for missing data for each required parameter are given in table 2.
-The total number of missing parameters is greater than the number of participants excluded for missing parameters.
-This is because some patients lacked multiple required parameters.
-
-```{r gt table, echo = FALSE}
-missing
 ```

### 2022-11-02 functions/auc_diff.R

@@ -1,8 +1,6 @@
-
-
-auc_diff <- function(roc1, roc2) {
+auc_delta <- function(roc1, roc2) {
   
-  roctest <- roc.test(roc1, roc2, method="delong")
+  roctest <- roc.test(roc1, roc2, method="delong") ## has to be forced or it will try to bootstrap; bootstrap doesnt give a CI
   
   diff <- roctest[['estimate']][[1]] - roctest[['estimate']][[2]]
   ci.low <- roctest[['conf.int']][[1]]

### 2022-11-02 main.R

@@ -32,9 +32,13 @@ df <- merge_data(datasets)
 pd <- prep_data(df) ## returns a list with the prepped dataset, inclusion counts, and missing parameters
 df <- pd[['df']]
 
-## exclusion data
-counts <- pd[['counts']] ## inclusion/exclusion counts
-missing <- pd[['missing']] ## nr of rows with missing data for each parameter
+## eligibility counts
+eligibility <- pd[['eligibility']] ## inclusion/exclusion counts
+
+## missing data
+incomplete.data.tbl <- gt(pd[['incomplete.data']]) ## table nr of rows with missing data for each parameter
+incomplete.data.tbl <- incomplete.data.tbl %>%
+  tab_header(title = "Table 1. Total no. missing data for each variable in cases excluded for incomplete data")
 
 ## add scores
 df$rts <- apply(df, 1, make_rts)
@@ -73,3 +77,5 @@ aucdiff.np <- auc_delta(stats.n[['roc']], stats.p[['roc']]) ## NORMIT AUC - PS12
 ## descriptive data
 table.one <- make_table_one(df)
 numbers <- table_one_stats(df) ## some manually generated descriptive statistics
+
+

### 2022-11-02 manuscript.Rmd
text in results -> participants

@@
 ------------
-A total of `r eligibility[6,3]` patients were excluded during processing of data.
-`r eligibility[2,3]` patients were excluded due to not having a recorded outcome from an M&M conference.
-`r eligibility[3,3]` patients were excluded for being under 15 years of age.
-`r eligibility[4,3]` patients were excluded due to being dead on arrival at hospital.
-`r eligibility[5,3]` patients were excluded due to incomplete data.
-Figure 2 shows the selection process for eligible participants.
+A total of `r eligibility[1,2]` patients were recorded in the trauma registry during the study period.
+Of these, `r eligibility[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion.
+`r post.ofi.exclusion` patients were excluded, due to either being of paediatric age (`r eligibility[3,3]`), being dead on arrival at hospital (`r eligibility[4,3]`), or due to incomplete data (`r eligibility[5,3]`).
+Figure 2 shows the inclusion process for eligible participants.
 

### 2022-11-02 main.R

@@ -34,6 +34,7 @@ df <- pd[['df']]
 
 ## eligibility counts
 eligibility <- pd[['eligibility']] ## inclusion/exclusion counts
+post.ofi.exclusion <- as.numeric(eligibility[3,3]) + as.numeric(eligibility[4,3]) + as.numeric(eligibility[5,3])
 
 ## missing data
 incomplete.data.tbl <- gt(pd[['incomplete.data']]) ## table nr of rows with missing data for each parameter

### 2022-11-02 manuscript.Rmd
more/better(?) descriptive data

@@
 ----------------
-A total of `r eligibility[5,2]` participants were identified as eligible for inclusion.
-Their demographical and clinical characteristics are described in Table 1.
-`r numbers['ofi.true']` cases had at least one OFI and `r numbers['ofi.false']` cases had no OFI.
-The mean age for all patients was `r numbers['mean.age']` and `r numbers['pc.male']`% were male.
+A total of `r eligibility[5,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in Table 2.
 
-For patients with at least one OFI, the mean TRISS score was `r numbers['t.ofi.t']` and the mean NORMIT score was `r numbers['n.ofi.t']`.
-For patients with no OFI, the mean TRISS score was `r numbers['t.ofi.f']` and the mean NORMIT score was `r numbers['n.ofi.f']`.
+The median age was `r demo['age.med']` (IQR: `r demo['age.iqr']`).
+`r demo['pc.male']`% of participants were male.
+`r demo['pc.blunt']`% of cases were due to blunt force trauma.
+Of cases where 30-day survival was available, there was a  `r demo['mort']`% 30-day mortality rate (`r demo['mort.missing']` cases with missing survival data).
+Of cases where registered data on the highest level of care was available, `r demo['icu']`% had the ICU as their highest level of care (`r demo['icu.missing']` cases with missing care level data).
+The median ISS score was `r demo['iss']` (IQR: `r demo['iss.iqr']`).
+
+At least one OFI was found in `r ofi.descr['ofi.true']` of cases (`r ofi.descr['pc.ofi.true']`%).
+In cases where death occurred within 30 days, at least one OFI was found in `r ofi.descr['ofi.mort']`% of cases.
+The median scores for TRISS, NORMIT, and PS were `r scores.descr['t.med']` (IQR: `r scores.descr['t.iqr']`), `r scores.descr['n.med']` (IQR: `r scores.descr['n.iqr']`), and `r scores.descr['n.med']` (IQR: `r scores.descr['n.iqr']`) respectively.
 
@@
 
-The odds ratio for TRISS was `r or.t['or']` (95% CI: `r or.t[2]`-`r or.t[3]`, p-value: `r stats.t[['or.p']]`).
+The odds ratio for TRISS was `r or.t['or']` (95% CI: `r or.t[2]`-`r or.t[3]`, p-value: `r or.t`).
 The ROC is plotted in Figure 3.
@@
 ### NORMIT 
-The odds ratio for NORMIT was `r or.n['or']` (95% CI: `r or.n[2]`-`r or.n[3]`, p-value: `r stats.n[['or.p']]`).
+The odds ratio for NORMIT was `r or.n['or']` (95% CI: `r or.n[2]`-`r or.n[3]`, p-value: `r or.n`).
 The ROC is plotted in Figure 5.
@@
 ### PS
+The odds ratio for PS was `r or.p['or']` (95% CI: `r or.p[2]`-`r or.p[3]`, p-value: `r or.p`).
+The ROC is plotted in Figure 3.
+The AUC for the ROC was `r stats.p[['auc']]` (95% CI: `r stats.p[['auc.ci']][1]`-`r stats.p[['auc.ci']][3]`).
+The integrated calibration index for PS was `r stats.p[['ici']]`.
+The accuracy curve is shown in Figure 4.
+The best accuracy was `r stats.p[['acc.nrs']][1]` and was found at the cutoff `r stats.p[['acc.nrs']][2]`.
+
+```{r}
+plot(stats.p[['roc']], main="Fig 6. PS Receiver Operating Characteristic Curve.", colorize = TRUE)
+```
+
+```{r}
+p.plot.acc <- plot(stats.p[['acc']], main="Fig 7. PS Accuracy Curve.")
 

### 2022-11-02 functions/make_table_one.R

@@ -57,18 +57,19 @@ make_table_one <- function(cd) {
     ofi = "Opportunity for improvement",
     rts = "Revised Trauma Score",
     triss = "TRISS",
-    normit = "NORMIT"
+    normit = "NORMIT",
+    ps = "PS"
   )
   
   
   ## Making table one
   ## Removed care.level and survival bc of missing data, not sure if to include that
-  vars2 <- c("age", "gender", "iss", "niss", "dom.inj", "gcs", "ed.rr", "ed.sbp", "asa", "ofi", "rts", "triss", "normit")
-  table.1 <- table1(~ age + gender + iss + niss + dom.inj + gcs + ed.rr + ed.sbp + asa + rts + triss + normit | ofi, 
+  vars2 <- c("age", "gender", "iss", "niss", "dom.inj", "gcs", "ed.rr", "ed.sbp", "asa", "ofi", "rts", "triss", "normit", "ps")
+  table.1 <- table1(~ age + gender + iss + niss + dom.inj + gcs + ed.rr + ed.sbp + asa + rts + triss + normit + ps | ofi, 
                     data = df[,vars2], 
                     overall = "Total", 
                     render.categorical = "FREQ (PCTnoNA%)", 
-                    caption = "Table 1. Characteristics of participants."
+                    caption = "Table 2. Characteristics of participants."
                     )
   
 }

### 2022-11-02 functions/stats.R

@@ -10,7 +10,7 @@ make_stats <-function(df) {
   auc <- auc(roc)
   auc.ci <- round(ci.auc(roc), digits = 2)
 
-  ## The console output of this says that it uses "mgcv::gam instead of loess to calculate ICI"
+  ## Console output: >With at least 1000 observations, using mgcv::gam instead of loess to calculate ICI.
   ici <- ici(pred.model, df$ofi)
   ici <- round(ici, digits = 2)
 

### 2022-11-02 functions/table_one_stats.R

@@ -1,25 +1,79 @@
 
 table_one_stats <- function(df) {
+  
+  ##
+  ## overall demographics
+  ##
+  
+  demo <- c()
+  total <- as.numeric(nrow(df))
 
-  numbers <- c()
-  numbers['ofi.true'] <- nrow(df[df$ofi == 1,])
-  numbers['ofi.false'] <- nrow(df[df$ofi == 0,])
-  numbers['mean.age'] <- round(mean(df$age), digits = 1)
-  numbers['pc.male'] <- round(nrow(df[df$gender == 1,]) / nrow(df) * 100, digits = 1)
-  numbers['t.mean'] <- round(mean(df$triss), digits = 2)
-  numbers['n.mean'] <- round(mean(df$normit), digits = 2)
+  ## age  
+  demo['age.med'] <- median(df$age)
+  age.iqr <- quantile(df$age)
+  demo['age.iqr'] <- paste0(age.iqr[2], '-', age.iqr[4])
 
-  t.numbers <- df %>% group_by(ofi) %>%
-    summarise(mean = round(mean(triss), digits = 2))
+  ## gender  
+  demo['pc.male'] <- round(nrow(df[df$gender == 1,]) / total * 100, digits = 1)
+
+  ## injury type  
+  demo['pc.blunt'] <- round(nrow(df[df$dom.inj == 1,]) / total * 100, digits = 1)
+
+  ## 30 day mortality  
+  demo['mort.missing'] <- sum(is.na(df$survival)) + nrow(df[df$survival == 999 & !is.na(df$survival),])
+  total.has.survival <- total - as.numeric(demo['mort.missing'])
+  total.dead <- nrow(df[df$survival == 1 & !is.na(df$survival),])
+  demo['mort'] <- round(total.dead / total.has.survival * 100, digits = 1) ## percentage 30 day mortality of patients that have survival data
+
+  ## % needing ICU care
+  demo['icu.missing'] <- sum(is.na(df$care.level)) + nrow(df[df$care.level == 999 & !is.na(df$care.level),])
+  total.has.care.level <- total - as.numeric(demo['icu.missing'])
+  total.icu <- nrow(df[df$care.level == 5 & !is.na(df$care.level),])
[…]

### 2022-11-02 main.R

@@ -45,19 +45,19 @@ incomplete.data.tbl <- incomplete.data.tbl %>%
 df$rts <- apply(df, 1, make_rts)
 df$triss <- apply(df, 1, make_triss)
 df$normit <- apply(df, 1, make_normit)
-df$ps12 <- apply(df, 1, make_ps12)
+df$ps <- apply(df, 1, make_ps12)
 
 ## cast to numeric to avoid Problems
 df$ofi <- as.numeric(df$ofi)
 df$triss <- as.numeric(df$triss)
 df$normit <- as.numeric(df$normit)
-df$ps12 <- as.numeric(df$ps12)
+df$ps <- as.numeric(df$ps)
 
 ## some subsetted data to make be friendly to stats functions
 ## this means my stats functions can be generic in case i add scores later
 df.t <- df[, c("ofi", "triss")] %>% rename(score = triss)
 df.n <- df[, c("ofi", "normit")] %>% rename(score = normit)
-df.p <- df[, c("ofi", "ps12")] %>% rename(score = ps12)
+df.p <- df[, c("ofi", "ps")] %>% rename(score = ps)
 
 ## make_stats returns a named list with all the things i want; see stats.R
 stats.t <- make_stats(df.t)
@@ -75,8 +75,15 @@ aucdiff.tn <- auc_delta(stats.t[['roc']], stats.n[['roc']]) ## TRISS AUC - NORMI
 aucdiff.tp <- auc_delta(stats.t[['roc']], stats.p[['roc']]) ## TRISS AUC - PS12 AUC
 aucdiff.np <- auc_delta(stats.n[['roc']], stats.p[['roc']]) ## NORMIT AUC - PS12 AUC
 
+## clean the dataset
+df$triss <- round(df$triss, digits = 2)
+df$normit <- round(df$normit , digits = 2)
+df$ps <- round(df$ps , digits = 2)
+
 ## descriptive data
 table.one <- make_table_one(df)
-numbers <- table_one_stats(df) ## some manually generated descriptive statistics
-
+descr.stats <- table_one_stats(df) ## some manually generated descriptive statistics
+demo <- descr.stats[['demo']]
+ofi.descr <- descr.stats[['ofi.descr']]
+scores.descr <- descr.stats[['scores.descr']]
 

### 2022-11-02 functions/auc_delta.R
rename

@@ -5,8 +5,9 @@ auc_delta <- function(roc1, roc2) {
   diff <- roctest[['estimate']][[1]] - roctest[['estimate']][[2]]
   ci.low <- roctest[['conf.int']][[1]]
   ci.hi <- roctest[['conf.int']][[2]]
+  ci <- paste0(ci.low, '-', ci.hi)
   p <- roctest[['p.value']]
   
-  ci.diff <- c(diff = diff, ci.low = ci.low, ci.hi = ci.hi, p = p)
+  ci.diff <- c(diff = diff, ci = ci, p = p)
   return(ci.diff)
 }

### 2022-11-08 functions/acc_ci.R
made much faster

@@ -12,9 +12,9 @@ acc_ci_bs <- function(df, index) {
 
 acc_ci <- function(df, boot.no) {
   results <- boot(data = df, statistic = acc_ci_bs, R = boot.no)
-  ci <- boot.ci(boot.out = results, conf = 0.95, type="all")
+  ci <- boot.ci(boot.out = results, conf = 0.95, type=c("norm"))
   
-  out <- round(c(ci[['normal']][[2]], ci[['normal']][[3]]), digits = 2) ## choosing "normal" for CI type
-  
+  out <- round(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
+
   return(out)
 }

### 2022-11-08 functions/auc_delta.R

@@ -3,10 +3,10 @@ auc_delta <- function(roc1, roc2) {
   roctest <- roc.test(roc1, roc2, method="delong") ## has to be forced or it will try to bootstrap; bootstrap doesnt give a CI
   
   delta <- round(roctest[['estimate']][[1]] - roctest[['estimate']][[2]], digits = 2)
-  ci.low <- round(roctest[['conf.int']][[1]], digits = 2)
+  ci.lo <- round(roctest[['conf.int']][[1]], digits = 2)
   ci.hi <- round(roctest[['conf.int']][[2]], digits = 2)
   p <- round(roctest[['p.value']], digits = 2)
   
-  out <- c(delta, ci.low, ci.hi, p)
+  out <- c(delta, ci.lo, ci.hi, p)
   return(out)
 }

### 2022-11-08 functions/descriptive_data.R

@@ -39,12 +39,12 @@ descriptive_data <- function(df) {
 
   
   dd['survival.na'] <- sum(is.na(df$survival)) + nrow(df[df$survival == 999 & !is.na(df$survival),]) ## # cases missing survival data
-  survival.not.na <- total - as.numeric(dd['mort.missing']) ## # cases that have survival data
+  survival.not.na <- total - as.numeric(dd['survival.na']) ## # cases that have survival data
   deceased <- nrow(df[df$survival == 1 & !is.na(df$survival),]) ## # cases deceased in 30 days
   dd['mort'] <- round(deceased / survival.not.na * 100, digits = 1) ## 30-day mortality as % in cases where data is not missing
 
-  dd['care.missing'] <- sum(is.na(df$care.level)) + nrow(df[df$care.level == 999 & !is.na(df$care.level),]) ## # cases missing care level data
-  care.not.na <- total - as.numeric(dd['icu.missing']) ## # cases that have care level data
+  dd['care.na'] <- sum(is.na(df$care.level)) + nrow(df[df$care.level == 999 & !is.na(df$care.level),]) ## # cases missing care level data
+  care.not.na <- total - as.numeric(dd['care.na']) ## # cases that have care level data
   total.icu <- nrow(df[df$care.level == 5 & !is.na(df$care.level),]) ## # cases that were in the ICU
   dd['icu'] <- round(total.icu / care.not.na * 100, digits = 1) ## % cases with ICU as highest level of care in cases where data is not missing
 

### 2022-11-08 functions/results.R

@@ -16,11 +16,10 @@ results <-function(df, bootstrap = FALSE) {
   t <- base_stats(df.t)
   n <- base_stats(df.n)
   p <- base_stats(df.p)
-  message("Base stats computed")
   
   if (bootstrap == TRUE) {
     message("Bootstrapping enabled")
-    boot.no <- 1500
+    boot.no <- 50
     message(paste0("# of bootstrap samples: ", boot.no))
     
     ## acc_ci.R
@@ -42,17 +41,17 @@ results <-function(df, bootstrap = FALSE) {
     message("AUC deltas + CIs computed")
     d.auc <- list(tn = auc.tn, tp = auc.tp, np = auc.np)
     
-    acc.tn <- t[['acc.nrs']][['acc.max']] - n[['acc.nrs']][['acc.max']]
-    acc.tp <- t[['acc.nrs']][['acc.max']] - p[['acc.nrs']][['acc.max']]
-    acc.np <- n[['acc.nrs']][['acc.max']] - p[['acc.nrs']][['acc.max']]
+    acc.tn <- t[['acc.max']] - n[['acc.max']]
+    acc.tp <- t[['acc.max']] - p[['acc.max']]
+    acc.np <- n[['acc.max']] - p[['acc.max']]
     message("Accuracy deltas computed")
     
-    ## acc_delta.R
+    ## acc_delta_ci.R
     acc.tn.ci <- acc_delta_ci(df.tn, boot.no)
     acc.tp.ci <- acc_delta_ci(df.tp, boot.no)
     acc.np.ci <- acc_delta_ci(df.np, boot.no)
     message("Accuracy delta CIs computed")
-    delta.acc <- list(tn = acc.tn, tn.ci = acc.tn.ci, tp = acc.tp, tp.ci = acc.tp.ci, np = acc.np, np.ci = acc.np.ci)
+    d.acc <- list(tn = acc.tn, tn.ci = acc.tn.ci, tp = acc.tp, tp.ci = acc.tp.ci, np = acc.np, np.ci = acc.np.ci)
     
     ici.tn <- t[['ici']] - n[['ici']]
     ici.tp <- t[['ici']] - p[['ici']]
@@ -66,7 +65,7 @@ results <-function(df, bootstrap = FALSE) {
     message("ICI delta CIs computed")
[…]

### 2022-11-08 main.R

@@ -54,7 +54,11 @@ table.one <- table_one(df) ## table_one.R
 dd <- descriptive_data(df) ## descriptive_data.R
 
 ## get results
-results <- results(df, bootstrap = FALSE) ## results.R
+results <- results(df, bootstrap = TRUE) ## results.R
+t <- results[['t']]
+n <- results[['n']]
+p <- results[['p']]
+d <- results[['delta']]
 
 ## done
 end.time <- Sys.time()

### 2022-11-08 manuscript.Rmd
adding new data to text

--- 2022-11-02 manuscript.Rmd
@@
 Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].
-<!-- Potential things to add about NKS:
-1. How many trauma patients per year?
-2. How often do conferences happen, or alternatively, roughly how many per year?
-3. Which specialties/professions participate in conferences? 
-4. Any other info about conferences at NKS that might be relevant like if there's a particular format or it follows particular guidelines etc 
-5. Also if there's any stats about trauma patients at NKS in general, like the avg ISS or similar
-... although I guess I have the data to calculate this myself-->
-
 The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
@@
 <!-- Double check these dates, maybe give exact dates of registration in the study -->
-<!-- Also is there a time limit for when patients are registered in the registry? Like x days/weeks following admission -->
 
@@
 The trauma care quality database is an internal hospital registry and registers the outcome of M&M conferences.
+M&M conferences are held at Karolinska University Hospital ten times per year and are attended by all involved specialties and professions.
 Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.
@@
 <!-- Double check these dates -->
-
-<!-- How exactly is OFI identified, is it according to specific criteria? -->
 
@@
 
-### PS
+#### PS
 
@@
 
-RTS, NORMIT and PS12 were calculated according to the original published algorithms [@rtsalgorithm, @normit, @psalgorithm]. TRISS was calculated using the revised 2009 coefficients [@triss2010].
+RTS, NORMIT and PS12 were computed according to the original published algorithms [@rtsalgorithm; @normit; @psalgorithm]. TRISS was calculated using the revised 2009 coefficients [@triss2010].
 
@@
 ------------
-A total of `r eligibility[1,2]` patients were recorded in the trauma registry during the study period.
-Of these, `r eligibility[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion.
-`r post.ofi.exclusion` patients were excluded, due to either being of paediatric age (`r eligibility[3,3]`), being dead on arrival at hospital (`r eligibility[4,3]`), or due to incomplete data (`r eligibility[5,3]`).
+A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period.
+Of these, `r exclusion[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion.
+`r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`).
 Figure 2 shows the inclusion process for eligible participants.
@@
 ```{r}
-  exclusion_flowchart(eligibility) ## Only works in HTML
+  exclusion_flowchart(exclusion) ## Only works in HTML
 ```
@@
 ```{r}
-  as_raw_html(incomplete.data.tbl)
+ as_raw_html(table(na.data, type = "incomplete"))
 ```
@@
 ----------------
-A total of `r eligibility[5,2]` participants were identified as eligible for inclusion.
-Their demographical and clinical characteristics are described in Table 1.
-`r numbers['ofi.true']` cases had at least one OFI and `r numbers['ofi.false']` cases had no OFI.
-The mean age for all patients was `r numbers['mean.age']` and `r numbers['pc.male']`% were male.
+A total of `r exclusion[6,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in table 2.
 
-For patients with at least one OFI, the mean TRISS score was `r numbers['t.ofi.t']` and the mean NORMIT score was `r numbers['n.ofi.t']`.
-For patients with no OFI, the mean TRISS score was `r numbers['t.ofi.f']` and the mean NORMIT score was `r numbers['n.ofi.f']`.
+The median age was `r dd['age']` (IQR: `r dd['age.lo']`-`r dd['age.hi']`).
+`r dd['male']`% of participants were male.
+`r dd['blunt']`% of cases were due to blunt force trauma.
+Of cases where 30-day survival was available, there was a `r dd['mort']`% 30-day mortality rate (`r dd['survival.na']` cases with missing survival data).
+Of cases where registered data on the highest level of care was available, `r dd['icu']`% had the ICU as their highest level of care (`r dd['care.na']` cases with missing care level data).
+The median ISS score was `r dd['iss']` (IQR: `r dd['iss.lo']`-`r dd['iss.hi']`).
+
+At least one OFI was found in `r dd['ofi.true']` of cases (`r dd['ofi.pc']`%).
+In cases where death occurred within 30 days, at least one OFI was found in `r dd['ofi.mort']`% of cases.
+The median scores for TRISS, NORMIT, and PS were `r dd['t.med']` (IQR: `r dd['t.lo']`-`r dd['t.lo']`), `r dd['n.med']` (IQR: `r dd['n.lo']`-`r dd['n.hi']`), and `r dd['p.med']` (IQR: `r dd['p.lo']`-`r dd['p.hi']`) respectively.
 
@@
 
-The odds ratio for TRISS was `r or.t['or']` (95% CI: `r or.t[2]`-`r or.t[3]`, p-value: `r stats.t[['or.p']]`).
+The odds ratio for TRISS was `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, p-value: `r t[['or.p']]`).
 The ROC is plotted in Figure 3.
-The AUC for the ROC was `r stats.t[['auc']]` (95% CI: `r stats.t[['auc.ci']][1]`-`r stats.t[['auc.ci']][3]`).
-The integrated calibration index for TRISS was `r stats.t[['ici']]`.
+The AUC for the ROC was `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`).
+The integrated calibration index for TRISS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`.
 The accuracy curve is shown in Figure 4.
-The best accuracy was `r stats.t[['acc.nrs']][1]` and was found at the cutoff `r stats.t[['acc.nrs']][2]`.
+The best accuracy was `r t[['acc.max']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`) and was found at the cutoff `r t[['acc.co']]`.
 
 ```{r}
-plot(stats.t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r}
-t.plot.acc <- plot(stats.t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
+plot(t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
 ```
@@
 ### NORMIT 
-The odds ratio for NORMIT was `r or.n['or']` (95% CI: `r or.n[2]`-`r or.n[3]`, p-value: `r stats.n[['or.p']]`).
-The ROC is plotted in Figure 5.
-The AUC for the ROC was `r stats.n[['auc']]` (95% CI: `r stats.n[['auc.ci']][1]`-`r stats.n[['auc.ci']][3]`).
-The integrated calibration index for NORMIT was `r stats.n[['ici']]`.
-The accuracy curve for NORMIT is shown in Figure 5.
-The best accuracy was `r stats.n[['acc.nrs']][1]` and was found at the cutoff `r stats.n[['acc.nrs']][2]`.
+The odds ratio for NORMIT was `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, p-value: `r n[['or.p']]`).
+The ROC is plotted in Figure 3.
+The AUC for the ROC was `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`).
+The integrated calibration index for NORMIT was `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]`.
+The accuracy curve is shown in Figure 4.
+The best accuracy was `r n[['acc.max']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and was found at the cutoff `r n[['acc.co']]`.
 
 ```{r, echo = FALSE}
-plot(stats.n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+plot(n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r, echo = FALSE}
-n.plot.acc <- plot(stats.n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
+plot(n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
 ```
@@
 ### PS
+The odds ratio for PS was `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, p-value: `r p[['or.p']]`).
+The ROC is plotted in Figure 3.
+The AUC for the ROC was `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`).
+The integrated calibration index for PS was `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]`.
+The accuracy curve is shown in Figure 4.
+The best accuracy was `r p[['acc.max']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) and was found at the cutoff `r p[['acc.co']]`.
+
+```{r}
+plot(p[['roc']], main="Fig 6. PS Receiver Operating Characteristic Curve.", colorize = TRUE)
+```
+
+```{r}
+plot(p[['acc']], main="Fig 7. PS Accuracy Curve.")
+```
 
 ### Comparison of scores
+
+Table 3 shows a comparison of ?? for the three models analysed.
+
+<!-- Make a table here? after I've run the full script -->
+
+```{r}
+  table(results, type = "or.plot")
+```
 

### 2022-11-08 functions/base_stats.R

@@ -25,23 +25,23 @@ base_stats <-function(df) {
   
   ## accuracy max + cutoff
   acc <- performance(pred, measure = "acc") ## plottable
-  ind <- which.max(slot(acc, "y.values")[[1]] ) ## find the index of the highest accuracy
+  ind <- which.max(slot(acc, "y.values")[[1]]) ## find the index of the highest accuracy
   acc.max <- slot(acc, "y.values")[[1]][ind] ## stores the highest accuracy
-  acc.max <- acc.max
+  acc.max <- acc.max[[1]]
   acc.co <- slot(acc, "x.values")[[1]][ind] ## stores the cutoff at the highest accuracy
-  acc.co <- acc.co
+  acc.co <- acc.co[[1]]
   
   numerics <- c(
     auc = auc, 
-    acc.max = acc.max,
-    acc.co = acc.co,
     ici = ici, 
     auc.ci.lo = auc.ci.lo,
     auc.ci.hi = auc.ci.hi,
     or = or, 
     or.ci.lo = or.ci.lo,
     or.ci.hi = or.ci.hi,
-    or.p = or.p
+    or.p = or.p,
+    acc.max = acc.max,
+    acc.co = acc.co
   )
   
   numerics <- lapply(numerics, round, digits = 2)

### 2022-11-08 functions/descriptive_data.R

@@ -6,32 +6,32 @@ descriptive_data <- function(df) {
 
   ## some medians 
   dd <- c(
-    "age" = median(df$age), 
-    "age.lo" = quantile(df$age)[[2]], 
-    "age.hi" = quantile(df$age)[[4]], 
-    "iss" = median(df$iss),
-    "iss.lo" = quantile(df$iss)[[2]], 
-    "iss.hi" = quantile(df$iss)[[4]], 
-    "ofi.true" = nrow(df[df$ofi == 1,]),
-    "ofi.false" = nrow(df[df$ofi == 0,]),
-    "t.med" = median(df$triss),
-    "t.lo" = quantile(df$triss)[[2]], 
-    "t.hi" = quantile(df$triss)[[4]], 
-    "n.med" = median(df$normit),
-    "n.lo" = quantile(df$normit)[[2]], 
-    "n.hi" = quantile(df$normit)[[4]], 
-    "p.med" = median(df$ps),
-    "p.lo" = quantile(df$ps)[[2]], 
-    "p.hi" = quantile(df$ps)[[4]]
+    age = median(df$age), 
+    age.lo = quantile(df$age)[[2]], 
+    age.hi = quantile(df$age)[[4]], 
+    iss = median(df$iss),
+    iss.lo = quantile(df$iss)[[2]], 
+    iss.hi = quantile(df$iss)[[4]], 
+    ofi.true = nrow(df[df$ofi == 1,]),
+    ofi.false = nrow(df[df$ofi == 0,]),
+    t.med = median(df$triss),
+    t.lo = quantile(df$triss)[[2]], 
+    t.hi = quantile(df$triss)[[4]], 
+    n.med = median(df$normit),
+    n.lo = quantile(df$normit)[[2]], 
+    n.hi = quantile(df$normit)[[4]], 
+    p.med = median(df$ps),
+    p.lo = quantile(df$ps)[[2]], 
+    p.hi = quantile(df$ps)[[4]]
   )   
   
   dd <- sapply(dd, round, digits = 2)
   
   ## some percentages
   pc <- c(
-    "ofi.pc" = as.numeric(nrow(df[df$ofi == 1,])) / total * 100, ## ofi as % true
-    "male" = nrow(df[df$gender == 1,]) / total * 100, ## gender as % male
-    "blunt" = nrow(df[df$dom.inj == 1,]) / total * 100 ## injury type as % blunt
[…]

### 2022-11-08 functions/prep_data.R

@@ -59,16 +59,16 @@ prep_data <- function(df, numbers = FALSE) {
   
   ## Recording data on each missing parameter
   na.data <- c(
-    "na.gcs" = sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),]),
-    "na.asa" = sum(is.na(df$asa)) + nrow(df[df$asa == 999 & !is.na(df$asa),]),
-    "na.rr" = sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0 & !is.na(df$ed.rr),]),
-    "na.sbp" = sum(is.na(df$ed.sbp)) + nrow(df[df$ed.sbp == 0 & !is.na(df$ed.sbp),]),
-    "na.dominj" = sum(is.na(df$dom.inj)) + nrow(df[df$dom.inj == 999 & !is.na(df$dom.inj),]),
-    "na.age" = sum(is.na(df$age)),
-    "na.iss" = sum(is.na(df$iss)),
-    "na.niss" = sum(is.na(df$niss)),
-    "na.gender" = sum(is.na(df$gender)) + nrow(df[df$gender == 999 & !is.na(df$gender),])
-    )
+    gcs = sum(is.na(df$ed.gcs)) + nrow(df[df$ed.gcs == 999 & !is.na(df$ed.gcs),]),
+    asa = sum(is.na(df$asa)) + nrow(df[df$asa == 999 & !is.na(df$asa),]),
+    rr = sum(is.na(df$ed.rr)) + nrow(df[df$ed.rr == 0 & !is.na(df$ed.rr),]),
+    sbp = sum(is.na(df$ed.sbp)) + nrow(df[df$ed.sbp == 0 & !is.na(df$ed.sbp),]),
+    dominj = sum(is.na(df$dom.inj)) + nrow(df[df$dom.inj == 999 & !is.na(df$dom.inj),]),
+    age = sum(is.na(df$age)),
+    iss = sum(is.na(df$iss)),
+    niss = sum(is.na(df$niss)),
+    gender = sum(is.na(df$gender)) + nrow(df[df$gender == 999 & !is.na(df$gender),])
+  )
                                             
   ## Parameter exclusion
   df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa, gender),all_vars(!is.na(.)))
@@ -80,7 +80,7 @@ prep_data <- function(df, numbers = FALSE) {
[…]

### 2022-11-08 functions/results.R

@@ -19,50 +19,50 @@ results <-function(df, bootstrap = FALSE) {
   
   if (bootstrap == TRUE) {
     message("Bootstrapping enabled")
-    boot.no <- 50
+    boot.no <- 25
     message(paste0("# of bootstrap samples: ", boot.no))
     
     ## acc_ci.R
     t <- list.append(t, "acc.ci" = acc_ci(df.t, boot.no))
     n <- list.append(n, "acc.ci" = acc_ci(df.n, boot.no))
     p <- list.append(p, "acc.ci" = acc_ci(df.p, boot.no))
-    message("Accuracy CIs computed")
+    message("Accuracy CIs done")
     
     ## ici_ci.R
     t <- list.append(t, "ici.ci" = ici_ci(df.t, boot.no))
     n <- list.append(n, "ici.ci" = ici_ci(df.n, boot.no))
     p <- list.append(p, "ici.ci" = ici_ci(df.p, boot.no))
-    message("ICI CIs computed")
+    message("ICI CIs done")
     
     ## auc_delta.R
     auc.tn <- auc_delta(t[['roc']], n[['roc']]) ## TRISS AUC - NORMIT AUC
     auc.tp <- auc_delta(t[['roc']], p[['roc']]) ## TRISS AUC - PS AUC
     auc.np <- auc_delta(n[['roc']], p[['roc']]) ## NORMIT AUC - PS AUC
-    message("AUC deltas + CIs computed")
+    message("AUC deltas + CIs done")
     d.auc <- list(tn = auc.tn, tp = auc.tp, np = auc.np)
     
     acc.tn <- t[['acc.max']] - n[['acc.max']]
     acc.tp <- t[['acc.max']] - p[['acc.max']]
     acc.np <- n[['acc.max']] - p[['acc.max']]
-    message("Accuracy deltas computed")
+    message("Accuracy deltas done")
     
     ## acc_delta_ci.R
     acc.tn.ci <- acc_delta_ci(df.tn, boot.no)
     acc.tp.ci <- acc_delta_ci(df.tp, boot.no)
     acc.np.ci <- acc_delta_ci(df.np, boot.no)
-    message("Accuracy delta CIs computed")
+    message("Accuracy delta CIs done")
     d.acc <- list(tn = acc.tn, tn.ci = acc.tn.ci, tp = acc.tp, tp.ci = acc.tp.ci, np = acc.np, np.ci = acc.np.ci)
     
[…]

### 2022-11-08 functions/table.R

@@ -1,0 +1,27 @@
+table <- function(data, type = "") {
+  
+  if(type == "incomplete") {
+    
+    names <- c("GCS", "ASA", "RR", "Systolic BP", "Dominant Injury Type", "Age", "ISS", "NISS", "Gender")
+    tib <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
+    out <- tib %>% gt() %>% tab_header(
+      title = "Table 1. Paramaters that were missing in cases excluded for incomplete data"
+    )
+    
+    return(out)
+  }
+
+  if(type == "or.plot") {
+    nrs <- data.frame(
+      score = c("TRISS, NORMIT, PS"),
+      or = c(data[['t']][['or']], data[['n']][['or']], data[['p']][['or']]),
+      lower = c(data[['t']][['or.ci.lo']], data[['n']][['or.ci.lo']], data[['p']][['or.ci.lo']]),
+      upper = c(data[['t']][['or.ci.hi']], data[['n']][['or.ci.hi']], data[['p']][['or.ci.hi']])
+    )
+    
+    out <- ggplot(nrs, aes(x = factor(score), y = or)) + geom_point() + geom_errorbar(aes(ymin = lower, ymax = upper)) + labs(title = "Figure 8. Comparison of Odds Ratios.")
+    return(out)
+  }
+  
+
+}

### 2022-11-08 main.R

@@ -18,10 +18,14 @@ library(pROC)
 library(gt)
 library(gtsummary)
 library(boot)
-library('rlist')
+library(rlist)
+library(ggplot2)
 
 ## TODO
 ## something else for boot.ci so I can run multiple stats in one bootstrap?
+## ci.boot level 0.90 or 0.95
+## ask if ok normit 2
+## how to present delta statistics?
 
 noacsr::source_all_functions()
 

### 2022-11-15 manuscript.Rmd
rounding, little bit of discussion

@@
 output:
+  pdf_document: default
   html_document:
@@
     number_sections: true
-  pdf_document: default
   word_document: default
@@
 * RTS - Revised Trauma Score
+* TARN - Trauma Audit & Research Network
 * TRISS - Trauma and Injury Severity Score
@@
 TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
-The resulting value the probability of survival given as a percentage.
-The ISS gives a score based on the three anatomical regions with the most severe injury.
-A limitation of the ISS is that it does not account for situations where one anatomical region has multiple injuries [@trissbetter].
-The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate.
-The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating probability of survival in certain populations [@trissbetter].
+The RTS requires GCS on arrival at hospital, which cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained.
+This gives TRISS limited value in estimating probability of survival in certain populations [@trissbetter].
+Another commonly used model is the Probability of Survival (PS) score, originally developed by the Trauma Audit & Research Network (TARN) in the UK in 2007 [@tarn07].
+In some ways the PS model is more complex than TRISS; it has coefficients for sex associated with age, and provides a formula that takes into account the non-linearity of ISS [@tarn07].
 
@@
 For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit].
-NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit].
+NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when used at the national level [@validationnormit].
 
@@
 Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
+Some hospitals use trauma severity models as audit filters is to only review deaths in cases that had a high predicted probability of survival, the idea being that deaths in cases with high predicted survival are more likely to have been preventable deaths.
+Research to date does not support this use case.
+Two major studies have looked at trauma severity models and preventable deaths.
+One study in Sweden investigated TRISS and NORMIT [@prevdeathsghorbani] and a study in the UK investigated TRISS and PS [@brohi16].
+Both of these studies showed that the studied models were unable to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani; @brohi16].
+
 However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
-A study at Karolinska University Hospital found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani].
-However, that study included only deaths and not other morbidities and thereby had a low sample size.
-It is not known whether trauma severity models have better performance when applied to the entire population of trauma patients.
-
-If it could be found that an existing trauma severity model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion.
-Conversely, if trauma severity models are not associated with OFI, using trauma severity models as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.
-
-There are far more trauma severity models than the two described above.
-These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is where this institution is located.
-It seems reasonable to begin by investigating these two models.
-
-<!-- S says to clarify the knowledge gap & motive -->
+The two studies mentioned included only deaths and not other morbidities and thereby had low sample sizes.
+To date there is no published research on the ability of trauma severity models to predict OFI across the board and not just in mortalities.
+If it could be found that an existing trauma severity model has value in predicting OFI, it would confirm the practice of using models as audit filters and would improve case selection and reduce resource burden.
+Conversely, if trauma severity models cannot reliably predict OFI, it is possible that using these models as audit filters causes cases to be unnecessarily flagged for review, which in turn wastes hospital resources.
+Either way, the answer to this question has the potential to benefit trauma QI programmes.
 
 ## Aim
-The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.
+The aim of this study is to evaluate how the trauma severity models TRISS, NORMIT and PS are associated with and to what extent they can predict opportunities for improvement.
 
@@
 Data is collected from the patient's electronic medical records.
-The trauma registry contains data on patients from 2012 to 2021.
-<!-- Double check these dates, maybe give exact dates of registration in the study -->
+The trauma registry contains data on patients from 2013 to 2021.
+<!-- Based on the dates I see in the Swetrau dataset -->
 
@@
 
-``` {r} 
-  ofi_diagram() ## Only works in HTML
+```{r, echo=FALSE, out.width = "400px"}
+## knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
 ```
@@
 
-```{r}
-  exclusion_flowchart(exclusion) ## Only works in HTML
+```{r, echo=FALSE, out.width = "400px"}
+## knitr::include_graphics("images/exclusion.png") ## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
 ```
@@
 ```{r}
- as_raw_html(table(na.data, type = "incomplete"))
+ as_latex(table(na.data, type = "incomplete"))
 ```
@@
 ```{r}
-plot(t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+## plot(t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r}
-plot(t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
+## plot(t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
 ```
@@
 ```{r, echo = FALSE}
-plot(n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+## plot(n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r, echo = FALSE}
-plot(n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
+## plot(n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
 ```
@@
 ```{r}
-plot(p[['roc']], main="Fig 6. PS Receiver Operating Characteristic Curve.", colorize = TRUE)
+## plot(p[['roc']], main="Fig 6. PS Receiver Operating Characteristic Curve.", colorize = TRUE)
 ```
@@
 ```{r}
-plot(p[['acc']], main="Fig 7. PS Accuracy Curve.")
+## plot(p[['acc']], main="Fig 7. PS Accuracy Curve.")
 ```
@@
 
-<!-- Make a table here? after I've run the full script -->
-
 ```{r}
-  table(results, type = "or.plot")
+  ### table(results, type = "or.plot")
+  ### table(d, type = "deltas")
+ ## as_latex(table(d, type = "deltas"))
 ```
@@
 
-<!-- 
 Key results
 -----------
-Summarise key results with reference to study objectives
+<!-- Summarise key results with reference to study objectives -->
 
--->
+This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI.
+These three models did not differ significantly in their results.
+The main findings indicate that increased probability of survival was associated with lower odds for OFI.
+<!-- OR --> All three models showed poor diagnostic ability in discriminating between OFI and no OFI (AUCs ranged from `r p[['auc']]` to `r t[['auc']]`.
+<!-- AUC -->
 
-Limitations
+Strengths and limitations
 -----------
-
 <!-- 
@@
 of any potential bias -->
+
+One of the strengths of this study is the duration of enrollment which made it possible to have a large study population.
+Karolinska University Hospital is the primary trauma centre for a catchment population of 3 million, which means that this hospital-based study was in effect a population-based study, which reduces the risk of selection bias.
+All patients that met inclusion criteria were included in the study.
+The code for data analysis was written by a single student researcher whilst blinded to the real-world data to minimise the risk of researcher bias.
+
+<!-- something about how staff at confs may have differed over the years -->
 

### 2022-11-15 functions/acc_ci.R

@@ -14,7 +14,7 @@ acc_ci <- function(df, boot.no) {
   results <- boot(data = df, statistic = acc_ci_bs, R = boot.no)
   ci <- boot.ci(boot.out = results, conf = 0.95, type=c("norm"))
   
-  out <- round(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
-
+  out <- signif(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
+  
   return(out)
 }

### 2022-11-15 functions/acc_delta_ci.R

@@ -25,7 +25,7 @@ acc_delta_ci <- function(df, boot.no) {
   results <- boot(data = df, statistic = acc_delta_bs, R = boot.no)
   ci <- boot.ci(boot.out = results, conf = 0.95, type=c("norm"))
   
-  out <- round(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
+  out <- c(ci[['normal']][2], ci[['normal']][3])
   
   return(out)
 }

### 2022-11-15 functions/auc_delta.R

@@ -2,10 +2,10 @@ auc_delta <- function(roc1, roc2) {
   
   roctest <- roc.test(roc1, roc2, method="delong") ## has to be forced or it will try to bootstrap; bootstrap doesnt give a CI
   
-  delta <- round(roctest[['estimate']][[1]] - roctest[['estimate']][[2]], digits = 2)
-  ci.lo <- round(roctest[['conf.int']][[1]], digits = 2)
-  ci.hi <- round(roctest[['conf.int']][[2]], digits = 2)
-  p <- round(roctest[['p.value']], digits = 2)
+  delta <- signif(roctest[['estimate']][[1]] - roctest[['estimate']][[2]], 2)
+  ci.lo <- signif(roctest[['conf.int']][[1]], 2)
+  ci.hi <- signif(roctest[['conf.int']][[2]], 2)
+  p <- format(signif(roctest[['p.value']], 1), scientific = FALSE)
   
   out <- c(delta, ci.lo, ci.hi, p)
   return(out)

### 2022-11-15 functions/base_stats.R

@@ -10,7 +10,7 @@ base_stats <-function(df) {
   or.ci <- exp(confint.default(model, level = 0.90))
   or.ci.lo <- or.ci[2,1]
   or.ci.hi <- or.ci[2,2]
-  or.p <- sum$coefficients[2,4]
+  or.p <- format(signif(sum$coefficients[2,4], 1), scientific = FALSE)
   
   ## ROC/AUC with pROC
   roc <- suppressMessages(roc(ofi ~ score, data = df))
@@ -39,19 +39,20 @@ base_stats <-function(df) {
     or = or, 
     or.ci.lo = or.ci.lo,
     or.ci.hi = or.ci.hi,
-    or.p = or.p,
     acc.max = acc.max,
     acc.co = acc.co
   )
   
-  numerics <- lapply(numerics, round, digits = 2)
+  numerics <- lapply(numerics, as.numeric)
+  numerics <- lapply(numerics, signif, digits = 2)
   
   out <- list(
     model = model, 
     roc = roc, 
-    acc = acc
+    acc = acc,
+    or.p = or.p
   )
-
+  
   out <- append(out, numerics)
   
   return(out)

### 2022-11-15 functions/descriptive_data.R

@@ -1,9 +1,9 @@
 ## manually calculating some of the stats found in table one
 
 descriptive_data <- function(df) {
-
+  
   total <- as.numeric(nrow(df))
-
+  
   ## some medians 
   dd <- c(
     age = median(df$age), 
@@ -36,18 +36,18 @@ descriptive_data <- function(df) {
   
   pc <- sapply(pc, round, digits = 1)
   dd <- append(dd, pc)
-
+  
   
   dd['survival.na'] <- sum(is.na(df$survival)) + nrow(df[df$survival == 999 & !is.na(df$survival),]) ## # cases missing survival data
   survival.not.na <- total - as.numeric(dd['survival.na']) ## # cases that have survival data
   deceased <- nrow(df[df$survival == 1 & !is.na(df$survival),]) ## # cases deceased in 30 days
   dd['mort'] <- round(deceased / survival.not.na * 100, digits = 1) ## 30-day mortality as % in cases where data is not missing
-
+  
   dd['care.na'] <- sum(is.na(df$care.level)) + nrow(df[df$care.level == 999 & !is.na(df$care.level),]) ## # cases missing care level data
   care.not.na <- total - as.numeric(dd['care.na']) ## # cases that have care level data
   total.icu <- nrow(df[df$care.level == 5 & !is.na(df$care.level),]) ## # cases that were in the ICU
   dd['icu'] <- round(total.icu / care.not.na * 100, digits = 1) ## % cases with ICU as highest level of care in cases where data is not missing
-
+  
   ## In cases where death occurred within 30 days, at least one OFI was found in % of cases.
   df.ofi.mort <- df[df$survival == 1 & !is.na(df$survival),]
   dd['ofi.mort'] <- round(nrow(df.ofi.mort[df.ofi.mort$ofi == 1,]) / survival.not.na * 100, digits = 1)

### 2022-11-15 functions/exclusion_flowchart.R

@@ -42,7 +42,7 @@ exclusion_flowchart <- function(exclusion) {
   [7]: paste0('Excluded due to <15 years of age \\n(n = ', exclusion[3,3], ')')
   [8]: paste0('Excluded due to DOA \\n(n = ', exclusion[4,3], ')')
   [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n(n = ', exclusion[5,3], ')')
-                               
                                ")
   return(out)
 }
+

### 2022-11-15 functions/ici_ci.R

@@ -6,7 +6,7 @@ ici_bs <- function(df, index) {
   pred.model <- predict(model, type="response")
   ici <- suppressMessages(ici(pred.model, sample$ofi))
   ici <- as.numeric(ici)
-
+  
   return(ici)
 }
 
@@ -14,7 +14,7 @@ ici_ci <- function(df, boot.no) {
   results <- boot(data = df, statistic = ici_bs, R = boot.no)
   ci <- boot.ci(boot.out = results, conf = 0.95, type=c("norm"))
   
-  out <- round(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
+  out <- signif(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
   
   return(out)
 }

### 2022-11-15 functions/ici_delta_ci.R

@@ -21,7 +21,7 @@ ici_delta_ci <- function(df, boot.no) {
   results <- boot(data = df, statistic = ici_delta_bs, R = boot.no)
   ci <- boot.ci(boot.out = results, conf = 0.95, type=c("norm"))
   
-  out <- round(c(ci[['normal']][2], ci[['normal']][3]), digits = 2)
+  out <- signif(c(ci[['normal']][2], ci[['normal']][3]), 2)
   
   return(out)
 }

### 2022-11-15 functions/make_normit.R

@@ -7,8 +7,8 @@ make_normit <- function(input.data) {
     asa = c(1, 2, 3, 4),
     b1 = c(-0.0713, -0.0565, -0.0487, -0.0081),
     b2 = c(0.6266, -0.2142, -0.8971, -3.8746)
-    )
-
+  )
+  
   ## Cast variables to numeric to ensure they can be used for calculations  
   n.niss <- as.numeric(input.data["niss"])
   n.rts <- as.numeric(input.data["rts"])
@@ -28,4 +28,4 @@ make_normit <- function(input.data) {
   
   return(out)
   
- }
+}

### 2022-11-15 functions/make_ps.R

@@ -1,75 +1,27 @@
-## https://www.tarn.ac.uk/Content.aspx?c=1895
+## Coefficients from Figure 1 here: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5813212/
+## i.e. NORMIT 2
 
-make_ps <- function (input.data) {
+make_normit <- function(input.data) {
+  
+  lookup <- data.frame(
+    asa = c(1, 2, 3, 4),
+    b1 = c(-0.0713, -0.0565, -0.0487, -0.0081),
+    b2 = c(0.6266, -0.2142, -0.8971, -3.8746)
+  )
   
   ## Cast variables to numeric to ensure they can be used for calculations  
-  ed.gcs <- as.numeric(input.data["ed.gcs"])
-  gcs <- as.numeric(input.data["gcs"] )
-  age <- as.numeric(input.data["age"])
-  gender <- as.numeric(input.data["gender"])
-  iss <- as.numeric(input.data["iss"])
+  n.niss <- as.numeric(input.data["niss"])
+  n.rts <- as.numeric(input.data["rts"])
+  n.asa <- as.numeric(input.data["asa"])
+  n.age <- as.numeric(input.data["age"])
   
-  b0 <- 4.9146
-
-  ## ed.gcs == 99 -> intubated
-  if (ed.gcs == 99) {gcs.val <- -2.31186} ## b6
-    else { 
-    if (ed.gcs >= 13) {gcs.val <- 0} ## b1
-    else if (ed.gcs >= 9) {gcs.val <- -1.27734} ## b2
-    else if (ed.gcs >= 5) {gcs.val <- -1.68936} ## b3
-    else if (ed.gcs >= 4) {gcs.val <- -2.52661} # b4
-    else {gcs.val <- -3.62339} #b5
-    }
+  ## Select the appropriate coefficients based on asa
+  coef <- lookup[lookup$asa == n.asa,]
   
-  ## coefficient for gender
-  if (gender == 1) {gender.val <- 0} # b9
-    else {gender.val <- -0.024416} #b10
-  
-  
-  ## coefficient for age; ignored coefficients for ages <15
-  if (age > 74) {age.val <- -3.01315} ## b18
-    else if (age > 64) {age.val <- -1.74081} ## b17
-    else if (age > 54) {age.val <- -0.995816} ## b16
-    else if (age > 44) {age.val <- -0.557926} ## b15
[…]

### 2022-11-15 functions/make_rts.R

@@ -15,12 +15,12 @@ make_rts <- function (input.data) {
     RRmin  = c(0, 1, 6, 30, 10),
     RRmax = c(0, 5, 9, 10000, 29)
   )
-
+  
   ## Cast variables to numeric to ensure they can be used for calculations
   gcs <- as.numeric(input.data["gcs"])
   rr <- as.numeric(input.data["ed.rr"])
   sbp <- as.numeric(input.data["ed.sbp"])
-
+  
   ## Select the appropriate score value for each parameter
   r.gcs <-  lookup.rts$score.value[gcs >= lookup.rts$GCSmin & gcs <= lookup.rts$GCSmax]
   r.sbp <-  lookup.rts$score.value[sbp >= lookup.rts$SBPmin & sbp <= lookup.rts$SBPmax]

### 2022-11-15 functions/make_triss.R

@@ -11,7 +11,7 @@ make_triss <- function(input.data) {
   gcs <- as.numeric(input.data["gcs"])
   sbp <- as.numeric(input.data["ed.sbp"])
   rr <- as.numeric(input.data["ed.rr"])
-
+  
   if (dom.inj == 1) { ## 1 = blunt
     b0 <- 1.6494 ## intercept
     b1 <- 0.0095 ## RR

### 2022-11-15 functions/ofi_diagram.R

@@ -1,8 +1,7 @@
 ## Makes a flowchart on how patients are selected for M&M
 
 ofi_diagram <- function() {
-  
-  out <- DiagrammeR::grViz("digraph flowchart {
+  diagram <- DiagrammeR::grViz("digraph flowchart {
   
     label='Fig. 1. OFI decision-making workflow';
     labelloc=bottom;
@@ -33,12 +32,11 @@ ofi_diagram <- function() {
     [6]: 'Opportunities for improvement found'
     [7]: 'No opportunities for improvement found'
   ") 
-
-  ## %>% DiagrammeRsvg::export_svg() %>% charToRaw() %>% rsvg::rsvg_pdf("swetrau_flowchart.pdf")
-  ## Cannot to install DiagrammeRsvg b/c can't install v8 b/c "fatal error: v8.h: no such file or directory"
-  return(out)
-
-}
   
+  return(diagram)
+}
 
+save_ofi_png <- function(diagram) {
+  diagram %>% export_svg() %>% charToRaw() %>% rsvg::rsvg_png("images/ofi.png")
+}
 

### 2022-11-15 functions/prep_data.R

@@ -1,7 +1,8 @@
 ## All values that are necessary: iss, niss, age, dom.inj, gcs, sbp, rr, asa, pre.gcs if ed.gcs == 99
 
-prep_data <- function(df, numbers = FALSE) {
+prep_data <- function(datasets) {
   
+  df <- merge_data(datasets)
   df$ofi <- create_ofi(df)
   
   ## Kept some possibly unnecessary variables in case they're interesting to have in table one
@@ -19,7 +20,7 @@ prep_data <- function(df, numbers = FALSE) {
               "ed_sbp_value",
               "pre_gcs_sum",
               "pt_age_yrs"
-              )]
+  )]
   
   df <- df %>% rename(gender = pt_Gender,
                       survival = res_survival,
@@ -34,15 +35,15 @@ prep_data <- function(df, numbers = FALSE) {
                       ed.sbp = ed_sbp_value,
                       pre.gcs = pre_gcs_sum,
                       age = pt_age_yrs
-                      )
+  )
   
   ## Storing the inclusion/exclusion counts at each step
   exclusion <- data.frame (step  = c("original"),
-                    included = c(nrow(df)),
-                    excluded = c(0))
+                           included = c(nrow(df)),
+                           excluded = c(0))
   
   ## OFI exclusion
-
+  
   df <- df %>% filter (!is.na(ofi))
   exclusion[nrow(exclusion) + 1,] <- c("ofi", nrow(df), exclusion[1,2] - nrow(df))
   
@@ -69,7 +70,7 @@ prep_data <- function(df, numbers = FALSE) {
     niss = sum(is.na(df$niss)),
     gender = sum(is.na(df$gender)) + nrow(df[df$gender == 999 & !is.na(df$gender),])
   )
-                                            
+  
   ## Parameter exclusion
   df <- df %>% filter_at(vars(iss, niss, age, ed.gcs, dom.inj, ed.sbp, ed.rr, asa, gender),all_vars(!is.na(.)))
   df <- df %>% filter(dom.inj != 999 & asa != 999 & ed.gcs != 999 & gender != 999)
@@ -99,15 +100,25 @@ prep_data <- function(df, numbers = FALSE) {
   df$ofi <- as.numeric(df$ofi)
   
[…]

### 2022-11-15 functions/results.R

@@ -2,7 +2,7 @@
 ## see results.md for a list of all the variables created in this function
 ## bootstrap TRUE/FALSE to turn off long loading times when testing things
 
-results <-function(df, bootstrap = FALSE) {
+results <-function(df, bootstrap = FALSE, boot.no = boot.no) {
   
   ## some subsetted data to be friendly to stats functions
   df.t <- df[, c("ofi", "triss")] %>% rename(score = triss)
@@ -12,14 +12,13 @@ results <-function(df, bootstrap = FALSE) {
   df.tp <- df[, c("ofi", "triss", "ps")] %>% rename(score1 = triss, score2 = ps)
   df.np <- df[, c("ofi", "normit", "ps")] %>% rename(score1 = normit, score2 = ps)
   
-  ## make "base" stats i.e. non-bootstrapped functions; see base_results.R
+  ## make "base" stats i.e. non-bootstrapped functions; base_results.R
   t <- base_stats(df.t)
   n <- base_stats(df.n)
   p <- base_stats(df.p)
   
   if (bootstrap == TRUE) {
     message("Bootstrapping enabled")
-    boot.no <- 25
     message(paste0("# of bootstrap samples: ", boot.no))
     
     ## acc_ci.R
@@ -41,9 +40,9 @@ results <-function(df, bootstrap = FALSE) {
     message("AUC deltas + CIs done")
     d.auc <- list(tn = auc.tn, tp = auc.tp, np = auc.np)
     
-    acc.tn <- t[['acc.max']] - n[['acc.max']]
-    acc.tp <- t[['acc.max']] - p[['acc.max']]
-    acc.np <- n[['acc.max']] - p[['acc.max']]
+    acc.tn <- signif(t[['acc.max']] - n[['acc.max']], 2)
+    acc.tp <- signif(t[['acc.max']] - p[['acc.max']], 2)
+    acc.np <- signif(n[['acc.max']] - p[['acc.max']], 2)
     message("Accuracy deltas done")
     
     ## acc_delta_ci.R
@@ -53,9 +52,9 @@ results <-function(df, bootstrap = FALSE) {
     message("Accuracy delta CIs done")
     d.acc <- list(tn = acc.tn, tn.ci = acc.tn.ci, tp = acc.tp, tp.ci = acc.tp.ci, np = acc.np, np.ci = acc.np.ci)
     
-    ici.tn <- t[['ici']] - n[['ici']]
[…]

### 2022-11-15 functions/table.R

@@ -5,12 +5,12 @@ table <- function(data, type = "") {
     names <- c("GCS", "ASA", "RR", "Systolic BP", "Dominant Injury Type", "Age", "ISS", "NISS", "Gender")
     tib <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
     out <- tib %>% gt() %>% tab_header(
-      title = "Table 1. Paramaters that were missing in cases excluded for incomplete data"
+      title = "Table 1. Parameters that were missing in cases excluded for incomplete data"
     )
     
     return(out)
   }
-
+  
   if(type == "or.plot") {
     nrs <- data.frame(
       score = c("TRISS, NORMIT, PS"),
@@ -19,9 +19,40 @@ table <- function(data, type = "") {
       upper = c(data[['t']][['or.ci.hi']], data[['n']][['or.ci.hi']], data[['p']][['or.ci.hi']])
     )
     
-    out <- ggplot(nrs, aes(x = factor(score), y = or)) + geom_point() + geom_errorbar(aes(ymin = lower, ymax = upper)) + labs(title = "Figure 8. Comparison of Odds Ratios.")
+    out <- ggplot(nrs, aes(x = factor(score), y = or)) + 
+      geom_point(position=position_dodge(.9)) + 
+      geom_errorbar(aes(ymin = lower, ymax = upper), position=position_dodge(.9), width = .25) + 
+      labs(
+        title = "Figure 8. Comparison of Odds Ratios.",
+        y = "Odds Ratio",
+        x = "Model"
+      )
     return(out)
   }
   
-
+  if(type == "deltas") {
+    comparisons <- c("TRISS - NORMIT", "TRISS - PS", "NORMIT - PS")
+    auc <- c(
+      paste0(data[['auc']][['tn']][1]," (", data[['auc']][['tn']][[2]], "-", data[['auc']][['tn']][[3]], ")"),
+      paste0(data[['auc']][['tp']][1]," (", data[['auc']][['tp']][[2]], "-", data[['auc']][['tp']][[3]], ")"),
+      paste0(data[['auc']][['np']][1]," (", data[['auc']][['np']][[2]], "-", data[['auc']][['np']][[3]], ")")
+    )
+    ici <- c(
+      paste0(data[['ici']][['tn']]," (", data[['ici']][['tn']][[1]], "-", data[['ici']][['tn']][[1]], ")"),
[…]

### 2022-11-15 functions/table_one.R

@@ -61,10 +61,10 @@ table_one <- function(cd) {
   ## Removed care.level and survival bc of missing data, not sure if to include that
   vars2 <- c("age", "gender", "iss", "niss", "dom.inj", "gcs", "ed.rr", "ed.sbp", "asa", "ofi", "rts", "triss", "normit", "ps")
   out <- table1(~ age + gender + iss + niss + dom.inj + gcs + ed.rr + ed.sbp + asa + rts + triss + normit + ps | ofi, 
-                    data = df[,vars2], 
-                    overall = "Total", 
-                    render.categorical = "FREQ (PCTnoNA%)", 
-                    caption = "Table 2. Characteristics of participants."
-                    )
+                data = df[,vars2], 
+                overall = "Total", 
+                render.categorical = "FREQ (PCTnoNA%)", 
+                caption = "Table 2. Characteristics of participants."
+  )
   return(out)
 }

### 2022-11-15 main.R

@@ -1,6 +1,5 @@
 start.time <- Sys.time()
 
-## Libraries I tried to install but failed: rsvg/diagrammersvg, kableextra
 library(rofi)
 library(dplyr)
 library(labelled)
@@ -20,12 +19,21 @@ library(gtsummary)
 library(boot)
 library(rlist)
 library(ggplot2)
+## library(kableExtra)
+## library(DiagrammeRsvg)
+## library(rsvg)
+library(mvbutils)
 
 ## TODO
 ## something else for boot.ci so I can run multiple stats in one bootstrap?
 ## ci.boot level 0.90 or 0.95
 ## ask if ok normit 2
 ## how to present delta statistics?
+## delta OR? 
+## p-values???
+## fix rounding of OR p value?
+## why is max accuracy the exact same for all 3?
+## find sources about ps12
 
 noacsr::source_all_functions()
 
@@ -33,36 +41,17 @@ set.seed(1112)
 
 datasets <- import_data(test = TRUE)
 
-## make dataset names that match merge_data() requirements
 datasets[['swetrau']] <- datasets$swetrau_scrambled
 datasets[['fmp']] <- datasets$fmp_scrambled
 datasets[['atgarder']] <- datasets$atgarder_scrambled
 datasets[['problem']] <- datasets$problem_scrambled
 datasets[['kvalgranskning2014.2017']] <- datasets$kvalgranskning2014.2017_scrambled
 
-## prepare data
-df <- merge_data(datasets)
-pd <- prep_data(df) ## returns a list with the prepared dataset, inclusion counts, and missing parameters
-df <- pd[['df']]
-exclusion <- pd[['exclusion']] ## inclusion/exclusion counts
-na.data <- pd[['na.data']] ## missing parameters
+extract.named(prep_data(datasets)) ## creates the prepared dataset, inclusion counts, and missing parameters
 
-## apply scores
-df$rts <- apply(df, 1, make_rts) ## make_rts.R
-df$triss <- apply(df, 1, make_triss) ## make_triss.R
-df$normit <- apply(df, 1, make_normit) ## make_normit.R
-df$ps <- apply(df, 1, make_ps) ## make_ps.R
-
-## get descriptive stats
 table.one <- table_one(df) ## table_one.R
 dd <- descriptive_data(df) ## descriptive_data.R
-
-## get results
-results <- results(df, bootstrap = TRUE) ## results.R
-t <- results[['t']]
[…]

### 2022-11-29 manuscript.Rmd
results plots

@@
 csl: vancouver.csl
-fontsize: 11pt
+fontsize: 11pts
 linestretch: 1
@@
 
-```{r main, echo=FALSE, message = FALSE}
+```{r main}
   source("main.R") 
@@
 ```{r, echo=FALSE, out.width = "400px"}
-## knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
+knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
 ```
@@
 
-RTS, NORMIT and PS12 were computed according to the original published algorithms [@rtsalgorithm; @normit; @psalgorithm]. TRISS was calculated using the revised 2009 coefficients [@triss2010].
+RTS and PS12 were computed according to the original published algorithms [@rtsalgorithm; @psalgorithm].
+NORMIT was calculated with the revised 2018 algorithm [@skagaValidatingPerformanceTRISS2018].
+TRISS was calculated using the revised 2009 coefficients [@triss2010].
 
@@
 ```{r, echo=FALSE, out.width = "400px"}
-## knitr::include_graphics("images/exclusion.png") ## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
+knitr::include_graphics("images/exclusion.png") ## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
 ```
@@
 `r dd['blunt']`% of cases were due to blunt force trauma.
+The median ISS score was `r dd['iss']` (IQR: `r dd['iss.lo']`-`r dd['iss.hi']`).
 Of cases where 30-day survival was available, there was a `r dd['mort']`% 30-day mortality rate (`r dd['survival.na']` cases with missing survival data).
-Of cases where registered data on the highest level of care was available, `r dd['icu']`% had the ICU as their highest level of care (`r dd['care.na']` cases with missing care level data).
-The median ISS score was `r dd['iss']` (IQR: `r dd['iss.lo']`-`r dd['iss.hi']`).
+`r dd['icu']`% of participants received intensive care as their highest level of care.
 
@@
 
-### TRISS
-
-The odds ratio for TRISS was `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, p-value: `r t[['or.p']]`).
-The ROC is plotted in Figure 3.
-The AUC for the ROC was `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`).
-The integrated calibration index for TRISS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`.
-The accuracy curve is shown in Figure 4.
-The best accuracy was `r t[['acc.max']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`) and was found at the cutoff `r t[['acc.co']]`.
+The fitted curves for the logistic regression models of each score are shown in figure x.
+The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.
 
 ```{r}
-## plot(t[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
+knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
 ```
 
+The receiver operating curves for all scores are plotted in figure x.
+The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.
+
 ```{r}
-## plot(t[['acc']], main="Fig 4. TRISS Accuracy Curve.")
+knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
 ```
 
-### NORMIT 
-The odds ratio for NORMIT was `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, p-value: `r n[['or.p']]`).
-The ROC is plotted in Figure 3.
-The AUC for the ROC was `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`).
-The integrated calibration index for NORMIT was `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]`.
-The accuracy curve is shown in Figure 4.
-The best accuracy was `r n[['acc.max']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and was found at the cutoff `r n[['acc.co']]`.
+The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively.
+The accuracy at all cutoffs is plotted for all scores in figure x.
+The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.
 
-```{r, echo = FALSE}
-## plot(n[['roc']], main="Fig 5. NORMIT Receiver Operating Characteristic Curve.", colorize = TRUE)
+```{r}
+knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
 ```
 
-```{r, echo = FALSE}
-## plot(n[['acc']], main="Fig 6. NORMIT Accuracy Curve.")
-```
-
-### PS
-The odds ratio for PS was `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, p-value: `r p[['or.p']]`).
-The ROC is plotted in Figure 3.
-The AUC for the ROC was `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`).
-The integrated calibration index for PS was `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]`.
-The accuracy curve is shown in Figure 4.
-The best accuracy was `r p[['acc.max']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) and was found at the cutoff `r p[['acc.co']]`.
+The differences between the scores for each result are shown in table x.
 
 ```{r}
-## plot(p[['roc']], main="Fig 6. PS Receiver Operating Characteristic Curve.", colorize = TRUE)
-```
-
-```{r}
-## plot(p[['acc']], main="Fig 7. PS Accuracy Curve.")
-```
-
-### Comparison of scores
-
-Table 3 shows a comparison of ?? for the three models analysed.
-
-```{r}
-  ### table(results, type = "or.plot")
-  ### table(d, type = "deltas")
- ## as_latex(table(d, type = "deltas"))
+ as_latex(table(d, type = "deltas"))
 ```

### 2022-11-29 functions/ici_delta_ci.R

@@ -2,10 +2,10 @@ ici_delta_bs <- function(df, index) {
   
   sample <- df[index,]
   
-  ici1 <- suppressMessages(ici1(sample$score, sample$ofi))
+  ici1 <- suppressMessages(ici(sample$score1, sample$ofi))
   ici1 <- as.numeric(ici1)
   
-  ici2 <- suppressMessages(ici2(sample$score, sample$ofi))
+  ici2 <- suppressMessages(ici(sample$score2, sample$ofi))
   ici2 <- as.numeric(ici2)
   
   delta <- ici1 - ici2

### 2022-11-29 functions/plots.R

@@ -1,0 +1,97 @@
+plot_glm <- function(df, type) {
+  
+  model <- glm(ofi ~ score, family="binomial", data = df)
+  pred <- predict(model, type = "response")
+  pred_df <- data.frame(score = df$score, pred)
+  
+  plot <- ggplot(df, aes(x = score, y = ofi)) +
+    geom_point() +
+    geom_smooth(method = "glm", 
+                method.args = list(family = "binomial"),
+                se = FALSE) +
+    geom_point(data = pred_df, aes(x = score, y = pred), colour = "#5FD4AB") +
+    geom_hline(data = data.frame(c(0.25, 0.50, 0.75)),
+               aes(yintercept = c(0.25, 0.50, 0.75)),
+               colour = "darkgrey", linetype = "dashed") +
+    labs(x = paste0(type), y = "OFI")
+  return(plot)
+}
+
+glm_x3 <- function(df) {
+  df.t <- df[, c("ofi", "triss")] %>% rename(score = triss)
+  df.n <- df[, c("ofi", "normit")] %>% rename(score = normit)
+  df.p <- df[, c("ofi", "ps")] %>% rename(score = ps)
+  
+  t <- plot_glm(df.t, "TRISS")
+  n <- plot_glm(df.n, "NORMIT")
+  p <- plot_glm(df.p, "PS")
+  x3 <- grid.arrange(t, n, p, ncol = 3)
+  x3 <- annotate_figure(x3, top = text_grob("Figure x: Probability of OFI vs probability of survival", size = 12))
+  ggsave(x3, filename = "images/glm.png",height = 667, width = 2001, units = "px")
+}
+
+plot_roc <- function(x, type) {
+  plot <- ggroc(x[['roc']], colour = "#5FD4AB") + ggtitle(paste0(type, " (AUC = ", t[['auc']], ")")) + theme(plot.title = element_text(size=12))
+  return(plot)
+}
+
+roc_x3 <- function(t, n, p) {
+  t.plot <- plot_roc(t, type = "TRISS")
+  n.plot <- plot_roc(n, type = "NORMIT")
+  p.plot <- plot_roc(p, type = "PS")
+  x3 <- grid.arrange(t.plot, n.plot, p.plot, ncol = 3)
+  x3 <- annotate_figure(x3, top = text_grob("Figure x: Receiver Operating Curves", size = 12))
[…]

### 2022-11-29 functions/table.R

@@ -5,28 +5,9 @@ table <- function(data, type = "", t = t, n = n, p = p) {
     names <- c("GCS", "ASA", "RR", "Systolic BP", "Dominant Injury Type", "Age", "ISS", "NISS", "Gender")
     tib <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
     out <- tib %>% gt() %>% tab_header(
-      title = "Table 1. Parameters that were missing in cases excluded for incomplete data"
+      title = "Table 1: Parameters that were missing in cases excluded for incomplete data"
     )
     
-    return(out)
-  }
-  
-  if(type == "or.plot") {
-    nrs <- data.frame(
-      score = c("TRISS, NORMIT, PS"),
-      or = c(data[['t']][['or']], data[['n']][['or']], data[['p']][['or']]),
-      lower = c(data[['t']][['or.ci.lo']], data[['n']][['or.ci.lo']], data[['p']][['or.ci.lo']]),
-      upper = c(data[['t']][['or.ci.hi']], data[['n']][['or.ci.hi']], data[['p']][['or.ci.hi']])
-    )
-    
-    out <- ggplot(nrs, aes(x = factor(score), y = or)) + 
-      geom_point(position=position_dodge(.9)) + 
-      geom_errorbar(aes(ymin = lower, ymax = upper), position=position_dodge(.9), width = .25) + 
-      labs(
-        title = "Figure 8. Comparison of Odds Ratios.",
-        y = "Odds Ratio",
-        x = "Model"
-      )
     return(out)
   }
   
@@ -48,7 +29,7 @@ table <- function(data, type = "", t = t, n = n, p = p) {
       paste0(data[['acc']][['np']]," (", data[['acc']][['np']][[1]], "-", data[['acc']][['np']][[1]], ")")
     )
     
-    tib <- tibble("Scores compared" = comparisons, "AUC difference (95% CI)" = auc, "ICI difference (95% CI)" = ici, "Accuracy difference (95% CI)" = acc)
+    tib <- tibble(" " = comparisons, "AUC difference (95% CI)" = auc, "ICI difference (95% CI)" = ici, "Accuracy difference (95% CI)" = acc)
     out <- tib %>% gt() %>% tab_header(
       title = "Table x. Comparison of model results"
     )
@@ -61,7 +42,7 @@ table <- function(data, type = "", t = t, n = n, p = p) {
[…]

### 2022-12-02 manuscript.Rmd
very very very rough beginning of discussion

@@
 In cases where death occurred within 30 days, at least one OFI was found in `r dd['ofi.mort']`% of cases.
-The median scores for TRISS, NORMIT, and PS were `r dd['t.med']` (IQR: `r dd['t.lo']`-`r dd['t.lo']`), `r dd['n.med']` (IQR: `r dd['n.lo']`-`r dd['n.hi']`), and `r dd['p.med']` (IQR: `r dd['p.lo']`-`r dd['p.hi']`) respectively.
+The median scores for TRISS, NORMIT, and PS were `r dd['t.med']` (IQR: `r dd['t.lo']`-`r dd['t.hi']`), `r dd['n.med']` (IQR: `r dd['n.lo']`-`r dd['n.hi']`), and `r dd['p.med']` (IQR: `r dd['p.lo']`-`r dd['p.hi']`) respectively.
 
@@
 These three models did not differ significantly in their results.
-The main findings indicate that increased probability of survival was associated with lower odds for OFI.
-<!-- OR --> All three models showed poor diagnostic ability in discriminating between OFI and no OFI (AUCs ranged from `r p[['auc']]` to `r t[['auc']]`.
-<!-- AUC -->
+The main findings indicate that increased probability of survival was associated with lower odds for OFI, with odds ratios ranging from `r t['or']` to `r n['or']`.
+The AUC values were almost identical for all three models (range: `r p[['auc']]`-`r t[['auc']]`), indicating poor diagnostic ability in discriminating between OFI and no OFI.
+The integrated calibration index ranged from `r n[['ici']]` to `r p[['ici']]`.
+Accuracy at a cutoff of 0.5 was low across the board (range: `r p[['acc']]`-`r n[['acc']]`).
+
+<!-- 
+NOTES:
+- Maybe repeat why it's useful to do this, like the bigger picture (improve case selection etc)
+- Similar results for all three models could be because:
+-- They all predict the same thing (Ps) and therefore they SHOULD give similar results
+-- They all have quite a few parameters in common
+-- (Less likely) OFI is so uncommon that it's unlikely any model would show a difference
+- Say that it probably wouldn't be useful to investigate more models + motivate
+- My results agree w/ previous studies on preventable deaths
+- Trauma models useful on populations but not individuals
+- OR - can it be that low severity cases (high Ps) are more likely to be ignored/non-selected and therefore skew results? -> Which would affect the association
+- Makes sense that models designed to predict death can't predict OFI - not what they are made for
+- If a hospital has a high case volume -> more practice -> less ofi -> harder to find predictors of ofi mb??
+- Health equity:
+-- "Deaths in elderly less likely to be found preventable" 
+-- Which is reasonable but maybe not 100% due to scientificness - element of bias?
+- My population is representative
+-- Normal to be majority men
+
+-->
 
@@
 -----------
-<!-- 
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias -->
 
-One of the strengths of this study is the duration of enrollment which made it possible to have a large study population.
-Karolinska University Hospital is the primary trauma centre for a catchment population of 3 million, which means that this hospital-based study was in effect a population-based study, which reduces the risk of selection bias.
+One of the strengths of this study is the duration of enrolment and Karolinska University Hospital's catchment population of 3 million, which allowed for a large study population.
+The outcome data is strengthened by Karolinska University Hospital's extensive quality review process involving personnel from all concerned specialities and professions.
 All patients that met inclusion criteria were included in the study.
-The code for data analysis was written by a single student researcher whilst blinded to the real-world data to minimise the risk of researcher bias.
 
-<!-- something about how staff at confs may have differed over the years -->
+There were several limitations to this study.
+One of them is that selection of cases for review at conferences was in large part based upon the use of audit filters.
+There is as yet no widespread consensus on which audit filters are the most effective at identifying cases and this is an ongoing topic of research.
+The choice of audit filters used affects case selection and in turn introduces an element of selection bias into this study.
+It is possible, and perhaps even likely, that some cases with opportunities for improvement were never selected for case review and therefore weren't available for inclusion in this study.
+Another limitation is that, due to the long duration of enrolment in this study, the composition of the staff present at conferences will have differed over time.
+Determining whether OFI exists or not is an essentially subjective evaluation and the conclusions reached by any given conference may vary depending on which individuals are present.
+It is also possible that standards and expectations of care have changed over time as knowledge and technology improves, which would also in turn affect assignment of OFI.
 
-<!-- 
+<!-- In this study the presence of OFI was a relatively uncommon outcome (`r dd['ofi.pc']`% of cases). 
+dont remember why i wrote this -->
+
+Interpretations
+-----------
+The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths <!-- cite here -->.
+Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e.
+whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study.
+Ghorbani et al found that ____.
+Brohi 2016 found that ____.
+Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI.
+The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.
+
+Our results do show an association between lower probability of survival and higher odds for assignment of OFI.
+Severe trauma is much more complex and time-dependent than mild trauma.
+Complexity and time pressure are two factors that increase the risk of management errors, so the association found is unsurprising.
+However, this result should be evaluated in the context of potential selection bias, as described above.
+Cases with low trauma severity are less likely to trigger audit filters and are in turn less likely to be reviewed at M&M conferences, which will lead to severe trauma being over-represented in OFI assignment.
+
+<!-- why important -->  
+<!-- similar -->
+
+Clinical applications 
+-----------
+
+### Generalisability 
+This study is based on a patient population at a level 1 trauma centre in Stockholm, Sweden and the results are likely applicable for other level 1 trauma centres in the Nordic countries, as these countries share similarities in patient demographics, comorbidities, and injury types.
+As mentioned previously, the Norwegian model NORMIT was found to be superior to the American model TRISS at predicting mortality at this hospital <!-- ghorbani -->.
+This study has less external validity with regards to international populations that differ significantly from the one in this study, for example populations with a higher proportion of penetrating trauma.
+
+### Health equity
+
+## Future research
+
+Future studies should seek to evaluate the predictive power of other audit filters that are currently used for case selection.
+
 Interpretation
@@
 
+<!--
 Give a cautious overall interpretation of results considering
@@
 
+- makes sense that models that predict death cant predict ofi very well
+- more severe trauma is associated w higher ofi but the scores cant predict ofi 
+- -> that 
+--> 
 Generalisability
@@
 
-Discuss the generalisability (external validity) of the study results
-
--->
+<!-- Relevant for KUH and other major scandinavian level 1 trauma centers, potentially 
+relevant for other hosps w similar pt populations + injury panorama-->
 

### 2022-12-02 functions/plots.R

@@ -31,7 +31,7 @@ glm_x3 <- function(df) {
 }
 
 plot_roc <- function(x, type) {
-  plot <- ggroc(x[['roc']], colour = "#5FD4AB") + ggtitle(paste0(type, " (AUC = ", t[['auc']], ")")) + theme(plot.title = element_text(size=12))
+  plot <- ggroc(x[['roc']], colour = "#5FD4AB") + ggtitle(paste0(type, " (AUC = ", x[['auc']], ")")) + theme(plot.title = element_text(size=12))
   return(plot)
 }
 

### 2022-12-02 main.R

@@ -23,7 +23,7 @@ library(kableExtra)
 library(DiagrammeRsvg)
 library(rsvg)
 library(mvbutils)
-library("gridExtra")
+library(gridExtra)
 library(ggpubr)
 
 noacsr::source_all_functions()
@@ -37,6 +37,13 @@ extract.named(prep_data(datasets)) ## creates the prepared dataset, inclusion co
 table.one <- table_one(df) ## table_one.R
 dd <- descriptive_data(df) ## descriptive_data.R
 extract.named(results(df, bootstrap = TRUE, boot.no = 10)) ## results.R
+
+## make figures
+##roc_x3(t, n, p)
+##acc_x3(t, n, p)
+##glm_x3(df)
+##save_png(exclusion_flowchart(exclusion), "exclusion")
+##save_png(ofi_diagram(), "ofi")
 
 ## done
 end.time <- Sys.time()

### 2022-12-02 manuscript.Rmd
generalisability + health equity

@@
 All patients that met inclusion criteria were included in the study.
+Data collection, patient inclusion and data analysis was prepared on scrambled data by a single student research before being given access to the real data, in order to minimise researcher bias.
 
@@
 
-Clinical applications 
------------
-
-### Generalisability 
+## Generalisability 
 This study is based on a patient population at a level 1 trauma centre in Stockholm, Sweden and the results are likely applicable for other level 1 trauma centres in the Nordic countries, as these countries share similarities in patient demographics, comorbidities, and injury types.
@@
 ### Health equity
+The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide.
+`r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally <!-- find a cite even tho i know this is true -->.
+The predominance of male patients invites one to question if female patients are less likely to receive quality care.
+In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected.
+This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes.
+None of these studies found a correlation beween sex and increased mortality or morbidity <!-- maybe write: or other serious outcomes --> <!-- also add cites -->, and in fact one study in the Netherlands found that female sex was correlated with decreased mortality <!-- cite -->.
+
+This study has not stratified for sex in the statistical analyses, and therefore cannot comment on any relation between sex and OFI at this centre.
+One of the trauma severity models we investigated, PS, does include patient sex as a coefficient in predicting mortality, but we did not examine whether patient sex affects the ability of PS to predict OFI.
+However, there is another currently ongoing project at this hospital that compares OFI between the male and female sex cohorts.
+The uneven distribution of sex in this study is not a limitation of this study in terms of external validation and generalisability since our demographics match the distribution in the Swedish trauma population.
+<!-- probably rephrase since htis is p similar to what elias said -->
+
+Another aspect of health equity is patient age.
+We know that peer review panels are more likely to classify trauma deaths as non-preventable in the geriatric population <!-- Radke 2019, Radke didnt cite his source tho -->.
+This is unlikely to be purely due to discrimination, since several factors make the elderly less likely to survive severe trauma, including a higher prevalence of comorbidities and in general lower margins for tolerating extreme physiological stress <!-- maybe some cites about response to injury in elderly -->.
+In fact, all three trauma severity models investigated in this study include coefficients for age, since age is a well-known predictor of mortality.
+Nevertheless, there is a possibility that healthcare providers are less likely to look for errors in older patients due to a predisposition to finding older patients less likely to be savable.
+The question of whether age is correlated with OFI was outside the scope of this study, but as with sex, is being investigated in other currently ongoing projects at this institution.
+The geriatric population differs from the average adult in terms of injury types, injury response, and management.
+This, in combination with worse trauma outcomes in the geriatric population, should prompt focus on this cohort in future research since this group has perhaps the most to gain from improvements in trauma care, especially improvements that recognise the unique challenges in this population and aim to specifically tailor their care to those challenges.
 
@@
 
+Contributions
+==========
+The code for statistical analysis and the writing of this paper has been carried out by myself, although of course with advice from my supervisors. 
+<!-- 
+I know it maybe seems weird to have this section, but the instructions specifically say: 
+Här anger du vilka delar och aspekter av arbetet som du personligen har utfört, och vilka som
+har gjorts av andra, t ex mer avancerad statistik, kemiska specialanalyser. Om du har
+samarbetat med en kurskamrat inom ett större projekt behöver du tydligt ange vilka delar man
+genomfört/skrivit tillsammans och vilka delar har utförts individuellt. Ange också namn på
+personen/ -er på din kurs du samarbetat med. Har du gjort det mesta själv anger du det.
+
+The last sentence makes it seem like I have to include this -->
+
 References

### 2022-12-02 manuscript.Rmd
future research

@@
 Future studies should seek to evaluate the predictive power of other audit filters that are currently used for case selection.
+As mentioned previously, trauma centres today use a large variety of audit filters and many of these have not yet been supported by research.
+Future studies could use logistic regression to identify factors associated with OFI, and investigate if any of these factors have predictive power.
+The ultimate goal of such a study study, as with this study, would be to improve case selection for review in conferences.
+As these conferences are the gold standard of trauma improvement programmes, improved case selection has the potential to directly improve trauma care, in addition to reducing the resource burden of identifying cases and holding said conferences.
+
+In addition, future studies should aim to confirm the findings of this study within a different population.
+The population in this study is a typically Scandinavian one, with a low incidence of penetrating trauma.
+As mentioned previously, the external validity of trauma severity models is in some degree dependent on the similarity of the populations they were developed from.
+Ideally a new study would investigate the predictive power of the models used in this study, in a country with a higher incidence of penetrating trauma and potentially other different population characteristics, in order to confirm the results found in this study.
+Such a study could use a study design similar to this one, i.e.
+use logistic regression and measures of discrimination and calibration.
 

### 2022-12-08 manuscript.Rmd
discussion

@@
 output:
-  pdf_document: default
+  word_document: default
   html_document:
@@
     number_sections: true
-  word_document: default
+  pdf_document: default
 bibliography: bibliography.bib
@@
 
-This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI.
-These three models did not differ significantly in their results.
+This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences.
+The three models investigated did not differ significantly in their results.
 The main findings indicate that increased probability of survival was associated with lower odds for OFI, with odds ratios ranging from `r t['or']` to `r n['or']`.
 The AUC values were almost identical for all three models (range: `r p[['auc']]`-`r t[['auc']]`), indicating poor diagnostic ability in discriminating between OFI and no OFI.
-The integrated calibration index ranged from `r n[['ici']]` to `r p[['ici']]`.
+Calibration was also low for all models (ICI range: `r n[['ici']]` - `r p[['ici']]`).
 Accuracy at a cutoff of 0.5 was low across the board (range: `r p[['acc']]`-`r n[['acc']]`).
-
-<!-- 
-NOTES:
-- Maybe repeat why it's useful to do this, like the bigger picture (improve case selection etc)
-- Similar results for all three models could be because:
--- They all predict the same thing (Ps) and therefore they SHOULD give similar results
--- They all have quite a few parameters in common
--- (Less likely) OFI is so uncommon that it's unlikely any model would show a difference
-- Say that it probably wouldn't be useful to investigate more models + motivate
-- My results agree w/ previous studies on preventable deaths
-- Trauma models useful on populations but not individuals
-- OR - can it be that low severity cases (high Ps) are more likely to be ignored/non-selected and therefore skew results? -> Which would affect the association
-- Makes sense that models designed to predict death can't predict OFI - not what they are made for
-- If a hospital has a high case volume -> more practice -> less ofi -> harder to find predictors of ofi mb??
-- Health equity:
--- "Deaths in elderly less likely to be found preventable" 
--- Which is reasonable but maybe not 100% due to scientificness - element of bias?
-- My population is representative
--- Normal to be majority men
-
--->
 
@@
 
-<!-- In this study the presence of OFI was a relatively uncommon outcome (`r dd['ofi.pc']`% of cases). 
-dont remember why i wrote this -->
+Interpretation
+-----------
 
-Interpretations
------------
-The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths <!-- cite here -->.
+Our results show an association between lower probability of survival and higher odds for assignment of OFI.
+This result should be evaluated in the context of potential selection bias, as described above.
+Cases with low trauma severity are less likely to trigger audit filters and are in turn less likely to be reviewed at M&M conferences, which will lead to severe trauma being over-represented in OFI assignment.
+However, it is unlikely that this selection bias is the main cause of the association found.
+Severe trauma is much more complex and time-dependent than mild trauma.
+Complexity and time pressure are two factors that increase the risk of management errors, and management errors have been found to be the main cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+In this context, the association between trauma severity and OFI has an obvious causality and our findings on this association are unsurprising.
+
+Despite this association, none of the trauma severity models we investigated are able to reliably predict OFI.
+The area under the curve of the receiver operating curves - a value that describes the ability of a model to discriminate between two outcomes - was around 0.65 for all three scores, which indicates that these models were not able to reliably predict OFI.
+We also showed that the models were poorly calibrated as to the outcomes.
+Intuitively, this makes sense, as these models are designed to predict death and not errors in care.
+These models do not take into account factors that might lead to suboptimal care - they are based on physiological and anatomical parameters that quantify the severity of an injury.
+Even though a higher trauma severity is associated with greater odds for OFI, these models cannot discriminate between a high trauma severity with OFI and one that has no OFI.
+
+We showed that all three models have a low accuracy at a 0.5 cutoff.
+One reason for the low accuracy is the low incidence of OFI, which means that maximal accuracy is found when assigning "no OFI" to all cases, although this would obviously lead to false negatives.
+Either way, whether you use maximum accuracy or accuracy at a 0.5 cutoff as a metric, none of these models have significant accuracy.
+In real-world terms, there is no decision-making cutoff for predicted survival that would lead to an improvement in case selection - these approaches would lead to either a high number of false positives or false negatives.
+
+These three models had essentially the same results, which is an expected outcome considering that they are all tools for estimating the same thing - the probability of survival.
+In addition, the models have a great degree of overlap in the parameters they incorporate, including GCS and age.
+The models we analysed are commonly-used and are sourced from three different geographical populations.
+The fact that they all had similar performance is a sign that investigation of other models is unlikely to lead to different results.
+
+The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths.
 Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e.
 whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study.
-Ghorbani et al found that ____.
-Brohi 2016 found that ____.
+Ghorbani et al found that both TRISS and NORMIT had poor performance in predicting preventable deaths [@prevdeathsghorbani].
+Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16].
 Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI.
@@
 
-Our results do show an association between lower probability of survival and higher odds for assignment of OFI.
-Severe trauma is much more complex and time-dependent than mild trauma.
-Complexity and time pressure are two factors that increase the risk of management errors, so the association found is unsurprising.
-However, this result should be evaluated in the context of potential selection bias, as described above.
-Cases with low trauma severity are less likely to trigger audit filters and are in turn less likely to be reviewed at M&M conferences, which will lead to severe trauma being over-represented in OFI assignment.
-
-<!-- why important -->  
-<!-- similar -->
+Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions.
+Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful.
+Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed.
+Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity.
+They are much less useful when applied to individuals, and were not intended for that use.
 
@@
 This study is based on a patient population at a level 1 trauma centre in Stockholm, Sweden and the results are likely applicable for other level 1 trauma centres in the Nordic countries, as these countries share similarities in patient demographics, comorbidities, and injury types.
-As mentioned previously, the Norwegian model NORMIT was found to be superior to the American model TRISS at predicting mortality at this hospital <!-- ghorbani -->.
-This study has less external validity with regards to international populations that differ significantly from the one in this study, for example populations with a higher proportion of penetrating trauma.
+This study is less generalisable for smaller trauma centres as those centres tend to manage less severe trauma and will have a different patient population.
+This study also has less external validity with regards to international populations that differ significantly from the one in this study, for example populations with a higher proportion of penetrating trauma.
 
@@
 The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide.
-`r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally <!-- find a cite even tho i know this is true -->.
+`r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden].
 The predominance of male patients invites one to question if female patients are less likely to receive quality care.
@@
 This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes.
-None of these studies found a correlation beween sex and increased mortality or morbidity <!-- maybe write: or other serious outcomes --> <!-- also add cites -->, and in fact one study in the Netherlands found that female sex was correlated with decreased mortality <!-- cite -->.
+None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pepe2019; @trentzsch2004], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].
 
-This study has not stratified for sex in the statistical analyses, and therefore cannot comment on any relation between sex and OFI at this centre.
+This study has not compared differences between the sexes, and therefore cannot comment on any relation between sex and OFI at this centre.
 One of the trauma severity models we investigated, PS, does include patient sex as a coefficient in predicting mortality, but we did not examine whether patient sex affects the ability of PS to predict OFI.
 However, there is another currently ongoing project at this hospital that compares OFI between the male and female sex cohorts.
-The uneven distribution of sex in this study is not a limitation of this study in terms of external validation and generalisability since our demographics match the distribution in the Swedish trauma population.
-<!-- probably rephrase since htis is p similar to what elias said -->
+The uneven distribution of sex in this study does not make for a limitation of this study in terms of generalisability since our demographics match the distribution of sex in trauma cases both globally and in the Swedish trauma population.
 
 Another aspect of health equity is patient age.
-We know that peer review panels are more likely to classify trauma deaths as non-preventable in the geriatric population <!-- Radke 2019, Radke didnt cite his source tho -->.
-This is unlikely to be purely due to discrimination, since several factors make the elderly less likely to survive severe trauma, including a higher prevalence of comorbidities and in general lower margins for tolerating extreme physiological stress <!-- maybe some cites about response to injury in elderly -->.
+We know that peer review panels are more likely to classify trauma deaths as non-preventable in the geriatric population [@radke2019].
+This is unlikely to be purely due to discrimination, since several factors make the elderly less likely to survive severe trauma, including a higher prevalence of comorbidities and in general lower margins for tolerating extreme physiological stress [@bardes2019].
 In fact, all three trauma severity models investigated in this study include coefficients for age, since age is a well-known predictor of mortality.
@@
 
-Interpretation
---------------
-
-<!--
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from
-similar studies, and other relevant evidence
-
-- makes sense that models that predict death cant predict ofi very well
-- more severe trauma is associated w higher ofi but the scores cant predict ofi 
-- -> that 
---> 
-Generalisability
-----------------
-
-<!-- Relevant for KUH and other major scandinavian level 1 trauma centers, potentially 
-relevant for other hosps w similar pt populations + injury panorama-->
-
 Conclusion
@@
 
-<!-- Give a short conclusion, corresponding to the aim -->
-
-Contributions
-==========
-The code for statistical analysis and the writing of this paper has been carried out by myself, although of course with advice from my supervisors. 
-<!-- 
-I know it maybe seems weird to have this section, but the instructions specifically say: 
-Här anger du vilka delar och aspekter av arbetet som du personligen har utfört, och vilka som
-har gjorts av andra, t ex mer avancerad statistik, kemiska specialanalyser. Om du har
-samarbetat med en kurskamrat inom ett större projekt behöver du tydligt ange vilka delar man
-genomfört/skrivit tillsammans och vilka delar har utförts individuellt. Ange också namn på
-personen/ -er på din kurs du samarbetat med. Har du gjort det mesta själv anger du det.
-
-The last sentence makes it seem like I have to include this -->
+This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, opportunities for improvement, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences.
+We showed that, while low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI.
+The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.
 

### 2022-12-09 manuscript.Rmd
readability

@@
 output:
-  word_document: default
+  pdf_document: default
   html_document:
@@
     number_sections: true
-  pdf_document: default
+  word_document: default
 bibliography: bibliography.bib
@@
 ============
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting around 9% of global deaths annually [@globalburden].
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden].
 The global health burden of trauma has led to it being described as a worldwide pandemic.
@@
 Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
-The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg].
-This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 [@sdg20].
-Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.
+The quality of trauma care is therefore a globally relevant issue with significance for a great number of people.
+Improvements in trauma outcomes benefit not only the individual, but also society at large by reducing the societal and healthcare costs of disability.
+The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg].
+The importance of improving trauma care named as one of the strategies in achieving this goal.
+The goal was not met by 2020 [@sdg20], which is yet another sign that there is significant progress to be made in trauma care.
 
+Efforts to advance the state of trauma care have been ongoing for many years.
+Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi].
+The most well-known example of a QI program is the Trauma Quality Improvement Program established by the American College of Surgeons in 2008 <!-- cite georg [@Hemmila2010-sr] -->.
+QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference.
+These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case.
+One method of improving the quality of M&M conferences is by improving case selection.
+This is usually achieved by using audit filters that automatically flag cases for review.
+One approach that is becoming more common is the use of trauma severity models as audit filters.
+Trauma severity models are algorithms that evaluate trauma severity in quantitative terms.
+The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper.
+
+<!-- georg mock2009
+earlier research focus on skill
+80s focus shifted
+2008 TQIP
+-->
+
+## Morbidity and mortality conferences
+
+The significance of M&M conferences in trauma QI is a consequence of the unique challenges that trauma poses.
 Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
-Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
-Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi].
-Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath].
-However, the implementation of QI programs differs widely between clinics [@tqipanglo], with smaller, non-specialised hospitals tending to have less organised programs [@whoguidelines].
+Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care.
+The only way to achieve this is through peer review and this is why M&M conferences are the cornerstone of trauma QI programs.
+
+As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care.
+Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement.
+For best results, conferences should be attended by all involved specialties and have a structured format.
+The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines].
+Despite theserecommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
+The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+
+The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
+For example, imagine a conference where every case reviewed was found to have no OFI.
+In such a situation the value of the conference would effectively be zero, as nothing would be learned from it.
+In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed.
+This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
+Some clinics have made further advances in standardising case election; most commonly by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
+Examples of audit filters include time to first medical contact, adverse event rate, and, in some clinics, trauma severity models [@developqi; @whoguidelines].
+
+The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
+The latter benefit is not insignificant considering the large caseload found in many urban trauma centres.
+The process of identifying cases is time-consuming and trauma centres that lack resources may not be able to allocate enough staff to case selection, which in turn lowers the effectiveness of M&M review.
+Despite the benefits of standardised case selection, several studies have shown that the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
+<!-- in specific areas?
+is it worse in poor places?
+-->
 
@@
 Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury.
-An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma.
-This question has spurred the development of multiple trauma severity models.
-Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality.
+One of several challenges posed by this heterogeneity is difficulty in expressing the severity of an injury in quantitative terms.
+For example, it would be difficult to intuitively say who was the more severely injured out of a patient with multiple gun shot wounds and a patient who was injured in a high speed road traffic accident.
+The ability to estimate trauma severity quantitatively is crucial for trauma QI programs as it allows trauma centres to assess outcomes and clinic performance.
+
+Trauma severity models are algorithms that seek to address this need.
+These models translate anatomical and physiological parameters into a quantitive value that represents the totality of the injury to any given patient.
 Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview].
-Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival.
+Today there are hundreds of trauma severity models that differ in methodology but have the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival.
 
 Trauma severity models have become a cornerstone of modern trauma QI programs [@historytss].
+These models allow clinics to compare their outcomes against an established benchmark.
 One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high predicted probability of survival.
 [@whoguidelines].
-Trauma severity models also allow clinics to compare their outcomes against an established benchmark [@whoguidelines].
-However, an inherent weakness of any trauma severity model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
-Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality].
-This weakness is one reason why models continue to be developed and improved on.
+They also provide a common language for expressing trauma severity, which enables clinics to compare their performance with other, comparable clinics [@whoguidelines].
+
+An inherent weakness of any trauma severity model is that estimates may be unreliable when applied to a population different from the one the model was developed from.
+This limitation in generalisability is a consequence of the heterogeneity of trauma, as the spectrum of traumatic injury varies across geographical regions [@assessingquality].
+This weakness means that there is no single "best" model, as the performance of any model will be best in the population the model was sourced from.
+This is one reason why new models continue to be developed.
 Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.
 
-The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter].
+The most commonly used model worldwide is The Trauma and Injury Severity Score (TRISS) [@trissbetter], first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010].
 TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss].
-The RTS requires GCS on arrival at hospital, which cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained.
+One weakness of TRISS is that the RTS requires GCS on arrival at hospital, which cannot be estimated in situations where the patient has already been intubated.
 This gives TRISS limited value in estimating probability of survival in certain populations [@trissbetter].
@@
 
-## Morbidity and mortality conferences
+<!-- how do they compare -->
 
-The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.
-M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
-The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care.
-The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines].
-Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
-The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
+## Knowledge gap
 
-The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review.
-In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed.
-This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo].
-The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
-The latter benefit is not insignificant considering the large caseload found in many urban trauma centres.
-Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
-
-The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
-Some clinics have already made efforts to standardise case election; for example, by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
-Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi].
-
-## Trauma severity models as quality indicators
-
-Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring.
-Some hospitals use trauma severity models as audit filters is to only review deaths in cases that had a high predicted probability of survival, the idea being that deaths in cases with high predicted survival are more likely to have been preventable deaths.
-Research to date does not support this use case.
+As previously mentioned, some trauma centres use trauma severity models as audit filters in selecting cases for M&M conferences [@whoguidelines].
+One example of their use as audit filters is in only reviewing deaths in cases that had a high predicted probability of survival.
+This strategy is based on the idea that deaths in cases with high predicted survival are more likely to have been preventable deaths.
+The factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e.
+whether there was a significant OFI.
+The use of trauma severity models as audit filters for preventable deaths is therefore based upon the expectation that these models can predict OFI in mortalities.
+Research to date does not support that trauma severity models can predict OFI in deaths.
 Two major studies have looked at trauma severity models and preventable deaths.
 One study in Sweden investigated TRISS and NORMIT [@prevdeathsghorbani] and a study in the UK investigated TRISS and PS [@brohi16].
-Both of these studies showed that the studied models were unable to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani; @brohi16].
+Both of these studies showed that these studied models were unable to reliably discriminate between preventable and non-preventable deaths and applying cutoffs led to some preventable deaths being missed [@prevdeathsghorbani; @brohi16].
 
+The use of trauma severity models as audit filters is not limited to solely mortalities; some hospitals also use them as audit filters in survivors.
 However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
-The two studies mentioned included only deaths and not other morbidities and thereby had low sample sizes.
-To date there is no published research on the ability of trauma severity models to predict OFI across the board and not just in mortalities.
-If it could be found that an existing trauma severity model has value in predicting OFI, it would confirm the practice of using models as audit filters and would improve case selection and reduce resource burden.
-Conversely, if trauma severity models cannot reliably predict OFI, it is possible that using these models as audit filters causes cases to be unnecessarily flagged for review, which in turn wastes hospital resources.
-Either way, the answer to this question has the potential to benefit trauma QI programmes.
+The two studies mentioned investigated only preventable deaths and to date there is no published research that investigates whether trauma severity models can predict OFI across the board.
+
+There is value in determining the effectiveness of trauma severity models in predicting OFI.
+If they can reliably predict OFI, their use as audit filters would be supported as they would improve case selection for M&M conferences.
+More accurate case selection will lead to M&M conferences becoming more effective at discovering flaws in care.
+There is also a secondary benefit in that effective audit filters allow case selection to become more automated, which reduces the large resource burden of identifying cases for review.
+Conversely, if trauma severity models cannot reliably predict OFI, we would know that their use as audit filters is harmful to the M&M process.
+Audit filters that are poor predictors can be harmful in two ways.
+On the one hand, they can over-predict OFI, causing more cases to be reviewed unnecessarily.
+This reduces the effectiveness of M&M conferences.
+On the other hand, they can miss cases with OFI, leading to flaws being undetected by M&M conferences.
+Given that some hospitals already use trauma severity models as audit filters, it is prudent to ascertain whether this is a helpful or harmful feature in the trauma QI process.
 

### 2022-12-09 manuscript.Rmd
proofread method

@@
 The quality of trauma care is therefore a globally relevant issue with significance for a great number of people.
-Improvements in trauma outcomes benefit not only the individual, but also society at large by reducing the societal and healthcare costs of disability.
+Improvements in trauma outcomes benefit both the individual and also society at large by reducing the societal and healthcare costs of disability.
 The significance of the consequences and burden of trauma has been recognised by the UN, who in their Sustainable Development Goals included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg].
 The importance of improving trauma care named as one of the strategies in achieving this goal.
-The goal was not met by 2020 [@sdg20], which is yet another sign that there is significant progress to be made in trauma care.
+The goal was not met by 2020 [@sdg20], which is another signal that there is significant progress to be made in trauma care.
 
@@
 Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi].
-The most well-known example of a QI program is the Trauma Quality Improvement Program established by the American College of Surgeons in 2008 <!-- cite georg [@Hemmila2010-sr] -->.
 QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference.
@@
 This is usually achieved by using audit filters that automatically flag cases for review.
-One approach that is becoming more common is the use of trauma severity models as audit filters.
+One approach is the use of trauma severity models as audit filters.
 Trauma severity models are algorithms that evaluate trauma severity in quantitative terms.
@@
 
-<!-- georg mock2009
-earlier research focus on skill
-80s focus shifted
-2008 TQIP
--->
-
 ## Morbidity and mortality conferences
 
-The significance of M&M conferences in trauma QI is a consequence of the unique challenges that trauma poses.
-Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers.
+The significance of M&M conferences in trauma QI is a consequence of the unique challenges inherent to trauma care.
+Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines.
 Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
 Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care.
-The only way to achieve this is through peer review and this is why M&M conferences are the cornerstone of trauma QI programs.
+This can only be achieved through peer review and consequently, M&M conferences are the cornerstone of trauma QI programs.
 
@@
 Regular conferences allow QI programs to identify patterns of errors and provide a basis for systemic quality improvement.
-For best results, conferences should be attended by all involved specialties and have a structured format.
+To maximise effectiveness, conferences should be attended by all involved specialities and have a structured format.
 The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines].
@@
 For example, imagine a conference where every case reviewed was found to have no OFI.
-In such a situation the value of the conference would effectively be zero, as nothing would be learned from it.
+The value of that conference would effectively be zero, as it would fail to discover any flaws in trauma care.
 In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal as it cannot be known whether a case was mismanaged prior to it being reviewed.
@@
 The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].
-Some clinics have made further advances in standardising case election; most commonly by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
+Some clinics have made further advances in standardising case election; most commonly by using quality indicators, or audit filters, to automatically flag cases for inclusion in M&M conferences [@tqipanglo; @developqi].
 Examples of audit filters include time to first medical contact, adverse event rate, and, in some clinics, trauma severity models [@developqi; @whoguidelines].
 
-The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
+The benefit of standardising and automating case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours.
 The latter benefit is not insignificant considering the large caseload found in many urban trauma centres.
@@
 Despite the benefits of standardised case selection, several studies have shown that the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo].
-<!-- in specific areas?
-is it worse in poor places?
--->
 
@@
 One of several challenges posed by this heterogeneity is difficulty in expressing the severity of an injury in quantitative terms.
-For example, it would be difficult to intuitively say who was the more severely injured out of a patient with multiple gun shot wounds and a patient who was injured in a high speed road traffic accident.
+For example, it is difficult to intuitively say who was the more severely injured out of a patient with multiple gun shot wounds and a patient who was injured in a high speed road traffic accident.
 The ability to estimate trauma severity quantitatively is crucial for trauma QI programs as it allows trauma centres to assess outcomes and clinic performance.
@@
 This is one reason why new models continue to be developed.
-Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.
+Another reason is that efforts to establish which risk factors best predict mortality are still ongoing.
 
@@
 NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when used at the national level [@validationnormit].
-
-<!-- how do they compare -->
 
@@
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
+These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT and PS and OFI using logistic regression.
 The predictive performance of these models was evaluated using measures of discrimination and calibration.
@@
 
-<!-- Describe the setting, locations, and relevant dates, including
-periods of recruitment, exposure, follow-up, and data collection -->
-
 Karolinska University Hospital is located in Solna, Sweden.
@@
 Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].
-The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
+
+The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and follows SweTrau guidelines.
 All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre.
@@
 The Integrated Calibration Index (ICI) was used to measure calibration.
-Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff.
-Results werepresented with a confidence interval of 95%.
+Accuracy was plotted and is reported as the accuracy at a 0.5 cutoff.
+Confidence intervals were calculated with bootstrapping.
+Results were presented with a confidence interval of 95%.
 A p-value of <0.05 was considered significant.

### 2022-12-09 manuscript.Rmd
proofread method

@@
 
-The fitted curves for the logistic regression models of each score are shown in figure x.
+The fitted curves for the logistic regression models of each score are shown in figure 3.
 The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.
@@
 
-The receiver operating curves for all scores are plotted in figure x.
+The receiver operating curves for all scores are plotted in 4.
 The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.
@@
 The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively.
-The accuracy at all cutoffs is plotted for all scores in figure x.
+The accuracy at all cutoffs is plotted for all models in  5.
 The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.
@@
 Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16].
-Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI.
+Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI across the board and not only in mortalities.
 The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.
 
-Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions.
+Even though our results were expected, it was still important to investigate this topic scientifically because trauma severity models are in use today as audit filters.
 Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful.
@@
 This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes.
-None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pepe2019; @trentzsch2004], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].
+None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].
 

### 2022-12-14 manuscript.Rmd

--- 2022-12-09 manuscript.Rmd
@@
 
+<!-- how do they compare -->
+
 ## Knowledge gap
@@
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT and PS and OFI using logistic regression.
+These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression.
 The predictive performance of these models was evaluated using measures of discrimination and calibration.
@@
 
+<!-- Describe the setting, locations, and relevant dates, including
+periods of recruitment, exposure, follow-up, and data collection -->
+
 Karolinska University Hospital is located in Solna, Sweden.
@@
 Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].
-
-The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and follows SweTrau guidelines.
+The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
 All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre.
@@
 The Integrated Calibration Index (ICI) was used to measure calibration.
-Accuracy was plotted and is reported as the accuracy at a 0.5 cutoff.
-Confidence intervals were calculated with bootstrapping.
-Results were presented with a confidence interval of 95%.
+Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff.
+Results werepresented with a confidence interval of 95%.
 A p-value of <0.05 was considered significant.

### 2022-12-15 manuscript.Rmd

@@
 <!-- Based on the dates I see in the Swetrau dataset -->
+<!--JA: Det har samlats ännu äldre data i olika former.
+Skulle bara skippa årtalen kring traumaregistret.
+KS har kört någon form av MoM sedan 2012.
+Se: https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/24625137/ -->
 
@@
 The database contains data on patients from 2014 to 2021.
-<!-- Double check these dates -->
+<!-- Double check these dates.
+JA: Se ovan, registrerat från 2012 kring dödsdall men tror morbiditet började 2014. Oavsett så är VÅR data från 2014-2021 -->
 
@@
 
+<!--JA: Jag tror du även kan använda denna länk om du inte vill ladda ner fil:
+https://rcsyd.se/swetrau/wp-content/uploads/sites/10/2019/09/Patientinformation-SweTrau.pdf
+-->
+
 This project is a retrospective cohort study and the project has thereby had no effect on patient treatment.
@@
 ------------
-
+<!--Figure X??? -->
 The fitted curves for the logistic regression models of each score are shown in figure x.
@@
 ```
-
+<!--Figure X??? -->
 The receiver operating curves for all scores are plotted in figure x.
@@
 ```
-
+<!--Figure X??? -->
 The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively.
@@
 ```
-
+<!--Table X? -->
 The differences between the scores for each result are shown in table x.
@@
 It is also possible that standards and expectations of care have changed over time as knowledge and technology improves, which would also in turn affect assignment of OFI.
+<!--JA: I takt med att man åtgärdar systematiskta fel så upkommer na -->
 

### 2022-12-15 manuscript.Rmd
.

--- 2022-12-09 manuscript.Rmd
@@
 knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
+## titles for plots will be added in word
 ```
@@
 ```{r} 
-  table.one ## will be improved later
+  table.one
 ```

### 2022-12-15 manuscript.Rmd
Comment on project paper and table function

--- 2022-12-15 manuscript.Rmd
@@
 ---
+## To use of cross-references and avoid having to manually number figures and tables 
+## you need to switch to the output format bookdown::pdf_document2 and 
+## bookdown::word_document2, see https://bookdown.org/yihui/rmarkdown-cookbook/cross-ref.html
 title: 'Association between trauma severity models and opportunities for improvement:
@@
 ============
+<!-- I think that it often makes sense to keep paragraphs short and focus on one key message per paragraph.
+In this first paragraph you present multiple important points, the burden of fatal trauma, the burden of non-fatal trauma, and the importance of improving trauma care.
+Each of these messages can have their own, shorter paragraphs, which in my opinion makes any text easier to digest.-->
 Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide and accounts for approximately 9% of global deaths annually [@globalburden].
@@
 
+<!-- See my comment above.
+You introduce several important concepts here, and one that is key to your paper, the use of trauma severity models for case selection.
+By introducing it late in the paragraph, you effectively hide it, and its importance is kind of lost on the reader.
+I suggest that you split this paragraph in two, and devote the second one to the use of trauma severity models-->
 Efforts to advance the state of trauma care have been ongoing for many years.
@@
 The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper.
+<!-- This last sentence is perhaps the most important sentence in the introduction, but that is not very clear from how it is presented.
+See if you can highlight it more, perhaps by having it as the first sentence in a separate paragraph.-->
 
@@
 Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines.
-Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Outcomes in severe trauma are dependent on the quality of care received [@ustqi]<!-- It's uncommon to introduce a citation in the middle of a sentence.
+See if you can rephrase the sentence to allow it to be placed last, or divide the sentence somehow, perhaps with a ",", and place the citation just before that.-->  and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
 Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care.
@@
 
-As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions<!-- It's okay to be more direct: "The purpose of M&M conferences is to review selected patient cases, establish if there were opportunities for improvement, and implement corrective actions"-->"" .
 The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care.
@@
 The WHO guidelines for trauma QI programs give specific recommendations as to the structure and format of M&M conferences [@whoguidelines].
-Despite theserecommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
+Despite these recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo].
 The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].
@@
 Karolinska University Hospital also has agreements with several other regions to provide trauma care for patients with severe trauma and the trauma centre has thereby a catchment population of almost 3 million residents [@nksstats].
+
 The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
@@
 ```
+<!-- In the figure it looks like all cases are reviewed by a second nurse, which is not true.
+Most cases are not reviewed by a second nurse, because either none of the audit filters are positive or the first nurse decides that there were no issues.
+So I would add an arrow to the right of the box "Audit filters and primary review..." stating that some cases are excluded at this stage.
+Then I would add a second arrow to the right of "Secondary review..." stating that additional cases are excluded at this step, if no issues are found.
+So it's only a small subset of all cases that are reviewed at a M&M conference-->
 
@@
 The algorithm for PS includes the patient's age, gender, GCS and ISS.
+<!-- This is the older PS, right?
+for the paper I would mention that you used that version because the data necessary for the new version is not included in the registry -->
 
@@
 
-The code for data analysis was written using simulated data to reduce the risk of research bias.
+The code for data analysis was written using simulated data to reduce the risk of researcher bias.
 Real-world data was only used once the data analysis model was completed and found to work correctly on simulated data.
@@
 
-R, a programming language for statistical computing, was used for data collection and analysis.
+R, a programming language for statistical computing, was used for data collection and analysis<!-- Add citation-->.
 Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR.
@@
 Accuracy was plotted and is reported as the highest value for accuracy in the plot together with the associated cutoff.
-Results werepresented with a confidence interval of 95%.
+Results are presented with a confidence interval of 95%.
 A p-value of <0.05 was considered significant.
@@
 A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period.
-Of these, `r exclusion[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion.
+Of these, `r exclusion[2,2]` were reviewed at an M&M conference<!-- Not really.
+Only a small subset of cases were reviewed at a conference, but all were part of the review process--> and therefore became eligible for study inclusion.
 `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`).
@@
 ```
+
+<!-- See my comment in your table.R, but the above could be: -->
+```{r incomplete-table, echo=FALSE}
+knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data")
+```
+
+<!-- And you can then reference this table from anywhere in the text with \@ref(tab:incomplete-table), which would take care of the numbering for you. -->
 
@@
 <!--Figure X??? -->
+<!-- You can also cross reference figures like this: \@ref(fig:fitted-glm) -->
 The fitted curves for the logistic regression models of each score are shown in figure x.
@@
 
-```{r}
+```{r fitted-glm, fig.cap = "Fitted curves for the logistic regression models of each score"}
 knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
+## Make sure you label the axes
 ```
@@
 knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
+## Write Sensitivity and Specificity, with capital S
 ```
@@
 The differences between the scores for each result are shown in table x.
+<!-- Elaborate on what differences are shown in the table.
+You are actually showing the differences in performance measures, right?
+So here you can also say something about what models performed best in terms of what measure-->
 
@@
 Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful.
-Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed.
+Audit filters that don't<!-- Avoid contractions--> improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed.
 Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity.
@@
 ### Health equity
+<!-- Is this section requested by KI?-->
 The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide.
@@
 
-This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, opportunities for improvement, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences.
+This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, opportunities for improvement, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences<!-- No need to state your aim again here in the conclusion, unless it's specifically asked for by KI.-->.
 We showed that, while low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI.

### 2022-12-20 pub.Rmd
saving pub method; still working on other stuff

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

<!--
- TRAUMA BAD 
- QI IMPORTANT
- CASE SELECTION
- AUTOMATE
- PPL LOOK AT PREV DEATH BUT NOT ACROSS THE BOARD 
- 
-->

Methods 
=======

<!-- please note that when i write done here, it doesn't mean i think it's perfect, it's just for my own tracking purposes -->

Study design
------------
<!-- Present key elements of study design -->
This is a single-centre retrospective cohort study that uses data from two registries at Karolinska University Hospital <!-- I would really like to abbreviate this? Ghorbani & Strömmer abbreviated KUH! --> in Stockholm, Sweden: the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models was evaluated using measures of discrimination and calibration.

<!-- "done" -->

Setting
-------
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
The trauma centre at Karolinska University Hospital is equivalent to a level 1 trauma centre according to American College of Surgeons standards and has a catchment population of almost 3 million residents [@nksstats] <!-- not really sure if this needs to be referenced? -->.

The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and uses the Utstein template. All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre. Data is collected from the patient's electronic medical records.

M&M conferences are held at Karolinska University Hospital ten times per year and are attended by all involved specialities and professions. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review. At Karolinska University Hospital, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are screened by research nursing staff to identify cases that potentially received sub-optimal care. These nurses also use audit filters that automatically highlight cases with abnormal parameters. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review. Figure 1 gives an overview of the workflow that leads to determination of OFI.

```{r, echo=FALSE, out.width = "400px"}
knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
## i'm going to fix this based on martin's feedback, ignore it for now 
## or maybe it isn't even useful in the pub version?
```

<!-- "done" -->

Participants
------------
 <!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

We retrospectively analysed patient registries that cover a seven year period, from 2014 to 2021. Participants were patients >15 years of age who were alive on arrival at hospital that were both registered in the trauma registry and also had a recorded outcome of an M&M review. We excluded patients where registry data was lacking that was necessary to calculate trauma scores.

Included in the trauma registry are all patients for whom the trauma team was activated after a potentially traumatic injury, regardless of trauma score, and all patients with a NISS >15 regardless of whether the trauma team was activated. Patients where the only traumatic injury is a chronic subdural haematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded.

<!-- "done" -->

Variables
---------

### Outcome
The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by M&M reviewe at Karolinska University Hospital. OFI is registered in the trauma care quality database and is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".

### Exposures
The exposures of interest were the trauma severity models TRISS, NORMIT, and PS. Variables taken from the registry were age, gender, mechanism of injury, Injury Severity Score (ISS), New Injury Severity Score (NISS), Glasgow Coma Scale (GCS), respiratory rate and systolic blood pressure. Conversion and handling of these variables was carried out according to the SweTrau manual.

The Revised Trauma Score (RTS) was calculated according to the original published algorithm [@rtsalgorithm]. We chose to use the PS12 algorithm instead of the newer PS19 model due to our registry lacking data needed to calculate PS19 [@psalgorithm]. NORMIT was calculated using the revised 2018 coefficients [@skagaValidatingPerformanceTRISS2018]. TRISS was calculated using the revised 2009 coefficients [@triss2010]. In order to avoid excluding patients who were intubated pre-hospitally, the pre-hospital GCS value was used when calculating TRISS for patients who were intubated before arrival at hospital.

<!-- "done" -->

Bias 
----

The code for data analysis was written by a student researcher using simulated data to reduce the risk of researcher bias. Real-world data was only used once the data analysis model was completed and found to work correctly on simulated data.

<!-- "done" -->

Statistical methods
-------------------

<!-- 
(a) Describe all statistical methods, including those used to control
for confounding
(b) Describe any methods used to examine subgroups and interactions
(c) Explain how missing data were addressed 
(d) 
Cohort study: If applicable, explain how loss to follow-up was addressed 
(e) Describe any sensitivity analyses
-->

The statistical programming language R was used for data collection and analysis [@r]. Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the derived OR. The CI for the OR was computed using the Wald estimator. Receiver operator characteristic (ROC) curves were used to assess diagnostic performance. The area under the curve (AUC) was calculated to compare the models with one another. The Integrated Calibration Index (ICI) was used to measure calibration. Accuracy was plotted and is reported as accuracy at a >=0.5 cutoff <!-- fix the >= later -->. Confidence intervals for the AUC, ICI and accuracy were computed with 1000 stratified bootstrap replicates. Results are presented with a confidence interval of 95%. A p-value of <0.05 was considered significant.

<!-- "done" -->

Results
=======

<!-- 
Participants
------------
(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed
(b) Give reasons for non-participation at each stage
(c) Consider use of a flow diagram
Descriptive data
----------------
(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders
(b) Indicate number of participants with missing data for each
variable of interest
(c) Cohort study — Summarise follow-up time (eg, average and total
amount)
Outcome data
------------
Cohort study — Report numbers of outcome events or summary measures
over time
Case-control study — Report numbers in each exposure category, or
summary measures of exposure
Cross-sectional study — Report numbers of outcome events or summary
measures
Main results
------------
(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included
(b) Report category boundaries when continuous variables were
categorized
(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period
Other analyses 
--------------
Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
## source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

Discussion
==========

<!--
Key results
-----------
Summarise key results with reference to study objectives
Limitations
-----------
Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias
Interpretation
--------------
Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence
Generalisability
----------------
Discuss the generalisability (external validity) of the study results
-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2022-12-20 pub.Rmd
very very rough draft of intro

@@
 
-<!--
-- TRAUMA BAD 
-- QI IMPORTANT
-- CASE SELECTION
-- AUTOMATE
-- PPL LOOK AT PREV DEATH BUT NOT ACROSS THE BOARD 
-- 
--->
+<!-- Check the sources in the heim article bc they seem to have some good stuff i missed -->
+<!-- also add a lot more cites -->
+Management of severe trauma is highly time-sensitive intervention that is dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines.
+Outcomes in severe trauma are dependent on the quality of care received [@ustqi], and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care.
+This can only be achieved through peer review and consequently, morbidity and mortality (M&M) review is the cornerstone of all modern trauma quality improvement (QI) programs.
+The purpose of M&M review is to evaluate selected patient cases, establish whether there was opportunity for improvement (OFI), and implement corrective actions.
+
+The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines] at M&M conferences.
+One of the drawbacks of M&M review is the process of identifying which cases should be reviewed.
+Urban trauma centres experience large caseloads and the time resource burden of reviewing even just fatalities can be considerable.
+The process of evaluating cases to determine which cases should be reviewed is also a time-consuming process.
+Due to this, many trauma centres have use audit filters that automatically flag cases for review.
+The potential benefits of this are twofold: it reduces the resource burden of case selection, and effective audit filters may help to identify the cases most likely to have been mismanaged.
+
+One proposed audit filter, that is already used in many hospitals, is the use of trauma severity models that estimate the probability of survival.
+The focus There are many trauma severity models and the choice of model may be dependent on the region, as these algorithms tend to perform best when used in populations similar to the one the model was developed from.
+TRISS, an American model, is the most commonly used model worldwide [@trissbetter].
+The British model PS is commonly used in the UK and Europe [@tarn07].
+NORMIT, a Norwegian model, was validated and found to have good performance in a Swedish trauma centre but was less-suited when applied to a national population [@validationnormit].
+
+One use case for severity models as audit filters is choosing to only review fatalities where the estimated probability of survival was high.
+The assumption in this case is that unexpected deaths are more likely to have OFI, and a fatality where OFI is found is a preventable death.
+Two studies have examined this assumption by investigating whether trauma severity models can predict preventable deaths.
+A 2016 study found that neither TRISS nor PS reliably predicted preventable deaths in a major trauma centre in London [@brohi16].
+Another study in 2019 found that NORMIT and TRISS performed poorly at predicting preventable and non-preventable deaths at a Scandinavian level-I centre [@prevdeathsghorbani].
+
+These studies indicate that trauma severity models are of limited use in improving case selection with regards to fatalities.
+However, to date there have been no studies that examined whether trauma severity models can predict OFI across the board and not just in fatalities.
+The aim of this study is to evaluate how the trauma severity models TRISS, NORMIT and PS are associated with, and to what extent they can predict OFI.
 
@@
 
-<!-- please note that when i write done here, it doesn't mean i think it's perfect, it's just for my own tracking purposes -->
+<!-- please note that when i write done here, it's just for my own tracking purposes to see what parts ive rewritten -->
 

### 2022-12-20 functions/ofi_diagram.R
fixed ofi diagram

@@ -1,42 +1,54 @@
 ## Makes a flowchart on how patients are selected for M&M
 
 ofi_diagram <- function() {
-  diagram <- DiagrammeR::grViz("digraph flowchart {
-  
-    label='Fig. 1. OFI decision-making workflow';
-    labelloc=bottom;
-    labeljust=left;
-    labelfontsize=18;
 
-    # node definitions with substituted label text
-    node [shape = rectangle, width = 8, fontsize = 16] 
-    tab1 [label = '@@1']
-    tab2 [label = '@@2']
-    tab3 [label = '@@3']
-    tab4 [label = '@@4']
-    tab5 [label = '@@5']
-    tab6 [label = '@@6']
-    tab7 [label = '@@7'] 
+  out <- grViz("digraph flowchart {
+  
+  
+    ## the real nodes
+    node [shape = rectangle, width = 6, fontsize = 18] 
+      1 [label = '@@1']
+      2 [label = '@@2']
+      3 [label = '@@3']
+      4 [label = '@@4']
+      5 [label = '@@5']
+      m1 [label = '@@6']
+      m2 [label = '@@7']
+      s1 [label = '@@8']
+      s2 [label = '@@9']
+
     
-    # edge definitions with the node IDs
-    tab1 -> tab2 -> tab3 -> tab4 -> tab5;
-    tab5 -> tab6;
-    tab5 -> tab7;
+    ## the fake nodes to create arrows halfway
+    node [shape=none, width=0, height=0, label='']
+    p1 -> 4; p2 -> 5;
+      {rank=same; 3 -> s1}
+      {rank=same; 4 -> s2}
+
+    edge [dir=none]
+    1 -> 2; 2 -> 3; 3 -> p1; 4 -> p2; 5 -> m1; 5 -> m2;
   }
+
+  [1]: 'Trauma team activation'
+  [2]: 'Inclusion in trauma registry according to SweTrau criteria'
+  [3]: 'Audit filters and primary review by experienced nurse'
+  [4]: 'Secondary review by another experienced nurse'
+  [5]: 'Selected for multidisciplinary M&M conference'
+  [6]: 'Opportunity for improvement found'
+  [7]: 'No opportunity for improvement found'
+  [8]: 'Case not selected for M&M review'
+  [9]: 'Case not selected for M&M review'
   
-    [1]: 'Trauma team activation'
-    [2]: 'Inclusion in registry according to SweTrau criteria'
[…]

### 2022-12-20 manuscript.Rmd
revised paper according to feedbck, wrote abstract

--- 2022-12-15 manuscript.Rmd
@@
 output:
-  pdf_document: default
-  html_document:
-    df_print: paged
-    number_sections: true
-  word_document: default
+  bookdown::pdf_document2: default
+  bookdown::html_document2: default
+  bookdown::word_document2: default
 bibliography: bibliography.bib
@@
 ```
+<!-- rmarkdown::render("manuscript.Rmd", html_document2(global_numbering = TRUE)) -->
 
@@
 what was done and what was found. Not more than 300 words. -->
+*Background:* Morbidity and mortality review, which aims to discover opportunities for improvement (OFI) is a cornerstone of trauma quality improvement worldwide.
+The process of case selection and review is time-consuming, leading many hospitals to use audit filters to automate case selection.
+Trauma severity models, which estimate probability of survival, have been proposed as one such audit filter.
+The ability of these models to predict opportunities for improvement (OFI) is not well understood.
+Two studies have found that they perform poorly at predicting preventable deaths, but as of yet there is no published literature that analyses the performance of these models across the board and not only in deaths.
+*Aim:* To evaluate how the commonly used models TRISS, NORMIT and PS are associated with, and to what extent they can predict, OFI.
+*Methods:* This is a retrospective registry-based cohort study at Karolinska University Hospital covering the period 2014-2021.
+Logistic regression was used to estimate the association between the models and OFI.
+Receiver operating characteristic curves with the corresponding area under the curve (AUC) were used to assess discrimination and the Integrated Calibration Index was used to assess calibration.
+*Results:* 5320 participants were included from a seven-year period.
+Results were similar for all models.
+Lower probability of survival was associated with higher probability of OFI, with TRISS having the strongest association (derived OR: `r t['or']`).
+Predictive performance and calibration was poor across the board.
+TRISS had the best AUC of `r t['auc']`.
+NORMIT had the best calibration with an ICI of `r n['ici']`.
+*Conclusions:* Although low probability of survival is associated with increased odds of OFI, none of the models were able to reliably predict OFI.
+The use of these models as audit filters in case selection for M&M conferences is not supported by our results and may lead to over-selection.
 
@@
 Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].
+
 The quality of trauma care is therefore a globally relevant issue with significance for a great number of people.
@@
 These conferences review selected patient cases and establish whether there was opportunity for improvement (OFI) in each case.
-One method of improving the quality of M&M conferences is by improving case selection.
-This is usually achieved by using audit filters that automatically flag cases for review.
-One approach is the use of trauma severity models as audit filters.
-Trauma severity models are algorithms that evaluate trauma severity in quantitative terms.
-The effectiveness of these models in case selection is understudied and their potential as audit filters is the focus of this paper.
+
+One area that is understudied is the use of trauma severity models as audit filters in case selection for M&M conferences.
+These models are algorithms that estimate the probability of survival.
+It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani ].
+Many hospitals have already implemented this approach [@brohi16]; however, the ability of trauma severity models to predict OFI is not well understood and will be investigated in this paper.
 
@@
 Management of severe trauma is highly time-sensitive and is also dependent on the actions of a chain of healthcare providers spanning multiple professions and disciplines.
-Outcomes in severe trauma are dependent on the quality of care received [@ustqi] and errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
+Outcomes in severe trauma are dependent on the quality of care received and errors in management are a common cause of preventable deaths [@ustqi; @prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira].
 Since mistakes by any one provider can lead to severe outcomes, attempts to improve trauma outcomes must seek to improve all aspects of the entire chain of care.
@@
 
-As mentioned, M&M conferences review a selection of patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions.
+The purpose of M&M conferences is to review selected patient cases, establish whether there were opportunities for improvement, and implement corrective action.
 The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care.
@@
 
-The use of trauma severity models as audit filters is not limited to solely mortalities; some hospitals also use them as audit filters in survivors.
-However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood.
-The two studies mentioned investigated only preventable deaths and to date there is no published research that investigates whether trauma severity models can predict OFI across the board.
+The studies mentioned signify that trauma severity models are not well suited to predicting OFI in fatalities and indicate that they should not be used for this purpose.
+However, the extent to which trauma severity models are associated with or can predict OFI is not well understood, as research thus far has only analysed their role in predicting preventable deaths.
+To date there is no published research that investigates whether trauma severity models can predict OFI across the board.
 
@@
 The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews.
-These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT and PS and OFI using logistic regression.
+These two databases were linked and then analysed to estimate the association between the common trauma severity models TRISS, NORMIT, and PS and OFI using logistic regression.
 The predictive performance of these models was evaluated using measures of discrimination and calibration.
@@
 
-The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and follows SweTrau guidelines.
+The trauma registry at Karolinska University Hospital is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines.
 All trauma patients meeting the SweTrau inclusion criteria are registered in the trauma registry by research nurses at the trauma centre.
 Data is collected from the patient's electronic medical records.
-The trauma registry contains data on patients from 2013 to 2021.
-<!-- Based on the dates I see in the Swetrau dataset -->
 
@@
 The database contains data on patients from 2014 to 2021.
-<!-- Double check these dates -->
 
@@
 Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
-Figure 1 gives an overview of the workflow that leads to determination of OFI.
+Figure \@ref(fig:ofi) gives an overview of the workflow that leads to determination of OFI.
 
-```{r, echo=FALSE, out.width = "400px"}
-knitr::include_graphics("images/ofi.png") ## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
-## titles for plots will be added in word
+```{r ofi, fig.cap = "Workflow for case selection and determining OFI", fig.align = "center", echo=FALSE, out.width = "600px"}
+knitr::include_graphics("images/ofi.png") 
+## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
 ```
@@
 ### Outcome
-The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital.
+The studied outcome was the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at Karolinska University Hospital.
 OFI is registered in the trauma care quality database and is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".
@@
 ### Exposures
-The exposures of interest are the trauma severity models TRISS, NORMIT, and PS.
+The exposures of interest were the trauma severity models TRISS, NORMIT, and PS.
 All three of these models estimate the probability of death, expressed as a probability.
@@
 
-The algorithm for PS includes the patient's age, gender, GCS and ISS.
+The algorithm for PS includes the patient's age, gender, GCS and ISS. We chose to use the PS12 algorithm instead of the newer PS19 model due to our registry lacking data needed to calculate PS19.
 
@@
 
-R, a programming language for statistical computing, was used for data collection and analysis.
+R, a programming language for statistical computing, was used for data collection and analysis [@r].
 Logistic regression was used to estimate association between the estimates of the trauma severity models and OFI and the effect size was given as the OR.
@@
 The Integrated Calibration Index (ICI) was used to measure calibration.
-Accuracy was plotted and is reported as the accuracy at a 0.5 cutoff.
-Confidence intervals were calculated with bootstrapping.
-Results were presented with a confidence interval of 95%.
+Accuracy was plotted and is reported as accuracy at a >=0.5 cutoff.
+Confidence intervals for the AUC, ICI and accuracy were computed with 1000 stratified bootstrap replicates.
+Results are presented with a confidence interval of 95%.
 A p-value of <0.05 was considered significant.
@@
 They also have the right to a copy of data that is stored about them on request.
-<!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->
 
@@
 This study has the potential to improve patient care by analysing a potential factor that could predict suboptimal trauma care.
-It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified by the fact that this population stands to gain the most from improved trauma care.
+It is true that the patient population for this study includes vulnerable populations, such as the deceased and the severely injured, but this is justified in the opinion of this author by the fact that this population stands to gain the most from improved trauma care.
 
@@
 A total of `r exclusion[1,2]` patients were recorded in the trauma registry during the study period.
-Of these, `r exclusion[2,2]` were reviewed at an M&M conference and therefore became eligible for study inclusion.
+Of these, `r exclusion[2,2]` were reviewed and therefore became eligible for study inclusion.
 `r exclusion[7,3]` patients were excluded, due to either being of paediatric age (`r exclusion[3,3]`), being dead on arrival at hospital (`r exclusion[4,3]`), or due to incomplete data (`r exclusion[5,3]`).
-Figure 2 shows the inclusion process for eligible participants.
+Figure \@ref(fig:exclusion) shows the inclusion process for eligible participants.
 
-```{r, echo=FALSE, out.width = "400px"}
-knitr::include_graphics("images/exclusion.png") ## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
+```{r exclusion, fig.cap = "Flow diagram for application of exclusion criteria", fig.align = "center", echo=FALSE, out.width = "600px"}
+knitr::include_graphics("images/exclusion.png") 
+## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
 ```
 
-Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table 1.
+Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table \@ref(tab:incomplete-table).
 Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.
 
-```{r}
- as_latex(table(na.data, type = "incomplete"))
+```{r incomplete-table, echo=FALSE, fig.align = "center"}
+knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data") %>% kable_styling(position = "center", full_width = FALSE)
 ```
@@
 ----------------
-A total of `r exclusion[6,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in table 2.
+A total of `r exclusion[6,2]` participants were identified as eligible for inclusion. Their demographical and clinical characteristics are described in table \@ref(tab:table-one).
 
@@
 
-```{r} 
+```{r table-one} 
   table.one
+## this thing breaks the table counting but ill just fix it manually in word
 ```
@@
 ------------
-
-The fitted curves for the logistic regression models of each score are shown in figure 3.
+The fitted curves for the logistic regression models of each score are shown in figure \@ref(fig:fitted-glm).
 The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.
 
-```{r}
+```{r fitted-glm, fig.cap = "Fitted curves for the logistic regression models of each score"}
 knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
@@
 
-The receiver operating curves for all scores are plotted in 4.
+The receiver operating curves for all scores are plotted in figure \@ref(fig:roc).
 The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.
 
-```{r}
+```{r roc, fig.cap = "Receiver operating curves for each score"}
 knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
@@
 
-The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively.
-The accuracy at all cutoffs is plotted for all models in  5.
+The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`), `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively.
+The accuracy at all cutoffs is plotted for all scores in figure \@ref(fig:acc).
 The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.
 
-```{r}
+```{r acc, fig.cap = "Accuracy at all cutoffs for each score"}
 knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
@@
 
-The differences between the scores for each result are shown in table x.
+The differences between the scores for each result are shown in table \@ref(tab:delta).
+TRISS was superior to the other models with regards to the AUC.
+NORMIT was superior to the other models with regards to accuracy and to the ICI.
 
-```{r}
- as_latex(table(d, type = "deltas"))
+```{r delta, centering = TRUE}
+knitr::kable(table(d, type = "deltas"), caption = "Comparison of model results") %>% kable_styling(position = "center", full_width = FALSE)
 ```
@@
 Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16].
-Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI across the board and not only in mortalities.
+Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI.
 The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.
 
-Even though our results were expected, it was still important to investigate this topic scientifically because trauma severity models are in use today as audit filters.
+Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions.
 Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful.
-Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed.
+Audit filters that do not improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed.
 Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity.

### 2022-12-20 pub.Rmd
revised paper according to feedbck, wrote abstract

@@
 ---
-title: "Manuscript title here"
+title: "Association between trauma severity models and opportunities for improvement:
+  A retrospective cohort study"
 output: html_document
-author: "Your name here"
+author: "Victoria Witthoff"
 bibliography: bibliography.bib
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 ```
 
-<!-- This is an R Markdown document. Markdown is a simple formatting
-syntax for authoring HTML, PDF, and MS Word documents. For more
-details on using R Markdown see http://rmarkdown.rstudio.com -->
-
-<!-- This is a comment and will not be present in the compiled
-document. Try "knitting" this document by clicking the knit button up
-to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
-the console and you should only see headings -->
-
-<!-- Structure and most comments are from the Strengthening the
-reporting of observational studies in epidemiology (STROBE) statement,
-see https://www.strobe-statement.org/. For more explanations see the
-paper Strengthening the Reporting of Observational Studies in
-Epidemiology (STROBE):explanation and elaboration.  -->
-
-<!-- Indicate the study’s design with a commonly used term in the
-title, for example: "Associations between resuscitation interventions
-and opportunities for improvement in adult trauma patients: A cohort
-study" -->
+```{r main}
+  ## source("main.R") 
+```
 

### 2022-12-20 functions/exclusion_flowchart.R

@@ -4,7 +4,6 @@ exclusion_flowchart <- function(exclusion) {
   
   out <- grViz("digraph flowchart {
   
-    label='Fig. 2. Flow diagram for application of exclusion criteria';
     labelloc=bottom;
     labeljust=left;
     labelfontsize=16;

### 2022-12-20 functions/plots.R

@@ -13,7 +13,7 @@ plot_glm <- function(df, type) {
     geom_hline(data = data.frame(c(0.25, 0.50, 0.75)),
                aes(yintercept = c(0.25, 0.50, 0.75)),
                colour = "darkgrey", linetype = "dashed") +
-    labs(x = paste0(type), y = "OFI")
+    labs(x = paste0(type, " probability of survival"), y = "Probability of OFI")
   return(plot)
 }
 
@@ -26,12 +26,11 @@ glm_x3 <- function(df) {
   n <- plot_glm(df.n, "NORMIT")
   p <- plot_glm(df.p, "PS")
   x3 <- grid.arrange(t, n, p, ncol = 3)
-  x3 <- annotate_figure(x3, top = text_grob("Figure x: Probability of OFI vs probability of survival", size = 12))
   ggsave(x3, filename = "images/glm.png",height = 667, width = 2001, units = "px")
 }
 
 plot_roc <- function(x, type) {
-  plot <- ggroc(x[['roc']], colour = "#5FD4AB") + ggtitle(paste0(type, " (AUC = ", x[['auc']], ")")) + theme(plot.title = element_text(size=12))
+  plot <- ggroc(x[['roc']], colour = "#5FD4AB") + ggtitle(paste0(type, " (AUC = ", x[['auc']], ")")) + theme(plot.title = element_text(size=12)) + labs(x = "Specificity", y = "Sensitivity")
   return(plot)
 }
 
@@ -40,7 +39,6 @@ roc_x3 <- function(t, n, p) {
   n.plot <- plot_roc(n, type = "NORMIT")
   p.plot <- plot_roc(p, type = "PS")
   x3 <- grid.arrange(t.plot, n.plot, p.plot, ncol = 3)
-  x3 <- annotate_figure(x3, top = text_grob("Figure x: Receiver Operating Curves", size = 12))
   ggsave(x3, filename = "images/roc.png",height = 667, width = 2001, units = "px")
 }
 
@@ -64,7 +62,6 @@ acc_x3 <- function(t, n, p){
   n.plot <- plot_acc(n, "NORMIT")
   p.plot <- plot_acc(p, "PS")
   x3 <- grid.arrange(t.plot, n.plot, p.plot, ncol = 3)
-  x3 <- annotate_figure(x3, top = text_grob("Figure x: Accuracy at all cutoffs", size = 12))
   ggsave(x3, filename = "images/acc.png",height = 667, width = 2001, units = "px")
   
 }
[…]

### 2022-12-20 functions/table.R

@@ -1,20 +1,10 @@
 table <- function(data, type = "", t = t, n = n, p = p) {
-## This is nice. I would suggest removing gt and just return the
-## tibble. Then call knitr::kable in your manuscript.Rmd and provide
-## a caption argument:
-## ```{r incomplete-table, echo=FALSE}
-## knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data")
-## ```
-## You can then cross-reference your tables in the text and do not
-## have to give them specific numbers here using for example
-## "See Table \@ref(tab:incomplete-table)."
+
   if(type == "incomplete") {
     
     names <- c("GCS", "ASA", "RR", "Systolic BP", "Dominant Injury Type", "Age", "ISS", "NISS", "Gender")
-    tib <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
-    out <- tib %>% gt() %>% tab_header(
-      title = "Table 1: Parameters that were missing in cases excluded for incomplete data"
-    )
+    out <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
+    
     
     return(out)
   }
@@ -37,10 +27,8 @@ table <- function(data, type = "", t = t, n = n, p = p) {
       paste0(data[['acc']][['np']]," (", data[['acc']][['np']][[1]], "-", data[['acc']][['np']][[1]], ")")
     )
     
-    tib <- tibble(" " = comparisons, "AUC difference (95% CI)" = auc, "ICI difference (95% CI)" = ici, "Accuracy difference (95% CI)" = acc)
-    out <- tib %>% gt() %>% tab_header(
-      title = "Table x. Comparison of model results"
-    )
+    out <- tibble(" " = comparisons, "AUC difference (95% CI)" = auc, "ICI difference (95% CI)" = ici, "Accuracy difference (95% CI)" = acc)
+    
     return(out)  
   }
   

### 2022-12-20 main.R

@@ -25,6 +25,7 @@ library(rsvg)
 library(mvbutils)
 library(gridExtra)
 library(ggpubr)
+library(bookdown)
 
 noacsr::source_all_functions()
 

### 2022-12-20 manuscript.Rmd
missing data

@@
 output:
+  bookdown::word_document2: default
+  bookdown::html_document2: default
   bookdown::pdf_document2: default
-  bookdown::html_document2: default
-  bookdown::word_document2: default
 bibliography: bibliography.bib
@@
 Figure \@ref(fig:exclusion) shows the inclusion process for eligible participants.
+Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.
 
@@
 knitr::include_graphics("images/exclusion.png") 
-## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion), "exclusion") to update
-```
-
-Detailed totals of missing data for each required parameter in cases that were excluded for incomplete data are given in table \@ref(tab:incomplete-table).
-Some patients had missing data in several parameters and therefore the total number of missing parameters is greater than the total number of patients excluded due to incomplete data.
-
-```{r incomplete-table, echo=FALSE, fig.align = "center"}
-knitr::kable(table(na.data, type = "incomplete"), caption = "Parameters that were missing in cases excluded for incomplete data") %>% kable_styling(position = "center", full_width = FALSE)
+## exclusion_flowchart.R; save_png(exclusion_flowchart(exclusion, na.data), "exclusion") to update
 ```
@@
 
-```{r table-one} 
-  table.one
-## this thing breaks the table counting but ill just fix it manually in word
+```{r table-one, centering = TRUE}
+t1kable(table.one)
+##knitr::kable(t1kable(table.one), caption = "Comparison of model results") %>% ##kable_styling(position = "center", full_width = FALSE)
 ```
@@
 -----------
-<!-- Summarise key results with reference to study objectives -->
-
 This aim of this study was to evaluate whether several commonly used trauma severity models (TRISS, NORMIT, and PS) are associated with, and to what extent they can predict, OFI, with the goal of improving case selection for multidisciplinary morbidity and mortality conferences.
@@
 It is also possible that standards and expectations of care have changed over time as knowledge and technology improves, which would also in turn affect assignment of OFI.
-<!--JA: I takt med att man åtgärdar systematiskta fel så upkommer na -->
 
@@
 ### Health equity
-<!-- Is this section requested by KI?-->
 The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide.

### 2022-12-20 pub.Rmd
missing data

@@
 ---
-title: "Association between trauma severity models and opportunities for improvement:
-  A retrospective cohort study"
-output: html_document
+title: 'Association between trauma severity models and opportunities for improvement:
+  A retrospective cohort study'
 author: "Victoria Witthoff"
+output:
+  bookdown::pdf_document2: default
+  bookdown::html_document2: default
+  bookdown::word_document2: default
 bibliography: bibliography.bib
 csl: vancouver.csl
+fontsize: 11pts
+linestretch: 1
+always_allow_html: yes
+editor_options: 
+   chunk_output_type: inline
 ---
@@
 
-You can include code in this document like this:
-
-```{r main, echo=FALSE}
-## source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
-
-You can also embed plots:
-
-```{r plot, echo=FALSE}
-plot(pressure)
-```
-
 Discussion

### 2022-12-20 functions/exclusion_flowchart.R

@@ -1,6 +1,6 @@
 ## makes a flowchart of the inclusion/exclusion process
 
-exclusion_flowchart <- function(exclusion) {
+exclusion_flowchart <- function(exclusion, na.data) {
   
   out <- grViz("digraph flowchart {
   
@@ -40,8 +40,11 @@ exclusion_flowchart <- function(exclusion) {
   [6]: paste0('Excluded due to no outcome from M&M review \\n(n =  ', exclusion[2,3], ')')
   [7]: paste0('Excluded due to <15 years of age \\n(n = ', exclusion[3,3], ')')
   [8]: paste0('Excluded due to DOA \\n(n = ', exclusion[4,3], ')')
-  [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n(n = ', exclusion[5,3], ')')
+  [9]: paste0('Excluded due to lacking necessary parameters for analysis \\n (n = ', exclusion[5,3], ')', '\\n \\n Breakdown of missing parameters:\\n ', 'GCS (n = ', na.data[1], ') \\n ASA (n = ', na.data[2], ') \\n RR (n = ', na.data[3], ') \\n Systolic BP (n = ', na.data[4], ') \\n ISS (n = ', na.data[7], ') \\n NISS (n = ', na.data[8], ')')
+  
                                ")
   return(out)
 }
+
+
 

### 2022-12-20 functions/table_one.R

@@ -64,7 +64,7 @@ table_one <- function(cd) {
                 data = df[,vars2], 
                 overall = "Total", 
                 render.categorical = "FREQ (PCTnoNA%)", 
-                caption = "Table 2. Characteristics of participants."
+                caption = "Characteristics of participants."
   )
   return(out)
 }

### 2022-12-20 main.R

@@ -26,6 +26,7 @@ library(mvbutils)
 library(gridExtra)
 library(ggpubr)
 library(bookdown)
+library(kableExtra)
 
 noacsr::source_all_functions()
 

### 2022-12-22 manuscript.Rmd
Minor edits

@@
 Trauma severity models, which estimate probability of survival, have been proposed as one such audit filter.
-The ability of these models to predict opportunities for improvement (OFI) is not well understood.
-Two studies have found that they perform poorly at predicting preventable deaths, but as of yet there is no published literature that analyses the performance of these models across the board and not only in deaths.
+Two studies have found that these models perform poorly at predicting preventable deaths, but the ability of these models to predict opportunities for improvement (OFI) is not well understood.
 *Aim:* To evaluate how the commonly used models TRISS, NORMIT and PS are associated with, and to what extent they can predict, OFI.
 *Methods:* This is a retrospective registry-based cohort study at Karolinska University Hospital covering the period 2014-2021.
-Logistic regression was used to estimate the association between the models and OFI.
-Receiver operating characteristic curves with the corresponding area under the curve (AUC) were used to assess discrimination and the Integrated Calibration Index was used to assess calibration.
+Logistic regression was used to estimate the association between the model scores and OFI.
+Predictive performance was measured using receiver operating characteristic curves with the corresponding area under the curve (AUC) to assess discrimination and the Integrated Calibration Index to assess calibration.
 *Results:* 5320 participants were included from a seven-year period.
@@
 Lower probability of survival was associated with higher probability of OFI, with TRISS having the strongest association (derived OR: `r t['or']`).
-Predictive performance and calibration was poor across the board.
+All model scores had poor predictive performance.
 TRISS had the best AUC of `r t['auc']`.
@@
 Efforts to advance the state of trauma care have been ongoing for many years.
-Today many trauma centres have implemented quality improvement (QI) programs, and it has been shown in the US that these programs improve outcomes [@ustqi].
+Today many trauma centres have implemented quality improvement (QI)<!-- Do you really need to abbreviate quality improvement?
+It's quite short as it is and you don't use it too much--> programs, and it has been shown in the US that these programs improve outcomes [@ustqi].
 QI programs use many tools, including statistical analysis and audit filters [@hakkenbrakprevdeath], but the gold standard of trauma QI is the multidisciplinary morbidity and mortality (M&M) conference.
@@
 These models are algorithms that estimate the probability of survival.
-It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani ].
+It has been proposed that using such models as audit filters to automatically flag cases for review could improve the effectiveness of M&M conferences while also reducing their resource burden [@brohi16; @prevdeathsghorbani].
 Many hospitals have already implemented this approach [@brohi16]; however, the ability of trauma severity models to predict OFI is not well understood and will be investigated in this paper.
@@
 The predominance of male patients invites one to question if female patients are less likely to receive quality care.
-In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected.
+In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority in treating males, whilst treatment of female patients will be neglected.
 This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes.

### 2024-06-09 manuscript.Rmd
committing old changes

@@
 output:
+  bookdown::html_document2: default
   bookdown::word_document2: default
-  bookdown::html_document2: default
   bookdown::pdf_document2: default

### 2024-06-09 test.Rmd
committing old changes

---
title: "test"
author: "Victoria Witthoff"
date: '2022-11-28'
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
  bookdown::word_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r ofi-diagram, fig.cap = "", out.width = "400px"}
knitr::include_graphics("images/ofi.png") 
## ofi_diagram.R; save_png(ofi_diagram(), "ofi") to update
```

blabla \@ref(fig:ofi-diagram)

```{r}
knitr::include_graphics("images/acc.png") ## update: acc_x3(t, n, p)
```

```{r}
knitr::include_graphics("images/roc.png") ## update: roc_x3(t, n, p)
```

```{r}
knitr::include_graphics("images/glm.png") ## update: glm_x3(df)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2024-06-09 functions/plots.R

@@ -13,7 +13,8 @@ plot_glm <- function(df, type) {
     geom_hline(data = data.frame(c(0.25, 0.50, 0.75)),
                aes(yintercept = c(0.25, 0.50, 0.75)),
                colour = "darkgrey", linetype = "dashed") +
-    labs(x = paste0(type, " probability of survival"), y = "Probability of OFI")
+    labs(x = paste0(type, " probability of survival"), y = "Probability of OFI") +
+    theme(axis.title.x = element_text(size = 8))  
   return(plot)
 }
 

### 2024-06-09 main.R

@@ -27,6 +27,8 @@ library(gridExtra)
 library(ggpubr)
 library(bookdown)
 library(kableExtra)
+library(pandoc)
+library(flextable)
 
 noacsr::source_all_functions()
 
@@ -38,7 +40,7 @@ extract.named(prep_data(datasets)) ## creates the prepared dataset, inclusion co
 
 table.one <- table_one(df) ## table_one.R
 dd <- descriptive_data(df) ## descriptive_data.R
-extract.named(results(df, bootstrap = TRUE, boot.no = 10)) ## results.R
+extract.named(results(df, bootstrap = TRUE, boot.no = 1000)) ## results.R
 
 ## make figures
 ##roc_x3(t, n, p)

### 2024-06-09 test.R

@@ -1,0 +1,12 @@
+scores <- c("TRISS", "NORMIT", "PS")
+or <- c()
+or.ci <- c()
+or.p <- c()
+
+##names <- c("GCS", "ASA", "RR", "Systolic BP", "Dominant Injury Type", "Age", "ISS", "NISS", "Gender")
+##tib <- tibble("Parameter" = names, "Total no. cases missing parameter" = data)
+##out <- tib %>% gt() %>% tab_header(
+##  title = "Table 1. Parameters that were missing in cases excluded for incomplete data"
+##)
+
+
